---
title: "MODEL - Rolling window comparison (3, 5, 10 games)"
output: html_document
date: "2026-04-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(MASS)
library(xgboost)
library(glmnet)
```

Helper functions
```{r}
log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

clip_prob <- function(p) {
  pmin(pmax(p, 1e-6), 1 - 1e-6)
}

make_fair_probs <- function(home_odds, draw_odds, away_odds) {
  raw_home <- 1 / home_odds
  raw_draw <- 1 / draw_odds
  raw_away <- 1 / away_odds
  overround <- raw_home + raw_draw + raw_away
  tibble(
    fair_home = raw_home / overround,
    fair_draw = raw_draw / overround,
    fair_away = raw_away / overround
  )
}

match_probs <- function(lambda_home, lambda_away, max_goals = 10) {
  g <- 0:max_goals
  score_matrix <- outer(dpois(g, lambda_home), dpois(g, lambda_away))
  list(
    prob_home = sum(score_matrix[lower.tri(score_matrix)]),
    prob_draw = sum(diag(score_matrix)),
    prob_away = sum(score_matrix[upper.tri(score_matrix)])
  )
}

make_calibration_table <- function(actual, pred, bucket_width = 0.1) {
  tibble(actual = actual, pred = pred) %>%
    filter(!is.na(actual), !is.na(pred)) %>%
    mutate(prob_bucket = cut(pred, breaks = seq(0, 1, by = bucket_width), include.lowest = TRUE)) %>%
    group_by(prob_bucket) %>%
    summarise(n = n(), avg_pred = mean(pred), actual_rate = mean(actual), .groups = "drop")
}

add_profit <- function(df) {
  df %>%
    mutate(odds = 1 / fair_home, profit = if_else(home_win.y == 1, odds - 1, -1))
}

# ── Core modelling function ──────────────────────────────────────
# Takes a feature set suffix (e.g. "3", "5", "10") and runs
# NB, LASSO and XGBoost, returning match-level win probabilities
run_models <- function(train_df, test_df, window) {

  feat <- c(
    paste0("roll", window, "_xG_openplay"),
    paste0("roll", window, "_xG_freekick"),
    paste0("opp_roll", window, "_xG_openplay"),
    paste0("opp_roll", window, "_xG_freekick"),
    paste0("sd", window, "_xG"),
    paste0("opp_sd", window, "_xG"),
    "home_flag"
  )

  # ── NB ──────────────────────────────────────────────────────
  nb_formula <- as.formula(paste("goals ~", paste(feat, collapse = " + ")))

  nb_mod <- glm.nb(nb_formula, data = train_df)
  cat("\n=== NB model (window =", window, ") ===\n")
  print(summary(nb_mod))
  cat("Theta:", nb_mod$theta, "\n")

  train_df[[paste0("pred_nb_", window)]] <- predict(nb_mod, newdata = train_df, type = "response")
  test_df[[paste0("pred_nb_", window)]]  <- predict(nb_mod, newdata = test_df,  type = "response")

  # ── LASSO ───────────────────────────────────────────────────
  lasso_form <- as.formula(paste("~", paste(feat, collapse = " + "), "- 1"))

  lasso_train_mat <- model.matrix(lasso_form, data = train_df)
  lasso_test_mat  <- model.matrix(lasso_form, data = test_df)

  lasso_cv <- cv.glmnet(x = lasso_train_mat, y = train_df$goals, family = "poisson", alpha = 1)

  cat("\n=== LASSO coefficients (window =", window, ") ===\n")
  print(coef(lasso_cv, s = "lambda.min"))

  train_df[[paste0("pred_lasso_", window)]] <- as.numeric(
    predict(lasso_cv, newx = lasso_train_mat, s = "lambda.min", type = "response"))
  test_df[[paste0("pred_lasso_", window)]]  <- as.numeric(
    predict(lasso_cv, newx = lasso_test_mat,  s = "lambda.min", type = "response"))

  # ── XGBoost ─────────────────────────────────────────────────
  train_mat <- xgb.DMatrix(data = as.matrix(train_df[, feat]), label = train_df$goals)
  test_mat  <- xgb.DMatrix(data = as.matrix(test_df[, feat]),  label = test_df$goals)

  xgb_mod <- xgb.train(
    params = list(
      objective = "count:poisson", eta = 0.05, max_depth = 4,
      subsample = 0.8, colsample_bytree = 0.8, eval_metric = "poisson-nloglik"
    ),
    data = train_mat, nrounds = 500,
    watchlist = list(train = train_mat, test = test_mat),
    early_stopping_rounds = 30, verbose = 1, print_every_n = 50
  )

  train_df[[paste0("pred_xgb_", window)]] <- predict(xgb_mod, train_mat)
  test_df[[paste0("pred_xgb_", window)]]  <- predict(xgb_mod, test_mat)

  # ── Feature importance ───────────────────────────────────────
  imp <- xgb.importance(feature_names = feat, model = xgb_mod)
  cat("\n=== XGBoost feature importance (window =", window, ") ===\n")
  print(imp)

  list(train_df = train_df, test_df = test_df, nb = nb_mod, lasso = lasso_cv, xgb = xgb_mod)
}

# ── Convert team-level predictions to match-level probabilities ─
to_match_probs <- function(df, window, split) {
  nb_col    <- paste0("pred_nb_",    window)
  lasso_col <- paste0("pred_lasso_", window)
  xgb_col   <- paste0("pred_xgb_",  window)

  home_df <- df %>%
    filter(home_flag == 1) %>%
    select(match_id, date, season,
           home_team = team, away_team = opponent, home_goals = goals,
           lambda_home_nb    = !!nb_col,
           lambda_home_lasso = !!lasso_col,
           lambda_home_xgb   = !!xgb_col)

  away_df <- df %>%
    filter(home_flag == 0) %>%
    select(match_id, away_goals = goals,
           lambda_away_nb    = !!nb_col,
           lambda_away_lasso = !!lasso_col,
           lambda_away_xgb   = !!xgb_col)

  match_df <- home_df %>%
    left_join(away_df, by = "match_id") %>%
    mutate(
      actual_result = case_when(
        home_goals > away_goals  ~ "home",
        home_goals == away_goals ~ "draw",
        TRUE                     ~ "away"
      ),
      home_win = as.integer(actual_result == "home")
    )

  probs_nb <- pmap_dfr(
    list(match_df$lambda_home_nb, match_df$lambda_away_nb),
    function(lh, la) { p <- match_probs(lh, la); tibble(p_home_nb = p$prob_home, p_draw_nb = p$prob_draw, p_away_nb = p$prob_away) }
  )
  probs_lasso <- pmap_dfr(
    list(match_df$lambda_home_lasso, match_df$lambda_away_lasso),
    function(lh, la) { p <- match_probs(lh, la); tibble(p_home_lasso = p$prob_home, p_draw_lasso = p$prob_draw, p_away_lasso = p$prob_away) }
  )
  probs_xgb <- pmap_dfr(
    list(match_df$lambda_home_xgb, match_df$lambda_away_xgb),
    function(lh, la) { p <- match_probs(lh, la); tibble(p_home_xgb = p$prob_home, p_draw_xgb = p$prob_draw, p_away_xgb = p$prob_away) }
  )

  bind_cols(match_df, probs_nb, probs_lasso, probs_xgb)
}
```

Load data
```{r}
pl <- readRDS("pl_clean.rds")
team_match_df <- readRDS("pl_team_match_df.rds")
cat("Columns in loaded RDS:", paste(names(team_match_df), collapse = ", "), "\n")

team_map <- c(
  "Man City" = "Manchester City", "Man United" = "Manchester United",
  "Newcastle" = "Newcastle United", "Nott'm Forest" = "Nottingham Forest",
  "West Brom" = "West Bromwich Albion", "Wolves" = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map),
    Date = as.Date(Date),
    home_win = as.integer(FTR == "H")
  )

fair_probs <- make_fair_probs(pl$B365H, pl$B365D, pl$B365A)
pl <- bind_cols(pl, fair_probs)

team_match_df <- team_match_df %>%
  mutate(date = as.Date(date)) %>%
  drop_na()
```

Train/test split
```{r}
train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows: ", nrow(test_df),  "\n")
```

Run all three window models
```{r}
results_3  <- run_models(train_df, test_df, "3")
results_5  <- run_models(results_3$train_df,  results_3$test_df,  "5")
results_10 <- run_models(results_5$train_df,  results_5$test_df,  "10")

train_final <- results_10$train_df
test_final  <- results_10$test_df
```

Convert to match-level probabilities
```{r}
match_train_3  <- to_match_probs(train_final, "3",  "train")
match_test_3   <- to_match_probs(test_final,  "3",  "test")

match_train_5  <- to_match_probs(train_final, "5",  "train")
match_test_5   <- to_match_probs(test_final,  "5",  "test")

match_train_10 <- to_match_probs(train_final, "10", "train")
match_test_10  <- to_match_probs(test_final,  "10", "test")
```

Merge bookmaker data
```{r}
book_train <- pl %>% filter(Date < as.Date("2023-08-01")) %>%
  select(Date, HomeTeam, AwayTeam, home_win, fair_home, fair_draw, fair_away)

book_test <- pl %>% filter(Date >= as.Date("2023-08-01")) %>%
  select(Date, HomeTeam, AwayTeam, home_win, fair_home, fair_draw, fair_away)

merge_book <- function(match_df, book_df) {
  match_df %>%
    left_join(book_df, by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")) %>%
    filter(!is.na(fair_home))
}

eval_3  <- merge_book(match_test_3,  book_test)
eval_5  <- merge_book(match_test_5,  book_test)
eval_10 <- merge_book(match_test_10, book_test)
```

Main results comparison across all windows
```{r}
make_results_table <- function(eval_df, window_label) {
  tibble(
    window   = window_label,
    model    = c("Bookmaker", "NB", "LASSO", "XGBoost"),
    log_loss = c(
      log_loss(eval_df$home_win.y, eval_df$fair_home),
      log_loss(eval_df$home_win.y, eval_df$p_home_nb),
      log_loss(eval_df$home_win.y, eval_df$p_home_lasso),
      log_loss(eval_df$home_win.y, eval_df$p_home_xgb)
    ),
    brier = c(
      brier_score(eval_df$home_win.y, eval_df$fair_home),
      brier_score(eval_df$home_win.y, eval_df$p_home_nb),
      brier_score(eval_df$home_win.y, eval_df$p_home_lasso),
      brier_score(eval_df$home_win.y, eval_df$p_home_xgb)
    )
  )
}

results_all <- bind_rows(
  make_results_table(eval_3,  "3-game"),
  make_results_table(eval_5,  "5-game"),
  make_results_table(eval_10, "10-game")
)

cat("\n=== Results by rolling window ===\n")
print(results_all)
```

```{r}
# Compare betting returns across all windows and models
betting_comparison <- bind_rows(
  eval_3 %>% mutate(lambda_sum = lambda_home_nb + lambda_away_nb) %>%
    filter(fair_home > 0.4 & fair_home < 0.6,
           lambda_sum > quantile(lambda_sum, 0.6, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(window = "3-game", model = "NB", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE)),

  eval_5 %>% mutate(lambda_sum = lambda_home_nb + lambda_away_nb) %>%
    filter(fair_home > 0.4 & fair_home < 0.6,
           lambda_sum > quantile(lambda_sum, 0.6, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(window = "5-game", model = "NB", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE)),

  eval_10 %>% mutate(lambda_sum = lambda_home_nb + lambda_away_nb) %>%
    filter(fair_home > 0.4 & fair_home < 0.6,
           lambda_sum > quantile(lambda_sum, 0.6, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(window = "10-game", model = "NB", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE)),

  eval_3 %>% mutate(lambda_sum = lambda_home_xgb + lambda_away_xgb) %>%
    filter(fair_home > 0.4 & fair_home < 0.6,
           lambda_sum > quantile(lambda_sum, 0.6, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(window = "3-game", model = "XGBoost", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE)),

  eval_5 %>% mutate(lambda_sum = lambda_home_xgb + lambda_away_xgb) %>%
    filter(fair_home > 0.4 & fair_home < 0.6,
           lambda_sum > quantile(lambda_sum, 0.6, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(window = "5-game", model = "XGBoost", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE)),

  eval_10 %>% mutate(lambda_sum = lambda_home_xgb + lambda_away_xgb) %>%
    filter(fair_home > 0.4 & fair_home < 0.6,
           lambda_sum > quantile(lambda_sum, 0.6, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(window = "10-game", model = "XGBoost", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE))
)

print(betting_comparison)
```


Calibration tables for best window (5-game)
```{r}
cal_book  <- make_calibration_table(eval_5$home_win.y, eval_5$fair_home)
cal_nb    <- make_calibration_table(eval_5$home_win.y, eval_5$p_home_nb)
cal_lasso <- make_calibration_table(eval_5$home_win.y, eval_5$p_home_lasso)
cal_xgb   <- make_calibration_table(eval_5$home_win.y, eval_5$p_home_xgb)

cal_book
cal_nb
cal_lasso
cal_xgb
```

Subgroup analysis by favourites — 5-game window
```{r}
subgroup_results <- eval_5 %>%
  mutate(
    fav_group = case_when(
      fair_home >= 0.60 ~ "Strong home favourite",
      fair_home >= 0.45 ~ "Moderate home favourite",
      fair_home >= 0.30 ~ "Balanced / slight away favourite",
      TRUE              ~ "Strong away favourite"
    )
  ) %>%
  group_by(fav_group) %>%
  summarise(
    n              = n(),
    book_log_loss  = log_loss(home_win.y, fair_home),
    nb_log_loss    = log_loss(home_win.y, p_home_nb),
    lasso_log_loss = log_loss(home_win.y, p_home_lasso),
    xgb_log_loss   = log_loss(home_win.y, p_home_xgb),
    .groups = "drop"
  )

print(subgroup_results)
```

Betting strategy — using 5-game NB lambda sum
```{r}
eval_5 <- eval_5 %>%
  mutate(
    lambda_sum_nb  = lambda_home_nb + lambda_away_nb,
    lambda_diff_nb = lambda_home_nb - lambda_away_nb
  )

bets_main <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
  add_profit()

cat("Main strategy — avg profit:", mean(bets_main$profit, na.rm = TRUE), "\n")
cat("N bets:", nrow(bets_main), "\n")
```

Betting strategy: median split
```{r}
bets_total_goals <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > median(lambda_sum_nb, na.rm = TRUE)) %>%
  add_profit()

mean(bets_total_goals$profit, na.rm = TRUE)
nrow(bets_total_goals)
```

Betting strategy: top 30 percent
```{r}
bets_stricter <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.7, na.rm = TRUE)) %>%
  add_profit()

mean(bets_stricter$profit, na.rm = TRUE)
nrow(bets_stricter)
```

Betting strategy: low expected goals
```{r}
bets_low <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb < quantile(lambda_sum_nb, 0.4, na.rm = TRUE)) %>%
  add_profit()

mean(bets_low$profit, na.rm = TRUE)
nrow(bets_low)
```

Profit by expected goal intensity quintile
```{r}
profit_by_bin <- eval_5 %>%
  mutate(bin = ntile(lambda_sum_nb, 5)) %>%
  add_profit() %>%
  group_by(bin) %>%
  summarise(avg_profit = mean(profit, na.rm = TRUE), n = n(), .groups = "drop")

print(profit_by_bin)
```

Robustness across different expected-goal cutoffs
```{r}
robustness_table <- purrr::map_dfr(c(0.50, 0.60, 0.70, 0.80), function(q) {
  bets_q <- eval_5 %>%
    filter(fair_home > 0.4 & fair_home < 0.6,
           lambda_sum_nb > quantile(lambda_sum_nb, q, na.rm = TRUE)) %>%
    add_profit()
  tibble(quantile_cutoff = q, n_bets = nrow(bets_q), avg_profit = mean(bets_q$profit, na.rm = TRUE))
})

print(robustness_table)
```

Robustness across alternative probability ranges
```{r}
range_results <- bind_rows(
  eval_5 %>% filter(fair_home > 0.35 & fair_home < 0.65,
                    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
    add_profit() %>% summarise(range = "0.35-0.65", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  eval_5 %>% filter(fair_home > 0.40 & fair_home < 0.60,
                    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
    add_profit() %>% summarise(range = "0.40-0.60", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  eval_5 %>% filter(fair_home > 0.45 & fair_home < 0.55,
                    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
    add_profit() %>% summarise(range = "0.45-0.55", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE))
)

print(range_results)
```

Heterogeneity across seasons
```{r}
season_results <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
  add_profit() %>%
  group_by(season) %>%
  summarise(n_bets = n(), avg_profit = mean(profit, na.rm = TRUE),
            actual_home_win = mean(home_win.y), avg_book_prob = mean(fair_home), .groups = "drop")

print(season_results)
```

Heterogeneity within mid-range segment
```{r}
mid_bin_results <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
  add_profit() %>%
  mutate(prob_bin = cut(fair_home, breaks = c(0.4, 0.45, 0.5, 0.55, 0.6), include.lowest = TRUE)) %>%
  group_by(prob_bin) %>%
  summarise(n_bets = n(), avg_profit = mean(profit, na.rm = TRUE),
            actual_home_win = mean(home_win.y), avg_book_prob = mean(fair_home), .groups = "drop")

print(mid_bin_results)
```

Direct comparison: high vs low expected goals
```{r}
high_low_comparison <- bind_rows(
  eval_5 %>% filter(fair_home > 0.4 & fair_home < 0.6,
                    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(group = "High expected goals", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE),
              actual_home_win = mean(home_win.y), avg_book_prob = mean(fair_home)),
  eval_5 %>% filter(fair_home > 0.4 & fair_home < 0.6,
                    lambda_sum_nb < quantile(lambda_sum_nb, 0.4, na.rm = TRUE)) %>%
    add_profit() %>%
    summarise(group = "Low expected goals", n_bets = n(),
              avg_profit = mean(profit, na.rm = TRUE),
              actual_home_win = mean(home_win.y), avg_book_prob = mean(fair_home))
)

print(high_low_comparison)
```

Bootstrap confidence interval
```{r}
set.seed(123)

bets_boot <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
  add_profit()

boot_means <- replicate(2000, {
  mean(sample(bets_boot$profit, size = nrow(bets_boot), replace = TRUE), na.rm = TRUE)
})

quantile(boot_means, probs = c(0.025, 0.5, 0.975))
```

Placebo test
```{r}
set.seed(123)

target_n <- nrow(eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)))

all_mid_bets <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6) %>%
  add_profit()

placebo_means <- replicate(2000, {
  mean(slice_sample(all_mid_bets, n = target_n, replace = FALSE)$profit, na.rm = TRUE)
})

quantile(placebo_means, probs = c(0.025, 0.5, 0.975))
```

Wider strategy check
```{r}
bets_wider <- eval_5 %>%
  filter(fair_home > 0.35 & fair_home < 0.65,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.5, na.rm = TRUE)) %>%
  add_profit()

mean(bets_wider$profit, na.rm = TRUE)
nrow(bets_wider)
t.test(bets_wider$profit)
```
```{r}
# Sample size checks
sample_size_checks <- tibble(
  segment = c(
    "Mid-range odds only",
    "Mid-range + high total goals (median split)",
    "Mid-range + top 40% total goals",
    "Mid-range + top 30% total goals",
    "Mid-range + low total goals"
  ),
  n = c(
    nrow(eval_5 %>% filter(fair_home > 0.4 & fair_home < 0.6)),
    nrow(bets_total_goals),
    nrow(eval_5 %>% filter(fair_home > 0.4 & fair_home < 0.6,
                           lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE))),
    nrow(bets_stricter),
    nrow(bets_low)
  )
)

print(sample_size_checks)
```

```{r}
# Concentration check
bets_concentration <- eval_5 %>%
  filter(fair_home > 0.4 & fair_home < 0.6,
         lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
  add_profit()

summary(bets_concentration$profit)

quantile(
  bets_concentration$profit,
  probs = c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1),
  na.rm = TRUE
)
```

```{r}
# T-test on main strategy
t.test(bets_main$profit)
```

