---
title: "independent model"
output: html_document
date: "2026-04-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(MASS)
library(xgboost)
```
Helper functions
```{r}
log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

clip_prob <- function(p) {
  pmin(pmax(p, 1e-6), 1 - 1e-6)
}

# Remove bookmaker margin from 1X2 odds
make_fair_probs <- function(home_odds, draw_odds, away_odds) {
  raw_home <- 1 / home_odds
  raw_draw <- 1 / draw_odds
  raw_away <- 1 / away_odds
  
  overround <- raw_home + raw_draw + raw_away
  
  tibble(
    fair_home = raw_home / overround,
    fair_draw = raw_draw / overround,
    fair_away = raw_away / overround
  )
}

# Convert predicted home/away goal means into home/draw/away probabilities
match_probs <- function(lambda_home, lambda_away, max_goals = 10) {
  g <- 0:max_goals
  score_matrix <- outer(dpois(g, lambda_home), dpois(g, lambda_away))
  
  list(
    prob_home = sum(score_matrix[lower.tri(score_matrix)]),
    prob_draw = sum(diag(score_matrix)),
    prob_away = sum(score_matrix[upper.tri(score_matrix)])
  )
}
```

Load data
```{r}
pl <- readRDS("pl_clean.rds")
team_match_df <- readRDS("epl_team_match_df.rds")
```

Align team names and clean bookmaker data
```{r}
team_map <- c(
  "Man City" = "Manchester City",
  "Man United" = "Manchester United",
  "Newcastle" = "Newcastle United",
  "Nott'm Forest" = "Nottingham Forest",
  "West Brom" = "West Bromwich Albion",
  "Wolves" = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map),
    Date = as.Date(Date),
    home_win = as.integer(FTR == "H")
  )
```

Create bookmaker fair probabilities
```{r}
fair_probs <- make_fair_probs(pl$B365H, pl$B365D, pl$B365A)

pl <- bind_cols(pl, fair_probs)
```

Check variables in team-level data
```{r}
names(team_match_df)
```
Keep only needed columns
```{r}
feature_names <- c(
  "roll_xG_openplay",
  "roll_xG_freekick",
  "opp_roll_xG_openplay",
  "opp_roll_xG_freekick",
  "sd_xG",
  "opp_sd_xG",
  "home_flag"
)

required_cols <- c(
  "match_id", "season", "date", "team", "opponent", "goals", feature_names
)

team_match_df <- team_match_df %>%
  select(all_of(required_cols)) %>%
  mutate(date = as.Date(date)) %>%
  drop_na()
```

Train/test split
```{r}
train_df <- team_match_df %>%
  filter(season <= 2022)

test_df <- team_match_df %>%
  filter(season >= 2023)

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows: ", nrow(test_df), "\n")
```

Fit Negative Binomial model
```{r}
nb_model <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

summary(nb_model)
cat("Theta:", nb_model$theta, "\n")
```
Predict team-level expected goals with NB
```{r}
train_matrix <- xgb.DMatrix(
  data = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

watchlist <- list(
  train = train_matrix,
  test  = test_matrix
)

xgb_model <- xgb.train(
  params = list(
    objective = "count:poisson",
    eta = 0.05,
    max_depth = 4,
    subsample = 0.8,
    colsample_bytree = 0.8,
    eval_metric = "poisson-nloglik"
  ),
  data = train_matrix,
  nrounds = 500,
  watchlist = watchlist,
  early_stopping_rounds = 30,
  verbose = 1,
  print_every_n = 50
)
```
Predict team-level expected goals with XGBoost
```{r}
train_df <- train_df %>%
  mutate(pred_xgb = predict(xgb_model, train_matrix))

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))
```

Convert team-level predictions to match-level rows
```{r}
home_train <- train_df %>%
  filter(home_flag == 1) %>%
  select(
    match_id, date, season,
    home_team = team,
    away_team = opponent,
    home_goals = goals,
    lambda_home_nb = pred_nb,
    lambda_home_xgb = pred_xgb
  )

away_train <- train_df %>%
  filter(home_flag == 0) %>%
  select(
    match_id,
    away_goals = goals,
    lambda_away_nb = pred_nb,
    lambda_away_xgb = pred_xgb
  )

match_train <- home_train %>%
  left_join(away_train, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals > away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      TRUE ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )
```

Same for test sample
```{r}
home_test <- test_df %>%
  filter(home_flag == 1) %>%
  select(
    match_id, date, season,
    home_team = team,
    away_team = opponent,
    home_goals = goals,
    lambda_home_nb = pred_nb,
    lambda_home_xgb = pred_xgb
  )

away_test <- test_df %>%
  filter(home_flag == 0) %>%
  select(
    match_id,
    away_goals = goals,
    lambda_away_nb = pred_nb,
    lambda_away_xgb = pred_xgb
  )

match_test <- home_test %>%
  left_join(away_test, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals > away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      TRUE ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )
```

Convert expected goals to home/draw/away probabilities
```{r}
probs_train_nb <- pmap_dfr(
  list(match_train$lambda_home_nb, match_train$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_nb = p$prob_home,
      p_draw_nb = p$prob_draw,
      p_away_nb = p$prob_away
    )
  }
)

probs_train_xgb <- pmap_dfr(
  list(match_train$lambda_home_xgb, match_train$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_xgb = p$prob_home,
      p_draw_xgb = p$prob_draw,
      p_away_xgb = p$prob_away
    )
  }
)

match_train <- bind_cols(match_train, probs_train_nb, probs_train_xgb)
```

Same probability conversion for test
```{r}
probs_test_nb <- pmap_dfr(
  list(match_test$lambda_home_nb, match_test$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_nb = p$prob_home,
      p_draw_nb = p$prob_draw,
      p_away_nb = p$prob_away
    )
  }
)

probs_test_xgb <- pmap_dfr(
  list(match_test$lambda_home_xgb, match_test$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_xgb = p$prob_home,
      p_draw_xgb = p$prob_draw,
      p_away_xgb = p$prob_away
    )
  }
)

match_test <- bind_cols(match_test, probs_test_nb, probs_test_xgb)
```

Merge bookmaker benchmarak into train sample
```{r}
book_train <- pl %>%
  filter(Date < as.Date("2023-08-01")) %>%
  select(
    Date, HomeTeam, AwayTeam, home_win,
    fair_home, fair_draw, fair_away
  )

match_train <- match_train %>%
  left_join(
    book_train,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )
```

Merge bookmaker benchmark into test sample
```{r}
book_test <- pl %>%
  filter(Date >= as.Date("2023-08-01")) %>%
  select(
    Date, HomeTeam, AwayTeam, home_win,
    fair_home, fair_draw, fair_away
  )

match_test <- match_test %>%
  left_join(
    book_test,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )
```

Check successful merge
```{r}
cat("Train matches with bookmaker odds:", sum(!is.na(match_train$fair_home)), "out of", nrow(match_train), "\n")
cat("Test matches with bookmaker odds: ", sum(!is.na(match_test$fair_home)), "out of", nrow(match_test), "\n")
```
Keep only matched rows
```{r}
match_train_eval <- match_train %>%
  filter(!is.na(fair_home))

match_test_eval <- match_test %>%
  filter(!is.na(fair_home))
```

Main out-of-sample comparison: bookmaker vs NB vs XGBoost
```{r}
results_main <- tibble(
  model = c("Bookmaker", "NB model", "XGBoost model"),
  log_loss = c(
    log_loss(match_test_eval$home_win.y, match_test_eval$fair_home),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_xgb)
  ),
  brier = c(
    brier_score(match_test_eval$home_win.y, match_test_eval$fair_home),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_xgb)
  )
)

results_main
```

Estimated combination model on training data
```{r}
logit <- function(p) {
  p <- clip_prob(p)
  log(p / (1 - p))
}

match_train_eval <- match_train_eval %>%
  mutate(
    logit_book = logit(fair_home),
    logit_nb = logit(p_home_nb),
    logit_xgb = logit(p_home_xgb)
  )

match_test_eval <- match_test_eval %>%
  mutate(
    logit_book = logit(fair_home),
    logit_nb = logit(p_home_nb),
    logit_xgb = logit(p_home_xgb)
  )
```

Combination: bookmaker + NB
```{r}
combo_nb <- glm(
  home_win.y ~ logit_book + logit_nb,
  data = match_train_eval,
  family = binomial
)

summary(combo_nb)
```
Predict combined NB model on test
```{r}
match_test_eval <- match_test_eval %>%
  mutate(
    p_combo_nb = predict(combo_nb, newdata = ., type = "response")
  )
```

Combination: bookmaker + XGboost
```{r}
combo_xgb <- glm(
  home_win.y ~ logit_book + logit_xgb,
  data = match_train_eval,
  family = binomial
)

summary(combo_xgb)
```
Predict combined XGboost model on test
```{r}
match_test_eval <- match_test_eval %>%
  mutate(
    p_combo_xgb = predict(combo_xgb, newdata = ., type = "response")
  )
```

Extended results table
```{r}
results_extended <- tibble(
  model = c(
    "Bookmaker",
    "NB model",
    "XGBoost model",
    "Bookmaker + NB",
    "Bookmaker + XGBoost"
  ),
  log_loss = c(
    log_loss(match_test_eval$home_win.y, match_test_eval$fair_home),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_xgb),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_combo_nb),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_combo_xgb)
  ),
  brier = c(
    brier_score(match_test_eval$home_win.y, match_test_eval$fair_home),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_xgb),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_combo_nb),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_combo_xgb)
  )
)

results_extended
```
Calibration table function
```{r}
make_calibration_table <- function(actual, pred, bucket_width = 0.1) {
  tibble(actual = actual, pred = pred) %>%
    filter(!is.na(actual), !is.na(pred)) %>%
    mutate(
      prob_bucket = cut(
        pred,
        breaks = seq(0, 1, by = bucket_width),
        include.lowest = TRUE
      )
    ) %>%
    group_by(prob_bucket) %>%
    summarise(
      n = n(),
      avg_pred = mean(pred),
      actual_rate = mean(actual),
      .groups = "drop"
    )
}
```

Calibration tables
```{r}
cal_book <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$fair_home)
cal_nb   <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$p_home_nb)
cal_xgb  <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$p_home_xgb)

cal_book
cal_nb
cal_xgb
```
Subgroup analysis: favourties vs non-favourites
```{r}
subgroup_results <- match_test_eval %>%
  mutate(
    fav_group = case_when(
      fair_home >= 0.60 ~ "Strong home favourite",
      fair_home >= 0.45 ~ "Moderate home favourite",
      fair_home >= 0.30 ~ "Balanced / slight away favourite",
      TRUE ~ "Strong away favourite"
    )
  ) %>%
  group_by(fav_group) %>%
  summarise(
    n = n(),
    book_log_loss = log_loss(home_win.y, fair_home),
    nb_log_loss   = log_loss(home_win.y, p_home_nb),
    xgb_log_loss  = log_loss(home_win.y, p_home_xgb),
    .groups = "drop"
  )

subgroup_results
```
Identify matches where our model disagrees most with bookmaker
```{r}
edge_table <- match_test_eval %>%
  mutate(
    edge_nb = p_home_nb - fair_home,
    edge_xgb = p_home_xgb - fair_home
  ) %>%
  arrange(desc(abs(edge_nb))) %>%
  select(
    date, home_team, away_team,
    home_goals, away_goals,
    fair_home, p_home_nb, p_home_xgb,
    edge_nb, edge_xgb
  )

head(edge_table, 20)
```

EXTENSION: KAN VI MODELLERE BOOKMAKERENS FORECAST-FEJL OG BRUGE DET TIL AT LAVE ET FORBEDRET HOME-WIN FORECAST?

Tjek at nødvendige objekter findes
```{r}
exists("match_train_eval")
exists("match_test_eval")
names(match_train_eval)
names(match_test_eval)
```
Lav bookmakernens forecast error i train og test
```{r}
match_train_eval <- match_train_eval %>%
  mutate(
    error_book = home_win.y - fair_home
  )

match_test_eval <- match_test_eval %>%
  mutate(
    error_book = home_win.y - fair_home
  )
```

Lav centrale difference- og uncertainty-features
```{r}
match_train_eval <- match_train_eval %>%
  mutate(
    pdiff_nb   = p_home_nb - fair_home,
    pdiff_xgb  = p_home_xgb - fair_home,
    lambda_diff_nb  = lambda_home_nb - lambda_away_nb,
    lambda_sum_nb   = lambda_home_nb + lambda_away_nb,
    lambda_diff_xgb = lambda_home_xgb - lambda_away_xgb,
    lambda_sum_xgb  = lambda_home_xgb + lambda_away_xgb
  )

match_test_eval <- match_test_eval %>%
  mutate(
    pdiff_nb   = p_home_nb - fair_home,
    pdiff_xgb  = p_home_xgb - fair_home,
    lambda_diff_nb  = lambda_home_nb - lambda_away_nb,
    lambda_sum_nb   = lambda_home_nb + lambda_away_nb,
    lambda_diff_xgb = lambda_home_xgb - lambda_away_xgb,
    lambda_sum_xgb  = lambda_home_xgb + lambda_away_xgb
  )
```

Kig på simple sammenhænge først
```{r}
summary(lm(error_book ~ pdiff_nb, data = match_train_eval))
summary(lm(error_book ~ pdiff_xgb, data = match_train_eval))
summary(lm(error_book ~ lambda_diff_nb, data = match_train_eval))
summary(lm(error_book ~ lambda_sum_nb, data = match_train_eval))
```
Baseline error model
```{r}
error_model_1 <- lm(
  error_book ~ pdiff_nb + lambda_diff_nb + lambda_sum_nb,
  data = match_train_eval
)

summary(error_model_1)
```
Mere fleksibel error model med nonlinearitet
```{r}
error_model_2 <- lm(
  error_book ~
    pdiff_nb +
    I(pdiff_nb^2) +
    lambda_diff_nb +
    I(lambda_diff_nb^2) +
    lambda_sum_nb,
  data = match_train_eval
)

summary(error_model_2)
```
Error model med både NB og XGBoost signal
```{r}
error_model_3 <- lm(
  error_book ~
    pdiff_nb +
    I(pdiff_nb^2) +
    pdiff_xgb +
    I(pdiff_xgb^2) +
    lambda_diff_nb +
    lambda_sum_nb,
  data = match_train_eval
)

summary(error_model_3)
```
Error model med interaktioner
```{r}
error_model_4 <- lm(
  error_book ~
    pdiff_nb * fair_home +
    lambda_diff_nb * fair_home +
    lambda_sum_nb,
  data = match_train_eval
)

summary(error_model_4)
```
Lav korrigerede bookmaker forecasts i test
```{r}
match_test_eval <- match_test_eval %>%
  mutate(
    error_hat_1 = predict(error_model_1, newdata = .),
    error_hat_2 = predict(error_model_2, newdata = .),
    error_hat_3 = predict(error_model_3, newdata = .),
    error_hat_4 = predict(error_model_4, newdata = .),

    p_adj_1 = clip_prob(fair_home + error_hat_1),
    p_adj_2 = clip_prob(fair_home + error_hat_2),
    p_adj_3 = clip_prob(fair_home + error_hat_3),
    p_adj_4 = clip_prob(fair_home + error_hat_4)
  )
```

Sammenlign bookmaker mod korrigerede forecasts
```{r}
results_error_models <- tibble(
  model = c(
    "Bookmaker",
    "Adjusted bookmaker: error model 1",
    "Adjusted bookmaker: error model 2",
    "Adjusted bookmaker: error model 3",
    "Adjusted bookmaker: error model 4"
  ),
  log_loss = c(
    log_loss(match_test_eval$home_win.y, match_test_eval$fair_home),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_adj_1),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_adj_2),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_adj_3),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_adj_4)
  ),
  brier = c(
    brier_score(match_test_eval$home_win.y, match_test_eval$fair_home),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_adj_1),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_adj_2),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_adj_3),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_adj_4)
  )
)

results_error_models
```
Kalibrering: bookmaker vs bedste korrigerede model
```{r}
make_calibration_table <- function(actual, pred, bucket_width = 0.1) {
  tibble(actual = actual, pred = pred) %>%
    filter(!is.na(actual), !is.na(pred)) %>%
    mutate(
      prob_bucket = cut(
        pred,
        breaks = seq(0, 1, by = bucket_width),
        include.lowest = TRUE
      )
    ) %>%
    group_by(prob_bucket) %>%
    summarise(
      n = n(),
      avg_pred = mean(pred),
      actual_rate = mean(actual),
      .groups = "drop"
    )
}

cal_bookmaker <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$fair_home)
cal_adjusted  <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$p_adj_1)

cal_bookmaker
cal_adjusted
```
Hvor virker justeringen bedst?
```{r}
subgroup_error_results <- match_test_eval %>%
  mutate(
    fav_group = case_when(
      fair_home >= 0.60 ~ "Strong home favourite",
      fair_home >= 0.45 ~ "Moderate home favourite",
      fair_home >= 0.30 ~ "Balanced / slight away favourite",
      TRUE ~ "Strong away favourite"
    )
  ) %>%
  group_by(fav_group) %>%
  summarise(
    n = n(),
    book_log_loss = log_loss(home_win.y, fair_home),
    adj1_log_loss = log_loss(home_win.y, p_adj_1),
    adj2_log_loss = log_loss(home_win.y, p_adj_2),
    adj3_log_loss = log_loss(home_win.y, p_adj_3),
    adj4_log_loss = log_loss(home_win.y, p_adj_4),
    .groups = "drop"
  )

subgroup_error_results
```
Test om bookmaker-fejl er større i kampe med høj "uenighed" mellem jeres model og bookmaker
```{r}
match_test_eval <- match_test_eval %>%
  mutate(
    abs_error_book = abs(error_book),
    abs_edge_nb = abs(pdiff_nb),
    abs_edge_xgb = abs(pdiff_xgb)
  )

summary(lm(abs_error_book ~ abs_edge_nb, data = match_test_eval))
summary(lm(abs_error_book ~ abs_edge_xgb, data = match_test_eval))
summary(lm(abs_error_book ~ abs_edge_nb + abs_edge_xgb + fair_home, data = match_test_eval))
```
Tjek de kampe, hvor error-model justerer mest
```{r}
largest_adjustments <- match_test_eval %>%
  mutate(
    adjustment = p_adj_3 - fair_home
  ) %>%
  arrange(desc(abs(adjustment))) %>%
  select(
    date, home_team, away_team,
    home_goals, away_goals,
    fair_home, p_home_nb, p_home_xgb, p_adj_3,
    adjustment
  )

head(largest_adjustments, 20)
```

EKSTRA

Combine bookmaker + our model
```{r}
# ── Combine bookmaker + model ─────────────────────────

logit_model <- glm(
  home_win.y ~ fair_home + p_home_nb + p_home_xgb,
  family = binomial,
  data = match_train_eval
)

summary(logit_model)
```
Predictions (out-of-sample)
```{r}
# ── Predict on test set ───────────────────────────────

match_test_eval <- match_test_eval %>%
  mutate(
    p_comb = predict(logit_model, newdata = ., type = "response")
  )
```

Compare models
```{r}
# ── Model comparison ──────────────────────────────────

results_models <- tibble(
  model = c("Bookmaker", "NB model", "XGBoost", "Combined model"),
  log_loss = c(
    log_loss(match_test_eval$home_win.y, match_test_eval$fair_home),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_xgb),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_comb)
  ),
  brier = c(
    brier_score(match_test_eval$home_win.y, match_test_eval$fair_home),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_xgb),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_comb)
  )
)

print(results_models)
```
Segment analysis
```{r}
# ── Where do models differ? ───────────────────────────

segment_results <- match_test_eval %>%
  mutate(
    fav_group = case_when(
      fair_home > 0.6  ~ "Strong home favourite",
      fair_home > 0.45 ~ "Moderate home favourite",
      fair_home > 0.35 ~ "Balanced",
      TRUE             ~ "Away favourite"
    )
  ) %>%
  group_by(fav_group) %>%
  summarise(
    n = n(),
    book_log_loss = log_loss(home_win.y, fair_home),
    nb_log_loss   = log_loss(home_win.y, p_home_nb),
    xgb_log_loss  = log_loss(home_win.y, p_home_xgb),
    comb_log_loss = log_loss(home_win.y, p_comb)
  )

print(segment_results)
```
Edge analysis
```{r}
# ── Model disagreement (true edge test) ───────────────

match_test_eval <- match_test_eval %>%
  mutate(
    edge_nb  = p_home_nb  - fair_home,
    edge_xgb = p_home_xgb - fair_home
  )

edge_results <- match_test_eval %>%
  mutate(edge_group = cut(edge_nb,
                         breaks = c(-1, -0.1, -0.05, 0, 0.05, 0.1, 1))) %>%
  group_by(edge_group) %>%
  summarise(
    n = n(),
    actual_rate = mean(home_win.y),
    model_prob  = mean(p_home_nb),
    market_prob = mean(fair_home)
  )

print(edge_results)
```
ROI test
```{r}
# ── Betting strategy based on model disagreement ──────

bet_results <- map_dfr(c(0.02, 0.05, 0.08), function(t) {

  bets <- match_test_eval %>%
    filter(edge_nb > t) %>%
    mutate(
      odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, odds - 1, -1)
    )

  tibble(
    threshold = t,
    n_bets = nrow(bets),
    avg_profit = mean(bets$profit, na.rm = TRUE)
  )
})

print(bet_results)
```
EKSTRA

Calibration test af bookmaker
```{r}
calib <- match_test_eval %>%
  filter(!is.na(fair_home)) %>%
  mutate(bin = cut(fair_home, breaks = seq(0,1,0.1), include.lowest = TRUE)) %>%
  group_by(bin) %>%
  summarise(
    actual = mean(home_win.y),
    pred = mean(fair_home),
    n = n()
  )

print(calib)
```

Find edge-segmenter
```{r}
match_test_eval %>%
  mutate(
    mid = fair_home > 0.4 & fair_home < 0.6
  ) %>%
  group_by(mid) %>%
  summarise(
    book = log_loss(home_win.y, fair_home),
    model = log_loss(home_win.y, p_home_nb)
  )
```

Betting kun i segment
```{r}
bets <- match_test_eval %>%
  filter(fair_home > 0.4 & fair_home < 0.6) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets$profit)
```

Kombineret filter
```{r}
bets <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    p_home_nb > fair_home   # <-- vigtigt!
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets$profit)
nrow(bets)
```

Sammenlign med random
```{r}
mean(match_test_eval$home_win.y - match_test_eval$fair_home)
```

```{r}
# ── Create model-based strength measure ───────────────────────

match_test_eval <- match_test_eval %>%
  mutate(
    lambda_diff_nb = lambda_home_nb - lambda_away_nb
  )
```

```{r}
# ── Subgroup test inside mid-range odds segment ───────────────

mid_segment_summary <- match_test_eval %>%
  filter(fair_home > 0.4 & fair_home < 0.6) %>%
  mutate(
    high_strength = lambda_diff_nb > median(lambda_diff_nb, na.rm = TRUE)
  ) %>%
  group_by(high_strength) %>%
  summarise(
    actual = mean(home_win.y),
    pred = mean(fair_home),
    n = n(),
    .groups = "drop"
  )

print(mid_segment_summary)
```
```{r}
# ── Betting only in mid-range + high model strength ───────────

bets_strength <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6
  ) %>%
  mutate(
    lambda_diff_nb = lambda_home_nb - lambda_away_nb,
    high_strength = lambda_diff_nb > median(lambda_diff_nb, na.rm = TRUE)
  ) %>%
  filter(high_strength) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets_strength$profit, na.rm = TRUE)
nrow(bets_strength)
```

```{r}
# ── Mid-range + model disagreement ────────────────────────────

bets_disagreement <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    p_home_nb - fair_home > 0.03
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets_disagreement$profit, na.rm = TRUE)
nrow(bets_disagreement)
```

```{r}
# ── Mid-range + high total expected goals ─────────────────────

bets_total_goals <- match_test_eval %>%
  mutate(
    lambda_sum_nb = lambda_home_nb + lambda_away_nb
  ) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > median(lambda_sum_nb, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets_total_goals$profit, na.rm = TRUE)
nrow(bets_total_goals)
```

```{r}
bets_strict <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets_strict$profit)
nrow(bets_strict)
```

```{r}
bets_stricter <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.7, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets_stricter$profit)
nrow(bets_stricter)
```

```{r}
bets_low <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb < quantile(lambda_sum_nb, 0.4, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets_low$profit)
nrow(bets_low)
```

```{r}
match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  mutate(bin = ntile(lambda_sum_nb, 5)) %>%
  group_by(bin) %>%
  summarise(
    profit = mean(
      if_else(home_win.y == 1, 1/fair_home - 1, -1)
    ),
    n = n()
  )
```
Check for sample size i de vigtigste betting-segmenter
```{r}
# ── Sample sizes in betting segments ──────────────────────────

sample_size_checks <- tibble(
  segment = c(
    "Mid-range odds only",
    "Mid-range + high total goals (median split)",
    "Mid-range + top 40% total goals",
    "Mid-range + top 30% total goals",
    "Mid-range + low total goals"
  ),
  n = c(
    nrow(match_test_eval %>% filter(fair_home > 0.4 & fair_home < 0.6)),
    nrow(bets_total_goals),
    nrow(bets_strict),
    nrow(bets_stricter),
    nrow(bets_low)
  )
)

print(sample_size_checks)
```
Robustness: flere cutoffs for høj expected goal intensity
```{r}
# ── Robustness across different lambda_sum cutoffs ────────────

robustness_cutoffs <- c(0.50, 0.60, 0.70, 0.80)

robustness_table <- purrr::map_dfr(robustness_cutoffs, function(q) {
  
  bets_q <- match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, q, na.rm = TRUE)
    ) %>%
    mutate(
      odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, odds - 1, -1)
    )
  
  tibble(
    quantile_cutoff = q,
    n_bets = nrow(bets_q),
    avg_profit = mean(bets_q$profit, na.rm = TRUE)
  )
})

print(robustness_table)
```
Robustness: alternative mid-range definitions
```{r}
# ── Robustness across alternative probability ranges ──────────

range_results <- bind_rows(
  match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(fair_home > 0.35 & fair_home < 0.65,
           lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
    mutate(odds = 1 / fair_home,
           profit = if_else(home_win.y == 1, odds - 1, -1)) %>%
    summarise(range = "0.35-0.65", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  
  match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(fair_home > 0.40 & fair_home < 0.60,
           lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
    mutate(odds = 1 / fair_home,
           profit = if_else(home_win.y == 1, odds - 1, -1)) %>%
    summarise(range = "0.40-0.60", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  
  match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(fair_home > 0.45 & fair_home < 0.55,
           lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)) %>%
    mutate(odds = 1 / fair_home,
           profit = if_else(home_win.y == 1, odds - 1, -1)) %>%
    summarise(range = "0.45-0.55", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE))
)

print(range_results)
```
Heterogeneity: virker strategien ens på tværs af sæsoner?
```{r}
# ── Heterogeneity across seasons ──────────────────────────────

season_results <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  ) %>%
  group_by(season) %>%
  summarise(
    n_bets = n(),
    avg_profit = mean(profit, na.rm = TRUE),
    actual_home_win = mean(home_win.y),
    avg_book_prob = mean(fair_home),
    .groups = "drop"
  )

print(season_results)
```
Heterogeneity: home probability bins inden for high-goal segmentet
```{r}
# ── Heterogeneity within the mid-range segment ────────────────

mid_bin_results <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  mutate(
    prob_bin = cut(
      fair_home,
      breaks = c(0.4, 0.45, 0.5, 0.55, 0.6),
      include.lowest = TRUE
    ),
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  ) %>%
  group_by(prob_bin) %>%
  summarise(
    n_bets = n(),
    avg_profit = mean(profit, na.rm = TRUE),
    actual_home_win = mean(home_win.y),
    avg_book_prob = mean(fair_home),
    .groups = "drop"
  )

print(mid_bin_results)
```
Heterogeneity: compare high- and low-goal enviromnents directly
```{r}
# ── Direct comparison: high vs low expected goal intensity ────

high_low_comparison <- bind_rows(
  match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    mutate(
      odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, odds - 1, -1)
    ) %>%
    summarise(
      group = "High expected goals",
      n_bets = n(),
      avg_profit = mean(profit, na.rm = TRUE),
      actual_home_win = mean(home_win.y),
      avg_book_prob = mean(fair_home)
    ),
  
  match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb < quantile(lambda_sum_nb, 0.4, na.rm = TRUE)
    ) %>%
    mutate(
      odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, odds - 1, -1)
    ) %>%
    summarise(
      group = "Low expected goals",
      n_bets = n(),
      avg_profit = mean(profit, na.rm = TRUE),
      actual_home_win = mean(home_win.y),
      avg_book_prob = mean(fair_home)
    )
)

print(high_low_comparison)
```
Check for concentration risk: kommer profit fra få store gevinster?
```{r}
# ── Check whether returns are concentrated in a few bets ──────

bets_concentration <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

summary(bets_concentration$profit)

quantile(bets_concentration$profit, probs = c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1), na.rm = TRUE)
```
Check for statistical uncertainty: bootstrap mean profit
```{r}
# ── Bootstrap confidence interval for average profit ──────────

set.seed(123)

bets_boot <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

boot_means <- replicate(2000, {
  sample_profit <- sample(bets_boot$profit, size = nrow(bets_boot), replace = TRUE)
  mean(sample_profit, na.rm = TRUE)
})

quantile(boot_means, probs = c(0.025, 0.5, 0.975))
```

Placebo test
```{r}
# ── Placebo test: random bets of the same size ────────────────

set.seed(123)

target_n <- nrow(
  match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    )
)

all_mid_bets <- match_test_eval %>%
  filter(fair_home > 0.4 & fair_home < 0.6) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

placebo_means <- replicate(2000, {
  placebo_sample <- all_mid_bets %>%
    slice_sample(n = target_n, replace = FALSE)
  mean(placebo_sample$profit, na.rm = TRUE)
})

quantile(placebo_means, probs = c(0.025, 0.5, 0.975))

mean(
  match_test_eval %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    mutate(
      odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, odds - 1, -1)
    ) %>%
    pull(profit)
)
```

```{r}
# mindre aggressiv strategi
bets_wider <- match_test_eval %>%
  mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb) %>%
  filter(
    fair_home > 0.35 & fair_home < 0.65,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.5, na.rm = TRUE)
  ) %>%
  mutate(
    odds = 1 / fair_home,
    profit = if_else(home_win.y == 1, odds - 1, -1)
  )

mean(bets_wider$profit)
nrow(bets_wider)
t.test(bets_wider$profit)
```

