---
title: "Data prep"
author: "Jonathan Arve Thomsen"
date: "2026-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(readr)
library(dplyr)
library(purrr)
library(stringr)
library(lubridate)
```

Read files
```{r}
files <- c(
  "PL 16-17.csv",
  "PL 17-18.csv",
  "PL 18-19.csv",
  "PL 19-20.csv",
  "PL 20-21.csv",
  "PL 21-22.csv",
  "PL 22-23.csv",
  "PL 23-24.csv",
  "PL 24-25.csv"
)
```

Add variable: season
```{r}
pl_raw <- map_dfr(files, function(file) {
  read_csv(file, show_col_types = FALSE) %>%
    mutate(season = str_remove(file, "\\.csv$"))
})
```

Choose variables from data set
```{r}
pl <- pl_raw %>%
  select(
    Date, HomeTeam, AwayTeam,
    FTHG, FTAG, FTR,
    B365H, B365D, B365A,
    season
  ) %>%
  mutate(Date = trimws(Date))
```

Make real dates
```{r}
pl$Date <- ifelse(
  nchar(pl$Date) == 8,
  as.character(as.Date(pl$Date, format = "%d/%m/%y")),
  as.character(as.Date(pl$Date, format = "%d/%m/%Y"))
)

pl$Date <- as.Date(pl$Date)
```

Sort data set by date
```{r}
pl <- pl %>%
  arrange(Date)
```

Create readable variable
```{r}
pl <- pl %>%
  mutate(
    match_result = case_when(
      FTR == "H" ~ "HomeWin",
      FTR == "D" ~ "Draw",
      FTR == "A" ~ "AwayWin",
      TRUE ~ NA_character_
    )
  )
table(pl$match_result, useNA = "ifany")
```
Create numerical variable for result
```{r}
pl <- pl %>%
  mutate(
    result_num = case_when(
      FTR == "H" ~ 1,
      FTR == "D" ~ 0,
      FTR == "A" ~ -1,
      TRUE ~ NA_real_
    )
  )
```

Create total goals
```{r}
pl <- pl %>%
  mutate(
    total_goals = FTHG + FTAG
  )
summary(pl$total_goals)
```

Indicator for over 2.5 goals
```{r}
pl <- pl %>%
  mutate(
    over_2_5 = ifelse(total_goals > 2.5, 1, 0)
  )
table(pl$over_2_5, useNA = "ifany")
prop.table(table(pl$over_2_5))
```
Create bookmaker implied probabilities
```{r}
pl <- pl %>%
  mutate(
    p_home_raw = 1 / B365H,
    p_draw_raw = 1 / B365D,
    p_away_raw = 1 / B365A
  )
```

Adjusting for bookmaker margin
```{r}
pl <- pl %>%
  mutate(
    prob_sum = p_home_raw + p_draw_raw + p_away_raw,
    p_home = p_home_raw / prob_sum,
    p_draw = p_draw_raw / prob_sum,
    p_away = p_away_raw / prob_sum
  )
summary(pl$prob_sum)
summary(pl$p_home + pl$p_draw + pl$p_away)
```
Create binary home team variable
```{r}
pl <- pl %>%
  mutate(
    home_win = ifelse(FTR == "H", 1, 0)
  )
```

Filter out NA rows
```{r}
pl <- pl %>%
  filter(
    !is.na(Date),
    !is.na(HomeTeam),
    !is.na(AwayTeam),
    !is.na(FTHG),
    !is.na(FTAG),
    !is.na(FTR),
    !is.na(B365H),
    !is.na(B365D),
    !is.na(B365A)
  )
```

Descriptive statistics per season
```{r}
pl %>%
  group_by(season) %>%
  summarise(
    matches = n(),
    avg_goals = mean(total_goals, na.rm = TRUE),
    home_win_rate = mean(home_win, na.rm = TRUE),
    over25_rate = mean(over_2_5, na.rm = TRUE)
  )
```
Bookmaker benchmark
```{r}
pl %>%
  mutate(
    home_prob_bucket = cut(
      p_home,
      breaks = c(0, 0.3, 0.5, 0.7, 1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(home_prob_bucket) %>%
  summarise(
    n = n(),
    avg_book_prob = mean(p_home, na.rm = TRUE),
    actual_home_win = mean(home_win, na.rm = TRUE)
  )
```

