---
title: "final outputs"
output: html_document
date: "2026-04-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read all files from other documents
```{r}
library(dplyr)
library(readr)
library(tibble)
library(purrr)
library(lubridate)
library(gt)
library(scales)

fix_bad_years <- function(x) {
  x <- as.Date(x)
  if_else(year(x) < 100, x %m+% years(2000), x)
}

# Load data
pl <- readRDS("pl_match_test_eval.rds")
ll <- readRDS("ll_match_test_eval.rds")
sa <- readRDS("sa_match_test_eval.rds")
bl <- readRDS("bl_match_test_eval.rds")

pooled <- bind_rows(
  pl %>% mutate(league = "PL"),
  ll %>% mutate(league = "LL"),
  sa %>% mutate(league = "SA"),
  bl %>% mutate(league = "BL")
)
```

MAIN PAPER OUTPUTS

Data overview table
```{r}
data_overview_table <- bind_rows(
  readRDS("pl_clean.rds") %>% mutate(league = "PL"),
  readRDS("ll_clean.rds") %>% mutate(league = "LL"),
  readRDS("sa_clean.rds") %>% mutate(league = "SA"),
  readRDS("bl_clean.rds") %>% mutate(league = "BL")
) %>%
  mutate(Date = fix_bad_years(Date)) %>%
  group_by(league) %>%
  summarise(
    bookmaker_matches = n(),
    first_date = min(Date, na.rm = TRUE),
    last_date = max(Date, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    bind_rows(
      readRDS("pl_team_match_df.rds") %>% mutate(league = "PL"),
      readRDS("ll_team_match_df.rds") %>% mutate(league = "LL"),
      readRDS("sa_team_match_df.rds") %>% mutate(league = "SA"),
      readRDS("bl_team_match_df.rds") %>% mutate(league = "BL")
    ) %>%
      group_by(league) %>%
      summarise(team_match_rows = n(), .groups = "drop"),
    by = "league"
  ) %>%
  left_join(
    pooled %>%
      group_by(league) %>%
      summarise(test_matches = n(), .groups = "drop"),
    by = "league"
  ) %>%
  left_join(
    bets_main_pooled %>%
      group_by(league) %>%
      summarise(main_strategy_bets = n(), .groups = "drop"),
    by = "league"
  )

data_overview_table %>%
  gt() %>%
  tab_header(title = "Data overview by league") %>%
  cols_label(
    league = "League",
    bookmaker_matches = "Bookmaker matches",
    first_date = "First date",
    last_date = "Last date",
    team_match_rows = "Team-match rows",
    test_matches = "Test matches",
    main_strategy_bets = "Main strategy bets"
  ) %>%
  fmt_date(columns = c(first_date, last_date), date_style = "iso") %>%
  tab_options(
    table.font.size = 13,
    data_row.padding = px(4)
  )
```
Pooled forecast performance
```{r}
forecast_performance_table <- results_pooled %>%
  mutate(
    log_loss = round(log_loss, 4),
    brier = round(brier, 4)
  )

forecast_performance_table %>%
  gt() %>%
  tab_header(
    title = "Pooled forecast performance",
    subtitle = "Comparison of model forecast accuracy across all leagues"
  ) %>%
  cols_label(
    model = "Model",
    log_loss = "Log loss",
    brier = "Brier score"
  ) %>%
  fmt_number(
    columns = c(log_loss, brier),
    decimals = 4
  ) %>%
  tab_options(
    table.font.size = 13,
    data_row.padding = px(4),
    column_labels.font.weight = "bold"
  )
```
Diebold-Mariano test
```{r}
library(forecast)

dm_nb <- dm.test(
  e1 = (pooled$home_win.y - pooled$p_home_nb)^2,
  e2 = (pooled$home_win.y - pooled$fair_home)^2,
  h = 1,
  power = 2
)

dm_xgb <- dm.test(
  e1 = (pooled$home_win.y - pooled$p_home_xgb)^2,
  e2 = (pooled$home_win.y - pooled$fair_home)^2,
  h = 1,
  power = 2
)

dm_table <- tibble(
  comparison = c("NB model vs Bookmaker", "XGBoost model vs Bookmaker"),
  dm_statistic = c(as.numeric(dm_nb$statistic), as.numeric(dm_xgb$statistic)),
  p_value = c(dm_nb$p.value, dm_xgb$p.value)
) %>%
  mutate(
    dm_statistic = round(dm_statistic, 3),
    p_value = signif(p_value, 3)
  )

dm_table %>%
  gt() %>%
  tab_header(
    title = "Diebold-Mariano test",
    subtitle = "Forecast accuracy compared with bookmaker probabilities"
  ) %>%
  cols_label(
    comparison = "Comparison",
    dm_statistic = "DM statistic",
    p_value = "p-value"
  ) %>%
  fmt_number(
    columns = dm_statistic,
    decimals = 3
  ) %>%
  fmt_number(
    columns = p_value,
    decimals = 3
  ) %>%
  tab_options(
    table.font.size = 13,
    data_row.padding = px(4),
    column_labels.font.weight = "bold"
  )
```
Main betting strategy with bootstrap CI
```{r}
main_strategy_table <- tibble(
  strategy = "Mid-range odds + high expected goal intensity",
  avg_profit = mean(bets_main_pooled$profit, na.rm = TRUE),
  n_bets = nrow(bets_main_pooled),
  ci_lower = quantile(boot_means_pooled, 0.025, na.rm = TRUE),
  ci_median = quantile(boot_means_pooled, 0.5, na.rm = TRUE),
  ci_upper = quantile(boot_means_pooled, 0.975, na.rm = TRUE)
) %>%
  mutate(
    avg_profit = round(avg_profit, 4),
    ci_lower = round(ci_lower, 4),
    ci_median = round(ci_median, 4),
    ci_upper = round(ci_upper, 4)
  )

main_strategy_table %>%
  gt() %>%
  tab_header(
    title = "Main betting strategy",
    subtitle = "Average profit with bootstrap confidence interval"
  ) %>%
  cols_label(
    strategy = "Strategy",
    avg_profit = "Average profit",
    n_bets = "Number of bets",
    ci_lower = "CI lower",
    ci_median = "CI median",
    ci_upper = "CI upper"
  ) %>%
  fmt_number(
    columns = c(avg_profit, ci_lower, ci_median, ci_upper),
    decimals = 4
  ) %>%
  tab_options(
    table.font.size = 13,
    data_row.padding = px(4),
    column_labels.font.weight = "bold"
  )
```
League heterogeneity table
```{r}
league_heterogeneity_table <- league_results %>%
  mutate(
    diff_actual_vs_book = actual_home_win - avg_book_prob,
    avg_profit = round(avg_profit, 4),
    actual_home_win = round(actual_home_win, 4),
    avg_book_prob = round(avg_book_prob, 4),
    diff_actual_vs_book = round(diff_actual_vs_book, 4)
  )

league_heterogeneity_table %>%
  gt() %>%
  tab_header(
    title = "League heterogeneity",
    subtitle = "Differences in bookmaker probabilities, realised outcomes, and betting performance"
  ) %>%
  cols_label(
    league = "League",
    n_bets = "Number of bets",
    avg_profit = "Average profit",
    actual_home_win = "Actual home win rate",
    avg_book_prob = "Average bookmaker probability",
    diff_actual_vs_book = "Actual minus bookmaker"
  ) %>%
  fmt_number(
    columns = where(is.numeric),
    decimals = 4
  ) %>%
  tab_options(
    table.font.size = 13,
    data_row.padding = px(4),
    column_labels.font.weight = "bold"
  )
```
Robustness across expected-goal cutoffs
```{r}
robustness_cutoff_table <- robustness_table_pooled %>%
  mutate(avg_profit = round(avg_profit, 4))

robustness_cutoff_table %>%
  gt() %>%
  tab_header(
    title = "Robustness across expected-goal cutoffs",
    subtitle = "Average betting profit under alternative cutoff values"
  ) %>%
  cols_label(
    quantile_cutoff = "Expected-goal cutoff",
    n_bets = "Number of bets",
    avg_profit = "Average profit"
  ) %>%
  fmt_number(
    columns = avg_profit,
    decimals = 4
  ) %>%
  tab_options(
    table.font.size = 13,
    data_row.padding = px(4),
    column_labels.font.weight = "bold"
  )
```

APPENDIX OUTPUTS
```{r}
make_appendix_table <- function(data, title, subtitle = NULL) {
  data %>%
    gt() %>%
    tab_header(
      title = title,
      subtitle = subtitle
    ) %>%
    tab_options(
      table.font.size = 11,
      data_row.padding = px(3),
      column_labels.font.weight = "bold"
    )
}
```

1. Bookmaker data cleaning summary
```{r}
bookmaker_cleaning_table <- bind_rows(
  readRDS("pl_clean.rds") %>% mutate(league = "Premier League"),
  readRDS("ll_clean.rds") %>% mutate(league = "La Liga"),
  readRDS("sa_clean.rds") %>% mutate(league = "Serie A"),
  readRDS("bl_clean.rds") %>% mutate(league = "Bundesliga")
) %>%
  mutate(Date = fix_bad_years(Date)) %>%
  group_by(league) %>%
  summarise(
    rows = n(),
    first_date = min(Date, na.rm = TRUE),
    last_date = max(Date, na.rm = TRUE),
    missing_odds = sum(is.na(B365H) | is.na(B365D) | is.na(B365A)),
    .groups = "drop"
  )

make_appendix_table(
  bookmaker_cleaning_table,
  "Bookmaker data cleaning summary"
) %>%
  cols_label(
    league = "League",
    rows = "Rows",
    first_date = "First date",
    last_date = "Last date",
    missing_odds = "Missing odds"
  ) %>%
  fmt_date(columns = c(first_date, last_date), date_style = "iso")
```
Understat construction summary
```{r}
understat_cleaning_table <- bind_rows(
  readRDS("pl_team_match_df.rds") %>% mutate(league = "Premier League"),
  readRDS("ll_team_match_df.rds") %>% mutate(league = "La Liga"),
  readRDS("sa_team_match_df.rds") %>% mutate(league = "Serie A"),
  readRDS("bl_team_match_df.rds") %>% mutate(league = "Bundesliga")
) %>%
  group_by(league) %>%
  summarise(
    team_match_rows = n(),
    first_date = min(date, na.rm = TRUE),
    last_date = max(date, na.rm = TRUE),
    teams = n_distinct(team),
    .groups = "drop"
  )

make_appendix_table(
  understat_cleaning_table,
  "Understat construction summary"
) %>%
  cols_label(
    league = "League",
    team_match_rows = "Team-match rows",
    first_date = "First date",
    last_date = "Last date",
    teams = "Teams"
  ) %>%
  fmt_date(columns = c(first_date, last_date), date_style = "iso")
```
Poisson and Negative Binomial comparison
```{r}
poisson_nb_table <- tibble(
  statistic = c(
    "Pearson dispersion",
    "Likelihood ratio statistic",
    "Likelihood ratio p-value",
    "Poisson AIC",
    "Negative Binomial AIC"
  ),
  value = c(
    pearson_disp,
    lrt_stat,
    lrt_pval,
    AIC(poisson_full),
    AIC(nb_for_test)
  )
) %>%
  mutate(value = round(value, 4))

make_appendix_table(
  poisson_nb_table,
  "Poisson and Negative Binomial comparison"
) %>%
  cols_label(
    statistic = "Statistic",
    value = "Value"
  )
```
Rolling mean and exponential weighting comparison
```{r}
weighting_comparison_table <- tibble(
  weighting = c("Simple rolling mean", "Exponential weighting"),
  aic = c(AIC(nb_full), AIC(nb_exp)),
  rmse = c(rmse_simple, rmse_exp)
) %>%
  mutate(
    aic = round(aic, 2),
    rmse = round(rmse, 5)
  )

make_appendix_table(
  weighting_comparison_table,
  "Rolling mean and exponential weighting comparison"
) %>%
  cols_label(
    weighting = "Weighting method",
    aic = "AIC",
    rmse = "RMSE"
  )
```
Standard deviation xg inclusion check
```{r}
sd_xg_table <- tibble(
  model = c("Without SD xG", "With SD xG"),
  aic = c(AIC(nb_no_sd), AIC(nb_with_sd))
) %>%
  mutate(aic = round(aic, 2))

make_appendix_table(
  sd_xg_table,
  "Standard deviation xG inclusion check"
) %>%
  cols_label(
    model = "Model",
    aic = "AIC"
  )
```
Home advantage model comparison
```{r}
home_advantage_table <- tibble(
  model = c("Without home advantage", "With home advantage"),
  aic = c(AIC(nb_no_home), AIC(nb_with_sd))
) %>%
  mutate(aic = round(aic, 2))

make_appendix_table(
  home_advantage_table,
  "Home advantage model comparison"
) %>%
  cols_label(
    model = "Model",
    aic = "AIC"
  )
```
Home advantage coefficient
```{r}
home_coefficient_table <- tibble(
  variable = "Home advantage",
  estimate = home_coef["Estimate"],
  p_value = home_coef["Pr(>|z|)"]
) %>%
  mutate(
    estimate = round(estimate, 4),
    p_value = signif(p_value, 3)
  )

make_appendix_table(
  home_coefficient_table,
  "Home advantage coefficient"
) %>%
  cols_label(
    variable = "Variable",
    estimate = "Estimate",
    p_value = "p-value"
  )
```
Variable reduction summary
```{r}
variable_reduction_table <- tibble(
  model = c(
    "Full model",
    "Without corner xG",
    "Without corner and penalty xG",
    "Without corner, penalty and setpiece xG"
  ),
  aic = c(
    AIC(nb_with_sd),
    AIC(nb_no_corner),
    AIC(nb_no_corner_penalty),
    AIC(nb_no_setpiece)
  ),
  variables_removed = c(
    "None",
    "Corner",
    "Corner, penalty",
    "Corner, penalty, setpiece"
  )
) %>%
  mutate(
    aic = round(aic, 2),
    aic_change = round(aic - first(aic), 2)
  )

make_appendix_table(
  variable_reduction_table,
  "Variable reduction summary"
) %>%
  cols_label(
    model = "Model",
    aic = "AIC",
    variables_removed = "Variables removed",
    aic_change = "AIC change"
  )
```
Final Negative Binomial model coefficients
```{r}
final_nb_coef_table <- as.data.frame(coef(summary(nb_no_setpiece))) %>%
  rownames_to_column("variable") %>%
  as_tibble() %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

make_appendix_table(
  final_nb_coef_table,
  "Final Negative Binomial model coefficients"
)
```
Model development journey
```{r}
development_journey_table <- tibble(
  step = c(
    "Poisson with simple rolling means and all xG types",
    "Negative Binomial with simple rolling means and all xG types",
    "Negative Binomial with exponential weighting and all xG types",
    "Negative Binomial with simple rolling means and SD xG",
    "Reduced Negative Binomial with open-play, free-kick, SD xG and home advantage"
  ),
  aic = c(
    AIC(poisson_full),
    AIC(nb_for_test),
    AIC(nb_exp),
    AIC(nb_with_sd),
    AIC(nb_no_setpiece)
  ),
  reason = c(
    "Baseline count model",
    "Overdispersion favours Negative Binomial",
    "Tested but not retained",
    "SD xG tested as a consistency measure (no AIC improvement)",
    "Corner, penalty and setpiece xG removed by AIC"
  )
) %>%
  mutate(aic = round(aic, 2))

make_appendix_table(
  development_journey_table,
  "Model development journey"
) %>%
  cols_label(
    step = "Model step",
    aic = "AIC",
    reason = "Reason"
  )
```
LASSO variable selection
```{r}
lasso_selection_table <- comparison_table %>%
  select(
    variable,
    nb_selected,
    lasso_min_sel,
    lasso_1se_sel
  )

make_appendix_table(
  lasso_selection_table,
  "LASSO variable selection"
) %>%
  cols_label(
    variable = "Variable",
    nb_selected = "Selected in NB model",
    lasso_min_sel = "Selected by LASSO lambda.min",
    lasso_1se_sel = "Selected by LASSO lambda.1se"
  )
```
LASSO agreement with Negative Binomial selection
```{r}
lasso_agreement_table <- comparison_table %>%
  summarise(
    agreement_lambda_min = mean(agree_min == "✓"),
    agreement_lambda_1se = mean(agree_1se == "✓")
  ) %>%
  mutate(
    agreement_lambda_min = round(agreement_lambda_min, 3),
    agreement_lambda_1se = round(agreement_lambda_1se, 3)
  )

make_appendix_table(
  lasso_agreement_table,
  "LASSO agreement with Negative Binomial selection"
) %>%
  cols_label(
    agreement_lambda_min = "Agreement with lambda.min",
    agreement_lambda_1se = "Agreement with lambda.1se"
  )
```
LASSO and Negative Binomial prediction performance
```{r}
lasso_performance_table <- tibble(
  model = c("Negative Binomial reduced", "LASSO lambda.min", "LASSO lambda.1se"),
  rmse = c(
    sqrt(mean((y_test - test_df$pred_final_nb)^2)),
    sqrt(mean((y_test - pred_lasso_min)^2)),
    sqrt(mean((y_test - pred_lasso_1se)^2))
  ),
  mae = c(
    mean(abs(y_test - test_df$pred_final_nb)),
    mean(abs(y_test - pred_lasso_min)),
    mean(abs(y_test - pred_lasso_1se))
  )
) %>%
  mutate(
    rmse = round(rmse, 5),
    mae = round(mae, 5)
  )

make_appendix_table(
  lasso_performance_table,
  "LASSO and Negative Binomial prediction performance"
) %>%
  cols_label(
    model = "Model",
    rmse = "RMSE",
    mae = "MAE"
  )
```

Random Forest and XGBoost prediction performance
```{r}
rf_xgb_table <- tibble(
  model = c("Random Forest", "XGBoost"),
  rmse = c(
    sqrt(mean((test_df$goals - pred_rf)^2)),
    sqrt(mean((test_df$goals - pred_xgb_comp)^2))
  ),
  mae = c(
    mean(abs(test_df$goals - pred_rf)),
    mean(abs(test_df$goals - pred_xgb_comp))
  ),
  bias = c(
    mean(pred_rf - test_df$goals),
    mean(pred_xgb_comp - test_df$goals)
  )
) %>%
  mutate(
    rmse = round(rmse, 5),
    mae = round(mae, 5),
    bias = round(bias, 5)
  )

make_appendix_table(
  rf_xgb_table,
  "Random Forest and XGBoost prediction performance"
) %>%
  cols_label(
    model = "Model",
    rmse = "RMSE",
    mae = "MAE",
    bias = "Bias"
  )
```
League-specific forecast performance
```{r}
make_league_forecast_results <- function(df, league_name) {
  tibble(
    league = league_name,
    model = c("Bookmaker", "Negative Binomial", "XGBoost"),
    log_loss = c(
      log_loss(df$home_win.y, df$fair_home),
      log_loss(df$home_win.y, df$p_home_nb),
      log_loss(df$home_win.y, df$p_home_xgb)
    ),
    brier = c(
      brier_score(df$home_win.y, df$fair_home),
      brier_score(df$home_win.y, df$p_home_nb),
      brier_score(df$home_win.y, df$p_home_xgb)
    )
  )
}

league_forecast_table <- bind_rows(
  make_league_forecast_results(pl, "Premier League"),
  make_league_forecast_results(ll, "La Liga"),
  make_league_forecast_results(sa, "Serie A"),
  make_league_forecast_results(bl, "Bundesliga")
) %>%
  mutate(
    log_loss = round(log_loss, 4),
    brier = round(brier, 4)
  )

make_appendix_table(
  league_forecast_table,
  "League-specific forecast performance"
) %>%
  cols_label(
    league = "League",
    model = "Model",
    log_loss = "Log loss",
    brier = "Brier score"
  )
```
League-specific betting strategy results
```{r}
make_league_betting_result <- function(df, league_name) {
  df <- df %>%
    mutate(lambda_sum_nb = lambda_home_nb + lambda_away_nb)
  
  bets <- df %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    add_profit()
  
  tibble(
    league = league_name,
    avg_profit = mean(bets$profit, na.rm = TRUE),
    n_bets = nrow(bets),
    actual_home_win = mean(bets$home_win.y, na.rm = TRUE),
    avg_book_prob = mean(bets$fair_home, na.rm = TRUE)
  )
}

league_betting_table <- bind_rows(
  make_league_betting_result(pl, "Premier League"),
  make_league_betting_result(ll, "La Liga"),
  make_league_betting_result(sa, "Serie A"),
  make_league_betting_result(bl, "Bundesliga")
) %>%
  mutate(
    avg_profit = round(avg_profit, 4),
    actual_home_win = round(actual_home_win, 4),
    avg_book_prob = round(avg_book_prob, 4)
  )

make_appendix_table(
  league_betting_table,
  "League-specific betting strategy results"
) %>%
  cols_label(
    league = "League",
    avg_profit = "Average profit",
    n_bets = "Number of bets",
    actual_home_win = "Actual home win rate",
    avg_book_prob = "Average bookmaker probability"
  )
```
Pooled calibration table
```{r}
calibration_combined <- bind_rows(
  cal_book_pooled %>% mutate(model = "Bookmaker"),
  cal_nb_pooled %>% mutate(model = "Negative Binomial"),
  cal_xgb_pooled %>% mutate(model = "XGBoost")
) %>%
  select(model, everything()) %>%
  mutate(
    avg_pred = round(avg_pred, 4),
    actual_rate = round(actual_rate, 4)
  )

make_appendix_table(
  calibration_combined,
  "Pooled calibration table"
) %>%
  cols_label(
    model = "Model",
    avg_pred = "Average predicted probability",
    actual_rate = "Actual home win rate"
  )
```
Pooled subgroup analysis
```{r}
subgroup_table <- subgroup_results_pooled %>%
  mutate(
    book_log_loss = round(book_log_loss, 4),
    nb_log_loss = round(nb_log_loss, 4),
    xgb_log_loss = round(xgb_log_loss, 4)
  )

make_appendix_table(
  subgroup_table,
  "Pooled subgroup analysis"
) %>%
  cols_label(
    book_log_loss = "Bookmaker log loss",
    nb_log_loss = "Negative Binomial log loss",
    xgb_log_loss = "XGBoost log loss"
  )
```
Robustness across alternative probability ranges
```{r}
probability_range_robustness_table <- range_results_pooled %>%
  mutate(avg_profit = round(avg_profit, 4))

make_appendix_table(
  probability_range_robustness_table,
  "Robustness across alternative probability ranges"
) %>%
  cols_label(
    avg_profit = "Average profit"
  )
```
Season-level betting heterogeneity
```{r}
season_heterogeneity_table <- season_results_pooled %>%
  mutate(
    avg_profit = round(avg_profit, 4),
    actual_home_win = round(actual_home_win, 4),
    avg_book_prob = round(avg_book_prob, 4)
  )

make_appendix_table(
  season_heterogeneity_table,
  "Season-level betting heterogeneity"
) %>%
  cols_label(
    avg_profit = "Average profit",
    actual_home_win = "Actual home win rate",
    avg_book_prob = "Average bookmaker probability"
  )
```
High and low expected-goal comparison
```{r}
high_low_table <- high_low_comparison_pooled %>%
  mutate(
    diff_actual_vs_book = actual_home_win - avg_book_prob,
    avg_profit = round(avg_profit, 4),
    actual_home_win = round(actual_home_win, 4),
    avg_book_prob = round(avg_book_prob, 4),
    diff_actual_vs_book = round(diff_actual_vs_book, 4)
  )

make_appendix_table(
  high_low_table,
  "High and low expected-goal comparison"
) %>%
  cols_label(
    avg_profit = "Average profit",
    actual_home_win = "Actual home win rate",
    avg_book_prob = "Average bookmaker probability",
    diff_actual_vs_book = "Actual minus bookmaker"
  )
```
Return concentration
```{r}
return_concentration_table <- tibble(
  statistic = c("Minimum", "10%", "25%", "Median", "75%", "90%", "Maximum"),
  profit = as.numeric(
    quantile(
      bets_concentration_pooled$profit,
      probs = c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1),
      na.rm = TRUE
    )
  )
) %>%
  mutate(profit = round(profit, 4))

make_appendix_table(
  return_concentration_table,
  "Return concentration"
) %>%
  cols_label(
    statistic = "Statistic",
    profit = "Profit"
  )
```
Placebo test
```{r}
placebo_table <- tibble(
  placebo_ci_lower = quantile(placebo_means, 0.025, na.rm = TRUE),
  placebo_median = quantile(placebo_means, 0.5, na.rm = TRUE),
  placebo_ci_upper = quantile(placebo_means, 0.975, na.rm = TRUE),
  actual_mean = mean(bets_main_pooled$profit, na.rm = TRUE)
) %>%
  mutate(across(everything(), ~ round(.x, 4)))

make_appendix_table(
  placebo_table,
  "Placebo test"
) %>%
  cols_label(
    placebo_ci_lower = "Placebo CI lower",
    placebo_median = "Placebo median",
    placebo_ci_upper = "Placebo CI upper",
    actual_mean = "Actual mean profit"
  )
```
Placebo test by league
```{r}
placebo_by_league_table <- placebo_by_league %>%
  mutate(
    actual_mean = round(actual_mean, 4),
    placebo_pvalue = round(placebo_pvalue, 4)
  )

make_appendix_table(
  placebo_by_league_table,
  "Placebo test by league"
) %>%
  cols_label(
    actual_mean = "Actual mean profit",
    placebo_pvalue = "Placebo p-value"
  )
```
Wider betting strategy check
```{r}
wider_strategy_test <- t.test(bets_wider_pooled$profit)

wider_strategy_table <- tibble(
  strategy = "Wider betting strategy",
  avg_profit = mean(bets_wider_pooled$profit, na.rm = TRUE),
  n_bets = nrow(bets_wider_pooled),
  t_statistic = as.numeric(wider_strategy_test$statistic),
  p_value = wider_strategy_test$p.value
) %>%
  mutate(
    avg_profit = round(avg_profit, 4),
    t_statistic = round(t_statistic, 3),
    p_value = signif(p_value, 3)
  )

make_appendix_table(
  wider_strategy_table,
  "Wider betting strategy check"
) %>%
  cols_label(
    strategy = "Strategy",
    avg_profit = "Average profit",
    n_bets = "Number of bets",
    t_statistic = "t-statistic",
    p_value = "p-value"
  )
```
ANOVA test of profit differences across leagues
```{r}
anova_model <- aov(profit ~ league, data = bets_main_pooled)
anova_summary <- summary(anova_model)[[1]]

anova_table <- tibble(
  test = "Profit differences across leagues",
  f_statistic = anova_summary$`F value`[1],
  p_value = anova_summary$`Pr(>F)`[1]
) %>%
  mutate(
    f_statistic = round(f_statistic, 3),
    p_value = signif(p_value, 3)
  )

make_appendix_table(
  anova_table,
  "ANOVA test of profit differences across leagues"
) %>%
  cols_label(
    test = "Test",
    f_statistic = "F-statistic",
    p_value = "p-value"
  )
```
High versus low expected-goal t-test
```{r}
high_low_ttest <- t.test(profit ~ group, data = compare_high_low)

high_low_ttest_table <- tibble(
  test = "High versus low expected goals",
  t_statistic = as.numeric(high_low_ttest$statistic),
  p_value = high_low_ttest$p.value
) %>%
  mutate(
    t_statistic = round(t_statistic, 3),
    p_value = signif(p_value, 3)
  )

make_appendix_table(
  high_low_ttest_table,
  "High versus low expected-goal t-test"
) %>%
  cols_label(
    test = "Test",
    t_statistic = "t-statistic",
    p_value = "p-value"
  )
```

