---
title: "LASSO PÅ ALLE VARIABLER"
output: html_document
date: "2026-04-22"
---
```{r}
library(worldfootballR)
library(tidyverse)
library(zoo)
library(MASS)
library(xgboost)
library(corrplot)
library(gt)
library(glmnet)
library(randomForest)

select <- dplyr::select
filter <- dplyr::filter
```

Helper chunk
```{r}
make_appendix_table <- function(data, title, subtitle = NULL) {
  data %>%
    gt() %>%
    tab_header(
      title = title,
      subtitle = subtitle
    ) %>%
    tab_options(
      table.font.size = 11,
      data_row.padding = px(3),
      column_labels.font.weight = "bold"
    )
}
```

Load and aggregate Understat data
```{r}
epl_shots <- load_understat_league_shots(league = "EPL")

match_xg <- epl_shots %>%
  filter(season >= 2016, season <= 2024) %>%
  mutate(h_a = coalesce(h_a, home_away)) %>%
  group_by(match_id, league, season, date, home_team, away_team, home_goals, away_goals) %>%
  summarise(
    home_xG_openplay = sum(xG[h_a == "h" & situation == "OpenPlay"],       na.rm = TRUE),
    away_xG_openplay = sum(xG[h_a == "a" & situation == "OpenPlay"],       na.rm = TRUE),
    home_xG_corner   = sum(xG[h_a == "h" & situation == "FromCorner"],     na.rm = TRUE),
    away_xG_corner   = sum(xG[h_a == "a" & situation == "FromCorner"],     na.rm = TRUE),
    home_xG_freekick = sum(xG[h_a == "h" & situation == "DirectFreekick"], na.rm = TRUE),
    away_xG_freekick = sum(xG[h_a == "a" & situation == "DirectFreekick"], na.rm = TRUE),
    home_xG_penalty  = sum(xG[h_a == "h" & situation == "Penalty"],        na.rm = TRUE),
    away_xG_penalty  = sum(xG[h_a == "a" & situation == "Penalty"],        na.rm = TRUE),
    home_xG_setpiece = sum(xG[h_a == "h" & situation == "SetPiece"],       na.rm = TRUE),
    away_xG_setpiece = sum(xG[h_a == "a" & situation == "SetPiece"],       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(date)
```

Construct team-level modelling data
```{r}
model_df <- bind_rows(
  match_xg %>% transmute(
    match_id, season, date,
    team        = home_team,
    opponent    = away_team,
    home_away   = "home",
    goals       = home_goals,
    xG_openplay = home_xG_openplay,
    xG_corner   = home_xG_corner,
    xG_freekick = home_xG_freekick,
    xG_penalty  = home_xG_penalty,
    xG_setpiece = home_xG_setpiece
  ),
  match_xg %>% transmute(
    match_id, season, date,
    team        = away_team,
    opponent    = home_team,
    home_away   = "away",
    goals       = away_goals,
    xG_openplay = away_xG_openplay,
    xG_corner   = away_xG_corner,
    xG_freekick = away_xG_freekick,
    xG_penalty  = away_xG_penalty,
    xG_setpiece = away_xG_setpiece
  )
) %>%
  arrange(team, date)
```

Rolling xG features and multicollinearity check
```{r}
model_df <- model_df %>%
  group_by(team) %>%
  mutate(
    roll_xG_openplay = lag(rollmean(xG_openplay, 5, fill = NA, align = "right"), 1),
    roll_xG_corner   = lag(rollmean(xG_corner,   5, fill = NA, align = "right"), 1),
    roll_xG_freekick = lag(rollmean(xG_freekick, 5, fill = NA, align = "right"), 1),
    roll_xG_penalty  = lag(rollmean(xG_penalty,  5, fill = NA, align = "right"), 1),
    roll_xG_setpiece = lag(rollmean(xG_setpiece, 5, fill = NA, align = "right"), 1)
  ) %>%
  ungroup()

team_match_df <- model_df %>%
  left_join(
    model_df %>%
      select(match_id, opponent = team,
             opp_roll_xG_openplay = roll_xG_openplay,
             opp_roll_xG_corner   = roll_xG_corner,
             opp_roll_xG_freekick = roll_xG_freekick,
             opp_roll_xG_penalty  = roll_xG_penalty,
             opp_roll_xG_setpiece = roll_xG_setpiece),
    by = c("match_id", "opponent")
  ) %>%
  mutate(home_flag = if_else(home_away == "home", 1, 0)) %>%
  select(
    match_id, season, date, team, opponent, home_away, home_flag, goals,
    roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
    roll_xG_penalty,  roll_xG_setpiece,
    opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
    opp_roll_xG_penalty,  opp_roll_xG_setpiece
  ) %>%
  drop_na() %>%
  arrange(date)

cat("Modelling rows:", nrow(team_match_df), "\n")
cat("Seasons:", paste(unique(team_match_df$season), collapse = ", "), "\n")

feature_cols <- team_match_df %>%
  select(
    roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
    roll_xG_penalty,  roll_xG_setpiece,
    opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
    opp_roll_xG_penalty,  opp_roll_xG_setpiece
  )

cor_matrix <- cor(feature_cols)
corrplot(cor_matrix, method = "number", type = "upper",
         tl.cex = 0.7, number.cex = 0.7,
         title = "Feature Correlation Matrix — All xG Types")

high_cor <- which(abs(cor_matrix) > 0.8 & cor_matrix != 1, arr.ind = TRUE)
if (nrow(high_cor) > 0) {
  cat("High correlations (>0.8) found:\n")
  print(high_cor)
} else {
  cat("No high correlations — safe to include all variables\n")
}
```

Train-test split
```{r}
train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "| Seasons:", min(train_df$season), "-", max(train_df$season), "\n")
cat("Test rows: ", nrow(test_df),  "| Seasons:", min(test_df$season),  "-", max(test_df$season),  "\n")
```

Poisson baseline and overdispersion test
```{r}
poisson_full <- glm(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece +
    home_flag,
  data   = train_df,
  family = poisson(link = "log")
)

summary(poisson_full)

test_df <- test_df %>%
  mutate(pred_poisson = predict(poisson_full, newdata = test_df, type = "response"))

cat("\n=== Poisson Baseline Evaluation ===\n")
test_df %>%
  summarise(
    rmse = sqrt(mean((goals - pred_poisson)^2)),
    mae  = mean(abs(goals - pred_poisson)),
    bias = mean(pred_poisson - goals)
  ) %>%
  print()

pearson_disp <- sum(residuals(poisson_full, type = "pearson")^2) / poisson_full$df.residual
cat("\n=== Overdispersion Check ===\n")
cat("Pearson dispersion statistic:", round(pearson_disp, 4), "\n")
cat("Value of 1 = perfect Poisson fit\n")
cat("Values > 1 indicate overdispersion\n")
```

Negative Binomial model comparison
```{r}
nb_for_test <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece +
    home_flag,
  data = train_df
)

lrt_stat <- 2 * (as.numeric(logLik(nb_for_test)) - as.numeric(logLik(poisson_full)))
lrt_pval <- pchisq(lrt_stat, df = 1, lower.tail = FALSE) / 2


cat("\n=== Likelihood Ratio Test: Poisson vs Negative Binomial ===\n")
cat("LRT statistic:", round(lrt_stat, 4), "\n")
cat("p-value (one-sided):", round(lrt_pval, 6), "\n")
cat("Theta (NB dispersion):", round(nb_for_test$theta, 4), "\n")
cat("If p < 0.05 → reject Poisson in favour of NB\n")
cat("High theta (>20) → modest overdispersion; NB still preferred\n")
cat("as it nests Poisson and AIC will confirm whether it is worth keeping\n")

cat("\n=== AIC: Poisson vs NB (full model) ===\n")
cat("Poisson AIC:", AIC(poisson_full),  "\n")
cat("NB AIC:     ", AIC(nb_for_test),   "\n")
cat("Lower is better\n")

nb_full <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece +
    home_flag,
  data = train_df
)

summary(nb_full)
cat("\nTheta:", round(nb_full$theta, 4), "\n")

test_df <- test_df %>%
  mutate(pred_nb_full = predict(nb_full, newdata = test_df, type = "response"))

cat("\n=== NB Full Model Evaluation ===\n")
test_df %>%
  summarise(
    rmse = sqrt(mean((goals - pred_nb_full)^2)),
    mae  = mean(abs(goals - pred_nb_full)),
    bias = mean(pred_nb_full - goals)
  ) %>%
  print()
```

Rolling mean versus exponential weighting
```{r}
exp_weights <- exp(seq(0, 1, length.out = 5))
exp_weights <- exp_weights / sum(exp_weights)
cat("\nExponential weights (oldest → most recent):\n")
print(round(exp_weights, 3))

wmean5 <- function(x) weighted.mean(x, w = exp_weights)

model_df_exp <- model_df %>%
  group_by(team) %>%
  mutate(
    roll_xG_openplay = lag(rollapply(xG_openplay, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_corner   = lag(rollapply(xG_corner,   5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_freekick = lag(rollapply(xG_freekick, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_penalty  = lag(rollapply(xG_penalty,  5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_setpiece = lag(rollapply(xG_setpiece, 5, FUN = wmean5, fill = NA, align = "right"), 1)
  ) %>%
  ungroup()

team_match_exp <- model_df_exp %>%
  left_join(
    model_df_exp %>%
      select(match_id, opponent = team,
             opp_roll_xG_openplay = roll_xG_openplay,
             opp_roll_xG_corner   = roll_xG_corner,
             opp_roll_xG_freekick = roll_xG_freekick,
             opp_roll_xG_penalty  = roll_xG_penalty,
             opp_roll_xG_setpiece = roll_xG_setpiece),
    by = c("match_id", "opponent")
  ) %>%
  mutate(home_flag = if_else(home_away == "home", 1, 0)) %>%
  select(match_id, season, date, team, opponent, home_away, home_flag, goals,
         roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
         roll_xG_penalty,  roll_xG_setpiece,
         opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
         opp_roll_xG_penalty,  opp_roll_xG_setpiece) %>%
  drop_na() %>%
  arrange(date)

train_exp <- team_match_exp %>% filter(season <= 2022)
test_exp  <- team_match_exp %>% filter(season >= 2023)

nb_exp <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece +
    home_flag,
  data = train_exp
)

test_exp <- test_exp %>%
  mutate(pred_nb_exp = predict(nb_exp, newdata = test_exp, type = "response"))

cat("\n=== Exponential vs Simple Rolling Mean Comparison ===\n")
cat("Simple rolling mean NB — AIC:", AIC(nb_full), "\n")
cat("Exponential weighted NB — AIC:", AIC(nb_exp),  "\n")

rmse_simple <- sqrt(mean((test_df$goals - test_df$pred_nb_full)^2))
rmse_exp    <- sqrt(mean((test_exp$goals - test_exp$pred_nb_exp)^2))
cat("Simple rolling mean RMSE:", round(rmse_simple, 5), "\n")
cat("Exponential weighted RMSE:", round(rmse_exp,    5), "\n")
cat("AIC prefers simple rolling mean — exponential weighting not adopted\n")
cat("Simple rolling means are retained as the final feature construction\n")
```

SD xG and home advantage checks
```{r}
model_df <- model_df %>%
  group_by(team) %>%
  mutate(
    sd_xG = lag(rollapply(
      xG_openplay + xG_corner + xG_freekick + xG_penalty + xG_setpiece,
      5, FUN = sd, fill = NA, align = "right"), 1)
  ) %>%
  ungroup()

team_match_df <- model_df %>%
  left_join(
    model_df %>%
      select(match_id, opponent = team,
             opp_roll_xG_openplay = roll_xG_openplay,
             opp_roll_xG_corner   = roll_xG_corner,
             opp_roll_xG_freekick = roll_xG_freekick,
             opp_roll_xG_penalty  = roll_xG_penalty,
             opp_roll_xG_setpiece = roll_xG_setpiece,
             opp_sd_xG            = sd_xG),
    by = c("match_id", "opponent")
  ) %>%
  mutate(home_flag = if_else(home_away == "home", 1, 0)) %>%
  select(match_id, season, date, team, opponent, home_away, home_flag, goals,
         roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
         roll_xG_penalty,  roll_xG_setpiece, sd_xG,
         opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
         opp_roll_xG_penalty,  opp_roll_xG_setpiece, opp_sd_xG) %>%
  drop_na() %>%
  arrange(date)

train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

nb_no_sd <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece +
    home_flag,
  data = train_df
)

nb_with_sd <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece + sd_xG +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece + opp_sd_xG +
    home_flag,
  data = train_df
)

cat("\n=== AIC: Without vs With SD xG ===\n")
cat("Without SD xG AIC:", AIC(nb_no_sd),   "\n")
cat("With SD xG AIC:   ", AIC(nb_with_sd), "\n")
cat("Lower is better — if difference > 2, SD xG is worth keeping\n")

nb_no_home <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece + sd_xG +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece + opp_sd_xG,
  data = train_df
)

cat("\n=== AIC: Without vs With Home Flag ===\n")
cat("Without home flag AIC:", AIC(nb_no_home),   "\n")
cat("With home flag AIC:   ", AIC(nb_with_sd),   "\n")

home_coef <- coef(summary(nb_with_sd))["home_flag", ]
cat("\nHome flag coefficient:", round(home_coef["Estimate"], 4), "\n")
cat("p-value:              ", round(home_coef["Pr(>|z|)"], 6), "\n")
```

Variable reduction
```{r}
cat("\n=== Full Model Coefficients (p-values) ===\n")
print(round(coef(summary(nb_with_sd)), 4))

nb_no_corner <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece + sd_xG +
    opp_roll_xG_openplay + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece + opp_sd_xG +
    home_flag,
  data = train_df
)

cat("\n=== Stage 1: Remove corner xG ===\n")
cat("Full model AIC:          ", AIC(nb_with_sd),  "\n")
cat("Without corner xG AIC:   ", AIC(nb_no_corner), "\n")
cat("If AIC does not increase by >2 → corner xG not needed\n")

nb_no_corner_penalty <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_freekick +
    roll_xG_setpiece + sd_xG +
    opp_roll_xG_openplay + opp_roll_xG_freekick +
    opp_roll_xG_setpiece + opp_sd_xG +
    home_flag,
  data = train_df
)

cat("\n=== Stage 2: Remove penalty xG ===\n")
cat("Without corner AIC:              ", AIC(nb_no_corner),          "\n")
cat("Without corner + penalty AIC:    ", AIC(nb_no_corner_penalty),  "\n")
cat("If AIC does not increase by >2 → penalty xG not needed\n")

nb_no_setpiece <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_freekick +
    sd_xG +
    opp_roll_xG_openplay + opp_roll_xG_freekick +
    opp_sd_xG +
    home_flag,
  data = train_df
)

cat("\n=== Stage 3: Remove setpiece xG ===\n")
cat("Without corner + penalty AIC:         ", AIC(nb_no_corner_penalty), "\n")
cat("Without corner + penalty + setpiece:  ", AIC(nb_no_setpiece),       "\n")
cat("If AIC does not increase by >2 → setpiece xG not needed\n")

cat("\n=== Variable Reduction Summary ===\n")
tibble(
  model        = c("Full (all 5 xG types)", "Remove corner",
                   "Remove corner + penalty", "Remove corner + penalty + setpiece"),
  aic          = c(AIC(nb_with_sd), AIC(nb_no_corner),
                   AIC(nb_no_corner_penalty), AIC(nb_no_setpiece)),
  variables_removed = c("—", "corner",
                        "corner, penalty", "corner, penalty, setpiece")
) %>%
  mutate(aic_change = round(aic - first(aic), 2)) %>%
  print()

cat("\nFinal retained variables: openplay, freekick, SD xG, home flag\n")

cat("\n=== Final Reduced NB Model Summary ===\n")
summary(nb_no_setpiece)
cat("\nTheta:", round(nb_no_setpiece$theta, 4), "\n")

test_df <- test_df %>%
  mutate(pred_final_nb = predict(nb_no_setpiece, 
                                 newdata = test_df, 
                                 type = "response"))

cat("\n=== Distribution Check: Actual vs Final NB Model ===\n")

test_df %>%
  count(goals) %>%
  mutate(pct_actual = round(n / sum(n), 3)) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded = round(pred_final_nb)) %>%
      count(pred_rounded) %>%
      mutate(pct_nb_final = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded),
    by = "goals"
  ) %>%
  print()
```

Final NB model evaluation
```{r}
cat("\n=== Final NB Model Evaluation ===\n")
test_df %>%
  summarise(
    rmse_nb_final = sqrt(mean((goals - pred_final_nb)^2)),
    mae_nb_final  = mean(abs(goals - pred_final_nb)),
    bias_nb_final = mean(pred_final_nb - goals)
  ) %>%
  print()

cat("\n=== Development Journey Summary ===\n")

tibble(
  step  = c(
    "1. Poisson (simple rolling, all 5 xG types)",
    "2. NB (simple rolling, all 5 xG types)",
    "3. NB (exponential weighting tested)",
    "4. NB (simple rolling + SD xG)",
    "5. NB reduced (openplay + freekick + SD + home)"
  ),
  aic   = c(
    AIC(poisson_full),
    AIC(nb_for_test),
    AIC(nb_exp),
    AIC(nb_with_sd),
    AIC(nb_no_setpiece)
  ),
  reason = c(
    "Baseline — Poisson is natural for count data",
    "Overdispersion test rejects Poisson",
    "No improvement — exponential weighting rejected",
    "SD xG tested as consistency measure",
    "Corner, penalty, setpiece not significant by AIC"
  )
) %>%
  print()

cat("\n=== Final variables: openplay, freekick, SD xG, home flag ===\n")
```

LASSO validation
```{r}
lasso_features <- c(
  "roll_xG_openplay", "roll_xG_corner", "roll_xG_freekick",
  "roll_xG_penalty",  "roll_xG_setpiece", "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_corner", "opp_roll_xG_freekick",
  "opp_roll_xG_penalty",  "opp_roll_xG_setpiece", "opp_sd_xG",
  "home_flag"
)

x_train <- as.matrix(train_df[, lasso_features])
y_train <- train_df$goals
x_test  <- as.matrix(test_df[, lasso_features])
y_test  <- test_df$goals

set.seed(123)
lasso_cv <- cv.glmnet(
  x      = x_train,
  y      = y_train,
  family = "poisson",
  alpha  = 1,
  nfolds = 10
)

cat("\n=== LASSO Cross-Validation ===\n")
cat("Lambda min (lowest CV error):       ", round(lasso_cv$lambda.min, 6), "\n")
cat("Lambda 1se (sparsest within 1 SE):  ", round(lasso_cv$lambda.1se, 6), "\n")

coefs_min <- coef(lasso_cv, s = "lambda.min") %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  setNames(c("variable", "lasso_min")) %>%
  filter(variable != "(Intercept)") %>%
  mutate(lasso_min = round(lasso_min, 4))

coefs_1se <- coef(lasso_cv, s = "lambda.1se") %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  setNames(c("variable", "lasso_1se")) %>%
  filter(variable != "(Intercept)") %>%
  mutate(lasso_1se = round(lasso_1se, 4))

nb_coef_df <- data.frame(
  variable = names(coef(nb_no_setpiece))[-1],
  nb_coef  = round(coef(nb_no_setpiece)[-1], 4)
)

cat("\n=== Variable Selection Comparison: NB vs LASSO ===\n")
cat("Zero in LASSO columns = variable removed by penalisation\n")
cat("NA in NB column = variable not selected by AIC reduction\n\n")

comparison_table <- data.frame(variable = lasso_features) %>%
  left_join(nb_coef_df,  by = "variable") %>%
  left_join(coefs_min,   by = "variable") %>%
  left_join(coefs_1se,   by = "variable") %>%
  mutate(
    nb_selected   = if_else(!is.na(nb_coef), "Yes", "No"),
    lasso_min_sel = if_else(lasso_min != 0,  "Yes", "No"),
    lasso_1se_sel = if_else(lasso_1se != 0,  "Yes", "No"),
    agree_min     = if_else(nb_selected == lasso_min_sel, "✓", "✗"),
    agree_1se     = if_else(nb_selected == lasso_1se_sel, "✓", "✗")
  )

print(comparison_table)

cat("\n=== Agreement Summary (lambda.min) ===\n")

cat("Both NB and LASSO selected:\n")
comparison_table %>%
  filter(nb_selected == "Yes", lasso_min_sel == "Yes") %>%
  pull(variable) %>% paste(collapse = ", ") %>% cat("\n")

cat("\nBoth NB and LASSO dropped:\n")
comparison_table %>%
  filter(nb_selected == "No", lasso_min_sel == "No") %>%
  pull(variable) %>% paste(collapse = ", ") %>% cat("\n")

cat("\nDisagreements (NB and LASSO differ):\n")
disagreements <- comparison_table %>% filter(agree_min == "✗")
if (nrow(disagreements) == 0) {
  cat("None — full agreement\n")
} else {
  disagreements %>% select(variable, nb_selected, lasso_min_sel) %>% print()
}

pred_lasso_min <- as.numeric(predict(lasso_cv, newx = x_test,
                                     s = "lambda.min", type = "response"))
pred_lasso_1se <- as.numeric(predict(lasso_cv, newx = x_test,
                                     s = "lambda.1se", type = "response"))

cat("\n=== Prediction Performance: NB vs LASSO ===\n")
tibble(
  model = c("NB reduced (AIC selection)", "LASSO (lambda.min)", "LASSO (lambda.1se)"),
  rmse  = c(
    round(sqrt(mean((y_test - test_df$pred_final_nb)^2)),  5),
    round(sqrt(mean((y_test - pred_lasso_min)^2)),         5),
    round(sqrt(mean((y_test - pred_lasso_1se)^2)),         5)
  ),
  mae   = c(
    round(mean(abs(y_test - test_df$pred_final_nb)), 5),
    round(mean(abs(y_test - pred_lasso_min)),        5),
    round(mean(abs(y_test - pred_lasso_1se)),        5)
  )
) %>%
  print()

cat("\nSimilar RMSE across models → same signal captured regardless of method\n")
cat("This provides a robustness check for the NB variable selection\n")
```

Random Forest and XGBoost comparison
```{r}
rf_features <- c(
  "roll_xG_openplay", "roll_xG_freekick", "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_freekick", "opp_sd_xG",
  "home_flag"
)

set.seed(123)
rf_model <- randomForest(
  x      = train_df[, rf_features],
  y      = train_df$goals,
  ntree  = 500,
  mtry   = floor(sqrt(length(rf_features))),
  importance = TRUE
)

cat("\n=== Random Forest Model Summary ===\n")
print(rf_model)

pred_rf <- predict(rf_model, newdata = test_df[, rf_features])

xgb_features <- rf_features

train_matrix_xgb <- xgb.DMatrix(
  data  = as.matrix(train_df[, xgb_features]),
  label = train_df$goals
)

test_matrix_xgb <- xgb.DMatrix(
  data  = as.matrix(test_df[, xgb_features]),
  label = test_df$goals
)

set.seed(123)
xgb_model_comp <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    eval_metric      = "poisson-nloglik"
  ),
  data                  = train_matrix_xgb,
  nrounds               = 500,
  watchlist             = list(train = train_matrix_xgb, test = test_matrix_xgb),
  early_stopping_rounds = 30,
  verbose               = 0
)

pred_xgb_comp <- predict(xgb_model_comp, test_matrix_xgb)

cat("\n=== Random Forest vs XGBoost: Head to Head ===\n")
cat("Same variable set, same train/test split\n")
cat("Any performance difference is due to model structure only\n\n")

tibble(
  model = c("Random Forest", "XGBoost"),
  rmse  = c(
    round(sqrt(mean((test_df$goals - pred_rf)^2)),       5),
    round(sqrt(mean((test_df$goals - pred_xgb_comp)^2)), 5)
  ),
  mae   = c(
    round(mean(abs(test_df$goals - pred_rf)),       5),
    round(mean(abs(test_df$goals - pred_xgb_comp)), 5)
  ),
  bias  = c(
    round(mean(pred_rf       - test_df$goals), 5),
    round(mean(pred_xgb_comp - test_df$goals), 5)
  )
) %>%
  print()

cat("\n=== Feature Importance: Random Forest ===\n")
importance_rf <- importance(rf_model)
print(round(importance_rf, 4))

cat("\n=== Feature Importance: XGBoost ===\n")
importance_xgb <- xgb.importance(feature_names = xgb_features, model = xgb_model_comp)
print(importance_xgb)

cat("\n=== Interpretation ===\n")
cat("If XGBoost RMSE < RF RMSE → boosting outperforms bagging on this dataset\n")
cat("This compares XGBoost with Random Forest using the same feature set\n")
cat("Similar feature importance rankings → both methods agree on what matters\n")
```

Final outputs for appendix:

Correlation matrix
```{r}
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  number.cex = 0.6,
  tl.cex = 0.7,
  tl.col = "black",
  tl.srt = 45,
  col = colorRampPalette(c("red", "white", "blue"))(200),
  title = "Correlation matrix of model predictors",
  mar = c(0, 0, 2, 0)
)
```
Random Forest feature importance
```{r}
rf_importance_table <- importance(rf_model) %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  as_tibble() %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  arrange(desc(`%IncMSE`))

make_appendix_table(
  rf_importance_table,
  "Random Forest feature importance"
) %>%
  cols_label(
    feature = "Feature",
    `%IncMSE` = "% increase in MSE",
    IncNodePurity = "Increase in node purity"
  )
```
XGBoost feature importance
```{r}
xgb_importance_table <- importance_xgb %>%
  as_tibble() %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  arrange(desc(Gain))

make_appendix_table(
  xgb_importance_table,
  "XGBoost feature importance"
) %>%
  cols_label(
    Feature = "Feature",
    Gain = "Gain",
    Cover = "Cover",
    Frequency = "Frequency"
  )
```

