---
title: "INDEPENDENT PL DONE"
output: html_document
date: "2026-04-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(MASS)
library(xgboost)
```

Helper functions
```{r}
log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

clip_prob <- function(p) {
  pmin(pmax(p, 1e-6), 1 - 1e-6)
}

make_fair_probs <- function(home_odds, draw_odds, away_odds) {
  raw_home <- 1 / home_odds
  raw_draw <- 1 / draw_odds
  raw_away <- 1 / away_odds
  
  overround <- raw_home + raw_draw + raw_away
  
  tibble(
    fair_home = raw_home / overround,
    fair_draw = raw_draw / overround,
    fair_away = raw_away / overround
  )
}

match_probs <- function(lambda_home, lambda_away, max_goals = 10) {
  g <- 0:max_goals
  score_matrix <- outer(dpois(g, lambda_home), dpois(g, lambda_away))
  
  list(
    prob_home = sum(score_matrix[lower.tri(score_matrix)]),
    prob_draw = sum(diag(score_matrix)),
    prob_away = sum(score_matrix[upper.tri(score_matrix)])
  )
}

make_calibration_table <- function(actual, pred, bucket_width = 0.1) {
  tibble(actual = actual, pred = pred) %>%
    filter(!is.na(actual), !is.na(pred)) %>%
    mutate(
      prob_bucket = cut(
        pred,
        breaks = seq(0, 1, by = bucket_width),
        include.lowest = TRUE
      )
    ) %>%
    group_by(prob_bucket) %>%
    summarise(
      n = n(),
      avg_pred = mean(pred),
      actual_rate = mean(actual),
      .groups = "drop"
    )
}

add_profit <- function(df) {
  df %>%
    mutate(
      odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, odds - 1, -1)
    )
}
```

Load data
```{r}
pl <- readRDS("pl_clean.rds")
team_match_df <- readRDS("epl_team_match_df.rds")
```

Align team names and clean bookmaker data
```{r}
team_map <- c(
  "Man City" = "Manchester City",
  "Man United" = "Manchester United",
  "Newcastle" = "Newcastle United",
  "Nott'm Forest" = "Nottingham Forest",
  "West Brom" = "West Bromwich Albion",
  "Wolves" = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map),
    Date = as.Date(Date),
    home_win = as.integer(FTR == "H")
  )
```

Create bookmaker fair probabilities
```{r}
fair_probs <- make_fair_probs(pl$B365H, pl$B365D, pl$B365A)

pl <- bind_cols(pl, fair_probs)
```

Keep only required columns
```{r}
feature_names <- c(
  "roll_xG_openplay",
  "roll_xG_freekick",
  "opp_roll_xG_openplay",
  "opp_roll_xG_freekick",
  "sd_xG",
  "opp_sd_xG",
  "home_flag"
)

required_cols <- c(
  "match_id", "season", "date", "team", "opponent", "goals", feature_names
)

team_match_df <- team_match_df %>%
  select(all_of(required_cols)) %>%
  mutate(date = as.Date(date)) %>%
  drop_na()
```

Train/test split
```{r}
train_df <- team_match_df %>%
  filter(season <= 2022)

test_df <- team_match_df %>%
  filter(season >= 2023)

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows: ", nrow(test_df), "\n")
```

Fit Negative Binomial model
```{r}
nb_model <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

summary(nb_model)
cat("Theta:", nb_model$theta, "\n")
```

Predict team-level expected goals with NB
```{r}
train_df <- train_df %>%
  mutate(pred_nb = predict(nb_model, newdata = train_df, type = "response"))

test_df <- test_df %>%
  mutate(pred_nb = predict(nb_model, newdata = test_df, type = "response"))
```

Fit XGBoost model
```{r}
train_matrix <- xgb.DMatrix(
  data = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

watchlist <- list(
  train = train_matrix,
  test  = test_matrix
)

xgb_model <- xgb.train(
  params = list(
    objective = "count:poisson",
    eta = 0.05,
    max_depth = 4,
    subsample = 0.8,
    colsample_bytree = 0.8,
    eval_metric = "poisson-nloglik"
  ),
  data = train_matrix,
  nrounds = 500,
  watchlist = watchlist,
  early_stopping_rounds = 30,
  verbose = 1,
  print_every_n = 50
)
```
Predict team-level expected goals with XGBoost
```{r}
train_df <- train_df %>%
  mutate(pred_xgb = predict(xgb_model, train_matrix))

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))
```

Convert team-level predictions to match-level rows: train
```{r}
home_train <- train_df %>%
  filter(home_flag == 1) %>%
  select(
    match_id, date, season,
    home_team = team,
    away_team = opponent,
    home_goals = goals,
    lambda_home_nb = pred_nb,
    lambda_home_xgb = pred_xgb
  )

away_train <- train_df %>%
  filter(home_flag == 0) %>%
  select(
    match_id,
    away_goals = goals,
    lambda_away_nb = pred_nb,
    lambda_away_xgb = pred_xgb
  )

match_train <- home_train %>%
  left_join(away_train, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals > away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      TRUE ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )
```

Convet team-level predictions to match-level rows: test
```{r}
home_test <- test_df %>%
  filter(home_flag == 1) %>%
  select(
    match_id, date, season,
    home_team = team,
    away_team = opponent,
    home_goals = goals,
    lambda_home_nb = pred_nb,
    lambda_home_xgb = pred_xgb
  )

away_test <- test_df %>%
  filter(home_flag == 0) %>%
  select(
    match_id,
    away_goals = goals,
    lambda_away_nb = pred_nb,
    lambda_away_xgb = pred_xgb
  )

match_test <- home_test %>%
  left_join(away_test, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals > away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      TRUE ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )
```

Convert expected goals to match probabilities: train
```{r}
probs_train_nb <- pmap_dfr(
  list(match_train$lambda_home_nb, match_train$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_nb = p$prob_home,
      p_draw_nb = p$prob_draw,
      p_away_nb = p$prob_away
    )
  }
)

probs_train_xgb <- pmap_dfr(
  list(match_train$lambda_home_xgb, match_train$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_xgb = p$prob_home,
      p_draw_xgb = p$prob_draw,
      p_away_xgb = p$prob_away
    )
  }
)

match_train <- bind_cols(match_train, probs_train_nb, probs_train_xgb)
```

Convert expected goals to match probabilities: test
```{r}
probs_test_nb <- pmap_dfr(
  list(match_test$lambda_home_nb, match_test$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_nb = p$prob_home,
      p_draw_nb = p$prob_draw,
      p_away_nb = p$prob_away
    )
  }
)

probs_test_xgb <- pmap_dfr(
  list(match_test$lambda_home_xgb, match_test$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(
      p_home_xgb = p$prob_home,
      p_draw_xgb = p$prob_draw,
      p_away_xgb = p$prob_away
    )
  }
)

match_test <- bind_cols(match_test, probs_test_nb, probs_test_xgb)
```
Merge bookmaker benchmark into train sample
```{r}
book_train <- pl %>%
  filter(Date < as.Date("2023-08-01")) %>%
  select(
    Date, HomeTeam, AwayTeam, home_win,
    fair_home, fair_draw, fair_away
  )

match_train <- match_train %>%
  left_join(
    book_train,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )
```

Merge bookmaker benchmark into test sample
```{r}
book_test <- pl %>%
  filter(Date >= as.Date("2023-08-01")) %>%
  select(
    Date, HomeTeam, AwayTeam, home_win,
    fair_home, fair_draw, fair_away
  )

match_test <- match_test %>%
  left_join(
    book_test,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )
```

Keep only matched rows
```{r}
match_train_eval <- match_train %>%
  filter(!is.na(fair_home))

match_test_eval <- match_test %>%
  filter(!is.na(fair_home))
```

Main out-of-sample comparison
```{r}
results_main <- tibble(
  model = c("Bookmaker", "NB model", "XGBoost model"),
  log_loss = c(
    log_loss(match_test_eval$home_win.y, match_test_eval$fair_home),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    log_loss(match_test_eval$home_win.y, match_test_eval$p_home_xgb)
  ),
  brier = c(
    brier_score(match_test_eval$home_win.y, match_test_eval$fair_home),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_nb),
    brier_score(match_test_eval$home_win.y, match_test_eval$p_home_xgb)
  )
)

print(results_main)
```
Calibration tables
```{r}
cal_book <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$fair_home)
cal_nb   <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$p_home_nb)
cal_xgb  <- make_calibration_table(match_test_eval$home_win.y, match_test_eval$p_home_xgb)

cal_book
cal_nb
cal_xgb
```

Subgroup analysis: favourites vs non-favourites
```{r}
subgroup_results <- match_test_eval %>%
  mutate(
    fav_group = case_when(
      fair_home >= 0.60 ~ "Strong home favourite",
      fair_home >= 0.45 ~ "Moderate home favourite",
      fair_home >= 0.30 ~ "Balanced / slight away favourite",
      TRUE ~ "Strong away favourite"
    )
  ) %>%
  group_by(fav_group) %>%
  summarise(
    n = n(),
    book_log_loss = log_loss(home_win.y, fair_home),
    nb_log_loss   = log_loss(home_win.y, p_home_nb),
    xgb_log_loss  = log_loss(home_win.y, p_home_xgb),
    .groups = "drop"
  )

print(subgroup_results)
```
Create betting-strategy variables
```{r}
match_test_eval <- match_test_eval %>%
  mutate(
    lambda_diff_nb = lambda_home_nb - lambda_away_nb,
    lambda_sum_nb  = lambda_home_nb + lambda_away_nb
  )
```

Main betting strategy: mid-range odds + high expected goal intensity
```{r}
bets_main <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  add_profit()

mean(bets_main$profit, na.rm = TRUE)
nrow(bets_main)
```

Betting strategy: median split
```{r}
bets_total_goals <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > median(lambda_sum_nb, na.rm = TRUE)
  ) %>%
  add_profit()

mean(bets_total_goals$profit, na.rm = TRUE)
nrow(bets_total_goals)
```

Betting strategy: top 40 percent
```{r}
bets_strict <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  add_profit()

mean(bets_strict$profit, na.rm = TRUE)
nrow(bets_strict)
```

Betting strategy: top 30 percent
```{r}
bets_stricter <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.7, na.rm = TRUE)
  ) %>%
  add_profit()

mean(bets_stricter$profit, na.rm = TRUE)
nrow(bets_stricter)
```

Betting strategy: low expected goals
```{r}
bets_low <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb < quantile(lambda_sum_nb, 0.4, na.rm = TRUE)
  ) %>%
  add_profit()

mean(bets_low$profit, na.rm = TRUE)
nrow(bets_low)
```

Profit by expected goal intensity quintile
```{r}
profit_by_bin <- match_test_eval %>%
  mutate(bin = ntile(lambda_sum_nb, 5)) %>%
  add_profit() %>%
  group_by(bin) %>%
  summarise(
    avg_profit = mean(profit, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

print(profit_by_bin)
```
Sample size checks
```{r}
sample_size_checks <- tibble(
  segment = c(
    "Mid-range odds only",
    "Mid-range + high total goals (median split)",
    "Mid-range + top 40% total goals",
    "Mid-range + top 30% total goals",
    "Mid-range + low total goals"
  ),
  n = c(
    nrow(match_test_eval %>% filter(fair_home > 0.4 & fair_home < 0.6)),
    nrow(bets_total_goals),
    nrow(bets_strict),
    nrow(bets_stricter),
    nrow(bets_low)
  )
)

print(sample_size_checks)
```
Robustness across different expected-goal cutoffs
```{r}
robustness_cutoffs <- c(0.50, 0.60, 0.70, 0.80)

robustness_table <- purrr::map_dfr(robustness_cutoffs, function(q) {
  
  bets_q <- match_test_eval %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, q, na.rm = TRUE)
    ) %>%
    add_profit()
  
  tibble(
    quantile_cutoff = q,
    n_bets = nrow(bets_q),
    avg_profit = mean(bets_q$profit, na.rm = TRUE)
  )
})

print(robustness_table)
```
Robustness across alternative probability ranges
```{r}
range_results <- bind_rows(
  match_test_eval %>%
    filter(
      fair_home > 0.35 & fair_home < 0.65,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    add_profit() %>%
    summarise(range = "0.35-0.65", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  
  match_test_eval %>%
    filter(
      fair_home > 0.40 & fair_home < 0.60,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    add_profit() %>%
    summarise(range = "0.40-0.60", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  
  match_test_eval %>%
    filter(
      fair_home > 0.45 & fair_home < 0.55,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    add_profit() %>%
    summarise(range = "0.45-0.55", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE))
)

print(range_results)
```
Heterogeneity across seasons
```{r}
season_results <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  add_profit() %>%
  group_by(season) %>%
  summarise(
    n_bets = n(),
    avg_profit = mean(profit, na.rm = TRUE),
    actual_home_win = mean(home_win.y),
    avg_book_prob = mean(fair_home),
    .groups = "drop"
  )

print(season_results)
```
Heterogeneity within the mid-range segment
```{r}
mid_bin_results <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  add_profit() %>%
  mutate(
    prob_bin = cut(
      fair_home,
      breaks = c(0.4, 0.45, 0.5, 0.55, 0.6),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bin) %>%
  summarise(
    n_bets = n(),
    avg_profit = mean(profit, na.rm = TRUE),
    actual_home_win = mean(home_win.y),
    avg_book_prob = mean(fair_home),
    .groups = "drop"
  )

print(mid_bin_results)
```
Direct comparison: high vs low expected goals
```{r}
high_low_comparison <- bind_rows(
  match_test_eval %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    add_profit() %>%
    summarise(
      group = "High expected goals",
      n_bets = n(),
      avg_profit = mean(profit, na.rm = TRUE),
      actual_home_win = mean(home_win.y),
      avg_book_prob = mean(fair_home)
    ),
  
  match_test_eval %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb < quantile(lambda_sum_nb, 0.4, na.rm = TRUE)
    ) %>%
    add_profit() %>%
    summarise(
      group = "Low expected goals",
      n_bets = n(),
      avg_profit = mean(profit, na.rm = TRUE),
      actual_home_win = mean(home_win.y),
      avg_book_prob = mean(fair_home)
    )
)

print(high_low_comparison)
```
Check whether returns are concentrated in a few bets
```{r}
bets_concentration <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  add_profit()

summary(bets_concentration$profit)

quantile(
  bets_concentration$profit,
  probs = c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1),
  na.rm = TRUE
)
```
Bootstrap confidence interval for average profit
```{r}
set.seed(123)

bets_boot <- match_test_eval %>%
  filter(
    fair_home > 0.4 & fair_home < 0.6,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
  ) %>%
  add_profit()

boot_means <- replicate(2000, {
  sample_profit <- sample(bets_boot$profit, size = nrow(bets_boot), replace = TRUE)
  mean(sample_profit, na.rm = TRUE)
})

quantile(boot_means, probs = c(0.025, 0.5, 0.975))
```

Placebo test
```{r}
set.seed(123)

target_n <- nrow(
  match_test_eval %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    )
)

all_mid_bets <- match_test_eval %>%
  filter(fair_home > 0.4 & fair_home < 0.6) %>%
  add_profit()

placebo_means <- replicate(2000, {
  placebo_sample <- all_mid_bets %>%
    slice_sample(n = target_n, replace = FALSE)
  mean(placebo_sample$profit, na.rm = TRUE)
})

quantile(placebo_means, probs = c(0.025, 0.5, 0.975))

mean(
  match_test_eval %>%
    filter(
      fair_home > 0.4 & fair_home < 0.6,
      lambda_sum_nb > quantile(lambda_sum_nb, 0.6, na.rm = TRUE)
    ) %>%
    add_profit() %>%
    pull(profit)
)
```

Wider strategy check
```{r}
bets_wider <- match_test_eval %>%
  filter(
    fair_home > 0.35 & fair_home < 0.65,
    lambda_sum_nb > quantile(lambda_sum_nb, 0.5, na.rm = TRUE)
  ) %>%
  add_profit()

mean(bets_wider$profit, na.rm = TRUE)
nrow(bets_wider)
t.test(bets_wider$profit)
```

