---
title: "Machine Learning model"
output: html_document
date: "2026-03-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(zoo)
library(corrplot)
library(poissonreg)  # Poisson model
library(xgboost)
library(caret)

# ── Step 1: Create target + rolling features ──────────────────

model_df <- match_xg %>%
  mutate(total_goals = home_goals + away_goals) %>%
  # Reshape to team level so rolling averages are per team
  # Home perspective
  bind_rows(
    match_xg %>% transmute(
      match_id, season, date,
      team        = home_team,
      opponent    = away_team,
      home_away   = "home",
      total_goals = home_goals + away_goals,
      xG_openplay = home_xG_openplay,
      xG_corner   = home_xG_corner,
      xG_freekick = home_xG_freekick,
      xG_penalty  = home_xG_penalty,
      xG_setpiece = home_xG_setpiece
    ),
    match_xg %>% transmute(
      match_id, season, date,
      team        = away_team,
      opponent    = home_team,
      home_away   = "away",
      total_goals = home_goals + away_goals,
      xG_openplay = away_xG_openplay,
      xG_corner   = away_xG_corner,
      xG_freekick = away_xG_freekick,
      xG_penalty  = away_xG_penalty,
      xG_setpiece = away_xG_setpiece
    )
  ) %>%
  arrange(team, date)

# ── Step 2: Rolling 5-game averages (lagged to avoid leakage) ─

model_df <- model_df %>%
  group_by(team) %>%
  mutate(
    roll_xG_openplay = lag(rollmean(xG_openplay, 5, fill = NA, align = "right"), 1),
    roll_xG_corner   = lag(rollmean(xG_corner,   5, fill = NA, align = "right"), 1),
    roll_xG_freekick = lag(rollmean(xG_freekick, 5, fill = NA, align = "right"), 1),
    roll_xG_penalty  = lag(rollmean(xG_penalty,  5, fill = NA, align = "right"), 1),
    roll_xG_setpiece = lag(rollmean(xG_setpiece, 5, fill = NA, align = "right"), 1)
  ) %>%
  ungroup()

# ── Step 3: Join home and away team features per match ─────────

home_features <- model_df %>%
  filter(home_away == "home") %>%
  select(match_id, date, season, total_goals,
         home_roll_xG_openplay = roll_xG_openplay,
         home_roll_xG_corner   = roll_xG_corner,
         home_roll_xG_freekick = roll_xG_freekick,
         home_roll_xG_penalty  = roll_xG_penalty,
         home_roll_xG_setpiece = roll_xG_setpiece)

away_features <- model_df %>%
  filter(home_away == "away") %>%
  select(match_id,
         away_roll_xG_openplay = roll_xG_openplay,
         away_roll_xG_corner   = roll_xG_corner,
         away_roll_xG_freekick = roll_xG_freekick,
         away_roll_xG_penalty  = roll_xG_penalty,
         away_roll_xG_setpiece = roll_xG_setpiece)

final_df <- home_features %>%
  left_join(away_features, by = "match_id") %>%
  drop_na()

# How many rows after dropping NA from rolling window?
nrow(final_df)

# ── Step 4: Check multicollinearity ───────────────────────────

feature_cols <- final_df %>%
  select(starts_with("home_roll"), starts_with("away_roll"))

# Correlation matrix
cor_matrix <- cor(feature_cols)
corrplot(cor_matrix, method = "number", type = "upper", 
         tl.cex = 0.7, number.cex = 0.7)

# Any correlations above 0.8?
high_cor <- which(abs(cor_matrix) > 0.8 & cor_matrix != 1, arr.ind = TRUE)
if (nrow(high_cor) > 0) {
  cat("High correlations found:\n")
  print(high_cor)
} else {
  cat("No high correlations — safe to proceed\n")
}
```




```{r}
library(tidyverse)
library(zoo)
library(corrplot)
library(xgboost)

# ═══════════════════════════════════════════════════════════════
# STEP 1: LOAD DATA
# ═══════════════════════════════════════════════════════════════

epl_shots <- load_understat_league_shots(league = "EPL")

# ═══════════════════════════════════════════════════════════════
# STEP 2: AGGREGATE TO MATCH LEVEL WITH xG BY SITUATION
# ═══════════════════════════════════════════════════════════════

match_xg <- epl_shots %>%
  filter(season >= 2016, season <= 2024) %>%
  mutate(h_a = coalesce(h_a, home_away)) %>%
  group_by(match_id, league, season, date, home_team, away_team, home_goals, away_goals) %>%
  summarise(
    # Total xG
    home_xG          = sum(xG[h_a == "h"], na.rm = TRUE),
    away_xG          = sum(xG[h_a == "a"], na.rm = TRUE),

    # Open play xG
    home_xG_openplay = sum(xG[h_a == "h" & situation == "OpenPlay"], na.rm = TRUE),
    away_xG_openplay = sum(xG[h_a == "a" & situation == "OpenPlay"], na.rm = TRUE),

    # From corner xG
    home_xG_corner   = sum(xG[h_a == "h" & situation == "FromCorner"], na.rm = TRUE),
    away_xG_corner   = sum(xG[h_a == "a" & situation == "FromCorner"], na.rm = TRUE),

    # Direct free kick xG
    home_xG_freekick = sum(xG[h_a == "h" & situation == "DirectFreekick"], na.rm = TRUE),
    away_xG_freekick = sum(xG[h_a == "a" & situation == "DirectFreekick"], na.rm = TRUE),

    # Penalty xG
    home_xG_penalty  = sum(xG[h_a == "h" & situation == "Penalty"], na.rm = TRUE),
    away_xG_penalty  = sum(xG[h_a == "a" & situation == "Penalty"], na.rm = TRUE),

    # Set piece xG (indirect)
    home_xG_setpiece = sum(xG[h_a == "h" & situation == "SetPiece"], na.rm = TRUE),
    away_xG_setpiece = sum(xG[h_a == "a" & situation == "SetPiece"], na.rm = TRUE),

    # All dead ball combined
    home_xG_deadball = sum(xG[h_a == "h" & situation != "OpenPlay"], na.rm = TRUE),
    away_xG_deadball = sum(xG[h_a == "a" & situation != "OpenPlay"], na.rm = TRUE),

    .groups = "drop"
  ) %>%
  arrange(date)

saveRDS(match_xg, "epl_match_xg.rds")

# ═══════════════════════════════════════════════════════════════
# STEP 3: RESHAPE TO TEAM LEVEL + ROLLING AVERAGES
# ═══════════════════════════════════════════════════════════════

# Reshape so each row = one team in one match
model_df <- bind_rows(
  match_xg %>% transmute(
    match_id, season, date,
    team      = home_team,
    opponent  = away_team,
    home_away = "home",
    goals     = home_goals,
    xG_openplay = home_xG_openplay,
    xG_corner   = home_xG_corner,
    xG_freekick = home_xG_freekick,
    xG_penalty  = home_xG_penalty,
    xG_setpiece = home_xG_setpiece
  ),
  match_xg %>% transmute(
    match_id, season, date,
    team      = away_team,
    opponent  = home_team,
    home_away = "away",
    goals     = away_goals,
    xG_openplay = away_xG_openplay,
    xG_corner   = away_xG_corner,
    xG_freekick = away_xG_freekick,
    xG_penalty  = away_xG_penalty,
    xG_setpiece = away_xG_setpiece
  )
) %>%
  arrange(team, date)

# Rolling 5-game averages — lagged by 1 to avoid data leakage
model_df <- model_df %>%
  group_by(team) %>%
  mutate(
    roll_xG_openplay = lag(rollmean(xG_openplay, 5, fill = NA, align = "right"), 1),
    roll_xG_corner   = lag(rollmean(xG_corner,   5, fill = NA, align = "right"), 1),
    roll_xG_freekick = lag(rollmean(xG_freekick, 5, fill = NA, align = "right"), 1),
    roll_xG_penalty  = lag(rollmean(xG_penalty,  5, fill = NA, align = "right"), 1),
    roll_xG_setpiece = lag(rollmean(xG_setpiece, 5, fill = NA, align = "right"), 1)
  ) %>%
  ungroup()

# ═══════════════════════════════════════════════════════════════
# STEP 4: BUILD FINAL MODELLING DATAFRAME
# ═══════════════════════════════════════════════════════════════

# Each row = one team's goals in one match
# Own rolling xG = attacking signal
# Opponent rolling xG = proxy for defensive weakness

team_match_df <- model_df %>%
  # Join opponent rolling stats
  left_join(
    model_df %>%
      select(
        match_id,
        opponent            = team,
        opp_roll_xG_openplay = roll_xG_openplay,
        opp_roll_xG_corner   = roll_xG_corner,
        opp_roll_xG_freekick = roll_xG_freekick,
        opp_roll_xG_penalty  = roll_xG_penalty,
        opp_roll_xG_setpiece = roll_xG_setpiece
      ),
    by = c("match_id", "opponent")
  ) %>%
  mutate(home_flag = if_else(home_away == "home", 1, 0)) %>%
  select(
    match_id, season, date, team, opponent, home_away, home_flag, goals,
    roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
    roll_xG_penalty,  roll_xG_setpiece,
    opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
    opp_roll_xG_penalty,  opp_roll_xG_setpiece
  ) %>%
  drop_na() %>%
  arrange(date)

cat("Final modelling rows:", nrow(team_match_df), "\n")
cat("Seasons:", paste(unique(team_match_df$season), collapse = ", "), "\n")

saveRDS(team_match_df, "epl_team_match_df.rds")

# ═══════════════════════════════════════════════════════════════
# STEP 5: MULTICOLLINEARITY CHECK
# ═══════════════════════════════════════════════════════════════

feature_cols <- team_match_df %>%
  select(
    roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
    roll_xG_penalty,  roll_xG_setpiece,
    opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
    opp_roll_xG_penalty,  opp_roll_xG_setpiece
  )

cor_matrix <- cor(feature_cols)
corrplot(cor_matrix, method = "number", type = "upper",
         tl.cex = 0.7, number.cex = 0.7,
         title = "Feature Correlation Matrix")

high_cor <- which(abs(cor_matrix) > 0.8 & cor_matrix != 1, arr.ind = TRUE)
if (nrow(high_cor) > 0) {
  cat("High correlations found:\n")
  print(high_cor)
} else {
  cat("No high correlations — safe to proceed\n")
}

# ═══════════════════════════════════════════════════════════════
# STEP 6: TRAIN / TEST SPLIT
# Always split by time — never random for time series data
# ═══════════════════════════════════════════════════════════════

train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "| Seasons:", min(train_df$season), "-", max(train_df$season), "\n")
cat("Test rows: ", nrow(test_df),  "| Seasons:", min(test_df$season),  "-", max(test_df$season),  "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 7: POISSON REGRESSION
# ═══════════════════════════════════════════════════════════════

poisson_model <- glm(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece +
    home_flag,
  data   = train_df,
  family = poisson(link = "log")
)

summary(poisson_model)

test_df <- test_df %>%
  mutate(pred_poisson = predict(poisson_model, newdata = test_df, type = "response"))

# step 8── XGBoost fix — create train matrix first ───────────────────

feature_names <- c(
  "roll_xG_openplay", "roll_xG_corner", "roll_xG_freekick",
  "roll_xG_penalty",  "roll_xG_setpiece",
  "opp_roll_xG_openplay", "opp_roll_xG_corner", "opp_roll_xG_freekick",
  "opp_roll_xG_penalty",  "opp_roll_xG_setpiece",
  "home_flag"
)

train_matrix <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

xgb_model <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix,
  nrounds               = 500,
  evals                 = list(train = train_matrix, test = test_matrix),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))

#step 9 ── Evaluate ──────────────────────────────────────────────────

results <- test_df %>%
  summarise(
    rmse_poisson = sqrt(mean((goals - pred_poisson)^2)),
    rmse_xgb     = sqrt(mean((goals - pred_xgb)^2)),
    mae_poisson  = mean(abs(goals - pred_poisson)),
    mae_xgb      = mean(abs(goals - pred_xgb)),
    bias_poisson = mean(pred_poisson - goals),
    bias_xgb     = mean(pred_xgb     - goals)
  )

cat("\n=== Model Evaluation ===\n")
print(results)

#  step 10── Feature importance ────────────────────────────────────────

importance <- xgb.importance(feature_names = feature_names, model = xgb_model)
xgb.plot.importance(importance, top_n = 11)
print(importance)

# step 11── Distribution check ────────────────────────────────────────

test_df %>%
  count(goals) %>%
  mutate(pct_actual = round(n / sum(n), 3)) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded_poisson = round(pred_poisson)) %>%
      count(pred_rounded_poisson) %>%
      mutate(pct_poisson = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded_poisson),
    by = "goals"
  ) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded_xgb = round(pred_xgb)) %>%
      count(pred_rounded_xgb) %>%
      mutate(pct_xgb = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded_xgb),
    by = "goals"
  ) %>%
  print()

# step 12── Visualise ─────────────────────────────────────────────────

test_df %>%
  select(goals, pred_poisson, pred_xgb) %>%
  pivot_longer(-goals, names_to = "model", values_to = "predicted") %>%
  ggplot(aes(x = goals, y = predicted, color = model)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~model) +
  labs(
    title = "Predicted vs Actual Goals (Per Team Per Match)",
    x     = "Actual Goals",
    y     = "Predicted Goals"
  ) +
  theme_minimal()

# ── Save ──────────────────────────────────────────────────────

saveRDS(poisson_model, "poisson_model.rds")
saveRDS(test_df,       "test_predictions.rds")
xgb.save(xgb_model,   "xgb_model.ubj")

cat("\n=== All models saved ===\n")
```

FROM UNDER HERE WE NOW use A EXPONENTIALLY WEIAGHTED MEAN and USE A NEGATIVE BINOMIAL IN ORDER TO SEE IF THE VARIANCE CAN NOT BE SUPPRESED:
Furthermore a standrad deviation is added to the xg of 5 prior matches



















```{r}
library(worldfootballR)
library(tidyverse)
library(zoo)
library(corrplot)
library(xgboost)
library(MASS)

select <- dplyr::select
filter <- dplyr::filter
# ═══════════════════════════════════════════════════════════════
# STEP 1: LOAD DATA
# ═══════════════════════════════════════════════════════════════

epl_shots <- load_understat_league_shots(league = "EPL")

# ═══════════════════════════════════════════════════════════════
# STEP 2: AGGREGATE TO MATCH LEVEL WITH xG BY SITUATION
# ═══════════════════════════════════════════════════════════════

match_xg <- epl_shots %>%
  filter(season >= 2016, season <= 2024) %>%
  mutate(h_a = coalesce(h_a, home_away)) %>%
  group_by(match_id, league, season, date, home_team, away_team, home_goals, away_goals) %>%
  summarise(
    home_xG          = sum(xG[h_a == "h"], na.rm = TRUE),
    away_xG          = sum(xG[h_a == "a"], na.rm = TRUE),
    home_xG_openplay = sum(xG[h_a == "h" & situation == "OpenPlay"], na.rm = TRUE),
    away_xG_openplay = sum(xG[h_a == "a" & situation == "OpenPlay"], na.rm = TRUE),
    home_xG_corner   = sum(xG[h_a == "h" & situation == "FromCorner"], na.rm = TRUE),
    away_xG_corner   = sum(xG[h_a == "a" & situation == "FromCorner"], na.rm = TRUE),
    home_xG_freekick = sum(xG[h_a == "h" & situation == "DirectFreekick"], na.rm = TRUE),
    away_xG_freekick = sum(xG[h_a == "a" & situation == "DirectFreekick"], na.rm = TRUE),
    home_xG_penalty  = sum(xG[h_a == "h" & situation == "Penalty"], na.rm = TRUE),
    away_xG_penalty  = sum(xG[h_a == "a" & situation == "Penalty"], na.rm = TRUE),
    home_xG_setpiece = sum(xG[h_a == "h" & situation == "SetPiece"], na.rm = TRUE),
    away_xG_setpiece = sum(xG[h_a == "a" & situation == "SetPiece"], na.rm = TRUE),
    home_xG_deadball = sum(xG[h_a == "h" & situation != "OpenPlay"], na.rm = TRUE),
    away_xG_deadball = sum(xG[h_a == "a" & situation != "OpenPlay"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(date)

saveRDS(match_xg, "epl_match_xg.rds")

# ═══════════════════════════════════════════════════════════════
# STEP 3: RESHAPE TO TEAM LEVEL
# ═══════════════════════════════════════════════════════════════

model_df <- bind_rows(
  match_xg %>% transmute(
    match_id, season, date,
    team        = home_team,
    opponent    = away_team,
    home_away   = "home",
    goals       = home_goals,
    xG_openplay = home_xG_openplay,
    xG_corner   = home_xG_corner,
    xG_freekick = home_xG_freekick,
    xG_penalty  = home_xG_penalty,
    xG_setpiece = home_xG_setpiece
  ),
  match_xg %>% transmute(
    match_id, season, date,
    team        = away_team,
    opponent    = home_team,
    home_away   = "away",
    goals       = away_goals,
    xG_openplay = away_xG_openplay,
    xG_corner   = away_xG_corner,
    xG_freekick = away_xG_freekick,
    xG_penalty  = away_xG_penalty,
    xG_setpiece = away_xG_setpiece
  )
) %>%
  arrange(team, date)

# ═══════════════════════════════════════════════════════════════
# STEP 4: EXPONENTIALLY WEIGHTED ROLLING AVERAGES
# Recent games weighted more than older games
# ═══════════════════════════════════════════════════════════════

exp_weights <- exp(seq(0, 1, length.out = 5))
exp_weights <- exp_weights / sum(exp_weights)
cat("Exponential weights (oldest to most recent):\n")  # unchanged
print(round(exp_weights, 3))                            # unchanged

wmean5 <- function(x) weighted.mean(x, w = exp_weights)

model_df <- model_df %>%
  group_by(team) %>%
  mutate(
    roll_xG_openplay = lag(rollapply(xG_openplay, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_corner   = lag(rollapply(xG_corner,   5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_freekick = lag(rollapply(xG_freekick, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_penalty  = lag(rollapply(xG_penalty,  5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_setpiece = lag(rollapply(xG_setpiece, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    # NEW — total xG standard deviation over last 5 games
    sd_xG = lag(rollapply(xG_openplay + xG_corner + xG_freekick +
                          xG_penalty  + xG_setpiece,
                          5, FUN = sd, fill = NA, align = "right"), 1)
  ) %>%
  ungroup()

# ═══════════════════════════════════════════════════════════════
# STEP 5: BUILD FINAL MODELLING DATAFRAME
# ═══════════════════════════════════════════════════════════════
exp_weights <- exp(seq(0, 1, length.out = 5))
exp_weights <- exp_weights / sum(exp_weights)
wmean5 <- function(x) weighted.mean(x, w = exp_weights)

team_match_df <- model_df %>%
  left_join(
    model_df %>%
      select(
        match_id,
        opponent             = team,
        opp_roll_xG_openplay = roll_xG_openplay,
        opp_roll_xG_corner   = roll_xG_corner,
        opp_roll_xG_freekick = roll_xG_freekick,
        opp_roll_xG_penalty  = roll_xG_penalty,
        opp_roll_xG_setpiece = roll_xG_setpiece,
        opp_sd_xG            = sd_xG  
      ),
    by = c("match_id", "opponent")
  ) %>%
  mutate(home_flag = if_else(home_away == "home", 1, 0)) %>%
  select(
    match_id, season, date, team, opponent, home_away, home_flag, goals,
    roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
    roll_xG_penalty,  roll_xG_setpiece, sd_xG,
    opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
    opp_roll_xG_penalty,  opp_roll_xG_setpiece, opp_sd_xG
  ) %>%
  drop_na() %>%
  arrange(date)

cat("Final modelling rows:", nrow(team_match_df), "\n")
cat("Seasons:", paste(unique(team_match_df$season), collapse = ", "), "\n")

saveRDS(team_match_df, "epl_team_match_df.rds")

# ═══════════════════════════════════════════════════════════════
# STEP 6: MULTICOLLINEARITY CHECK
# ═══════════════════════════════════════════════════════════════

feature_cols <- team_match_df %>%
  select(
    roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
    roll_xG_penalty,  roll_xG_setpiece,
    opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
    opp_roll_xG_penalty,  opp_roll_xG_setpiece
  )

cor_matrix <- cor(feature_cols)
corrplot(cor_matrix, method = "number", type = "upper",
         tl.cex = 0.7, number.cex = 0.7,
         title = "Feature Correlation Matrix")

high_cor <- which(abs(cor_matrix) > 0.8 & cor_matrix != 1, arr.ind = TRUE)
if (nrow(high_cor) > 0) {
  cat("High correlations found:\n")
  print(high_cor)
} else {
  cat("No high correlations — safe to proceed\n")
}

# ═══════════════════════════════════════════════════════════════
# STEP 7: TRAIN / TEST SPLIT
# ═══════════════════════════════════════════════════════════════

train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "| Seasons:", min(train_df$season), "-", max(train_df$season), "\n")
cat("Test rows: ", nrow(test_df),  "| Seasons:", min(test_df$season),  "-", max(test_df$season),  "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 8: NEGATIVE BINOMIAL MODEL
# ═══════════════════════════════════════════════════════════════

nb_model <- glm.nb(
  goals ~
    roll_xG_openplay + roll_xG_corner + roll_xG_freekick +
    roll_xG_penalty  + roll_xG_setpiece +  sd_xG +
    opp_roll_xG_openplay + opp_roll_xG_corner + opp_roll_xG_freekick +
    opp_roll_xG_penalty  + opp_roll_xG_setpiece + opp_sd_xG +
    home_flag,
  data = train_df
)

summary(nb_model)

# Dispersion parameter — how much extra variance vs Poisson
cat("\nDispersion parameter (theta):", nb_model$theta, "\n")
cat("If theta is large → close to Poisson\n")
cat("If theta is small → much more overdispersion than Poisson\n")

test_df <- test_df %>%
  mutate(pred_nb = predict(nb_model, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 9: XGBOOST
# ═══════════════════════════════════════════════════════════════

feature_names <- c(
  "roll_xG_openplay", "roll_xG_corner", "roll_xG_freekick",
  "roll_xG_penalty",  "roll_xG_setpiece", "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_corner", "opp_roll_xG_freekick",
  "opp_roll_xG_penalty",  "opp_roll_xG_setpiece", "opp_sd_xG",
  "home_flag"
)

train_matrix <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

xgb_model <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix,
  nrounds               = 500,
  evals                 = list(train = train_matrix, test = test_matrix),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))

# ═══════════════════════════════════════════════════════════════
# STEP 10: EVALUATE ALL MODELS
# ═══════════════════════════════════════════════════════════════

results <- test_df %>%
  summarise(
    rmse_nb  = sqrt(mean((goals - pred_nb)^2)),
    rmse_xgb = sqrt(mean((goals - pred_xgb)^2)),
    mae_nb   = mean(abs(goals - pred_nb)),
    mae_xgb  = mean(abs(goals - pred_xgb)),
    bias_nb  = mean(pred_nb  - goals),
    bias_xgb = mean(pred_xgb - goals)
  )

cat("\n=== Model Evaluation ===\n")
print(results)

# ═══════════════════════════════════════════════════════════════
# STEP 11: FEATURE IMPORTANCE
# ═══════════════════════════════════════════════════════════════

importance <- xgb.importance(feature_names = feature_names, model = xgb_model)
xgb.plot.importance(importance, top_n = 11)
print(importance)

# ═══════════════════════════════════════════════════════════════
# STEP 12: DISTRIBUTION CHECK
# ═══════════════════════════════════════════════════════════════

test_df %>%
  count(goals) %>%
  mutate(pct_actual = round(n / sum(n), 3)) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded_nb = round(pred_nb)) %>%
      count(pred_rounded_nb) %>%
      mutate(pct_nb = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded_nb),
    by = "goals"
  ) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded_xgb = round(pred_xgb)) %>%
      count(pred_rounded_xgb) %>%
      mutate(pct_xgb = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded_xgb),
    by = "goals"
  ) %>%
  print()

# ═══════════════════════════════════════════════════════════════
# STEP 13: VISUALISE
# ═══════════════════════════════════════════════════════════════

test_df %>%
  select(goals, pred_nb, pred_xgb) %>%
  pivot_longer(-goals, names_to = "model", values_to = "predicted") %>%
  ggplot(aes(x = goals, y = predicted, color = model)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~model) +
  labs(
    title = "Predicted vs Actual Goals (Per Team Per Match)",
    x     = "Actual Goals",
    y     = "Predicted Goals"
  ) +
  theme_minimal()

# ═══════════════════════════════════════════════════════════════
# STEP 14: SAVE
# ═══════════════════════════════════════════════════════════════

saveRDS(nb_model,  "nb_model.rds")
saveRDS(test_df,   "test_predictions.rds")
xgb.save(xgb_model, "xgb_model.ubj")

cat("\n=== All models saved ===\n")
```





UNDER HERE THE STANDARD DEVIATION HAS BEEN ADDED BEFORE THE MODEL IN ORDER TO SHRINK ESTIMATES OF INCOSISTENT TEAM. ALSO A SIMPLE NEGATIVE BINOMIAL MODEL HAS BEEN MADE BY REMOVING NON SIGNIFICANT VARIABLES. aLSO THE XGBOOST IS KEPT WITH THE STANDARD DEVIATION FROM BEFORE, AS IT SEEMS TO BE IMPORTANT FOR THE XGBOOST BUT NOT THE NEGATIVE BINOMIAL MODEL.


```{r}
# ═══════════════════════════════════════════════════════════════
# STEP 4: EXPONENTIALLY WEIGHTED ROLLING AVERAGES + SD ADJUSTMENT
# ═══════════════════════════════════════════════════════════════

exp_weights <- exp(seq(0, 1, length.out = 5))
exp_weights <- exp_weights / sum(exp_weights)
cat("Exponential weights (oldest to most recent):\n")
print(round(exp_weights, 3))

wmean5 <- function(x) weighted.mean(x, w = exp_weights)

model_df <- model_df %>%
  group_by(team) %>%
  mutate(
    # Weighted rolling means
    roll_xG_openplay = lag(rollapply(xG_openplay, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_corner   = lag(rollapply(xG_corner,   5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_freekick = lag(rollapply(xG_freekick, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_penalty  = lag(rollapply(xG_penalty,  5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_setpiece = lag(rollapply(xG_setpiece, 5, FUN = wmean5, fill = NA, align = "right"), 1),

    # Total xG sd over last 5 games
    sd_xG = lag(rollapply(xG_openplay + xG_corner + xG_freekick +
                          xG_penalty  + xG_setpiece,
                          5, FUN = sd, fill = NA, align = "right"), 1),

    # ── SD adjustment BEFORE model ─────────────────────────────
    # Shrinks estimates for inconsistent teams
    # High sd → less reliable mean → shrink toward zero
    # Low sd  → reliable mean → keep close to original
    roll_xG_openplay_adj = roll_xG_openplay / (1 + sd_xG),
    roll_xG_corner_adj   = roll_xG_corner   / (1 + sd_xG),
    roll_xG_freekick_adj = roll_xG_freekick / (1 + sd_xG),
    roll_xG_penalty_adj  = roll_xG_penalty  / (1 + sd_xG),
    roll_xG_setpiece_adj = roll_xG_setpiece / (1 + sd_xG)
  ) %>%
  ungroup()

# ═══════════════════════════════════════════════════════════════
# STEP 5: BUILD FINAL MODELLING DATAFRAME
# ═══════════════════════════════════════════════════════════════

team_match_df <- model_df %>%
  left_join(
    model_df %>%
      dplyr::select(
        match_id,
        opponent                 = team,
        opp_roll_xG_openplay_adj = roll_xG_openplay_adj,
        opp_roll_xG_corner_adj   = roll_xG_corner_adj,
        opp_roll_xG_freekick_adj = roll_xG_freekick_adj,
        opp_roll_xG_penalty_adj  = roll_xG_penalty_adj,
        opp_roll_xG_setpiece_adj = roll_xG_setpiece_adj,
        opp_sd_xG                = sd_xG
      ),
    by = c("match_id", "opponent")
  ) %>%
  mutate(home_flag = if_else(home_away == "home", 1, 0)) %>%
  dplyr::select(
    match_id, season, date, team, opponent, home_away, home_flag, goals,
    roll_xG_openplay_adj, roll_xG_corner_adj, roll_xG_freekick_adj,
    roll_xG_penalty_adj,  roll_xG_setpiece_adj,
    sd_xG,
    opp_roll_xG_openplay_adj, opp_roll_xG_corner_adj, opp_roll_xG_freekick_adj,
    opp_roll_xG_penalty_adj,  opp_roll_xG_setpiece_adj,
    opp_sd_xG,
    home_flag
  ) %>%
  drop_na() %>%
  arrange(date)

cat("Final modelling rows:", nrow(team_match_df), "\n")
cat("Seasons:", paste(unique(team_match_df$season), collapse = ", "), "\n")
saveRDS(team_match_df, "epl_team_match_df.rds")

# ═══════════════════════════════════════════════════════════════
# STEP 6: MULTICOLLINEARITY CHECK
# ═══════════════════════════════════════════════════════════════

feature_cols <- team_match_df %>%
  dplyr::select(
    roll_xG_openplay_adj, roll_xG_corner_adj, roll_xG_freekick_adj,
    roll_xG_penalty_adj,  roll_xG_setpiece_adj, sd_xG,
    opp_roll_xG_openplay_adj, opp_roll_xG_corner_adj, opp_roll_xG_freekick_adj,
    opp_roll_xG_penalty_adj,  opp_roll_xG_setpiece_adj, opp_sd_xG
  )

cor_matrix <- cor(feature_cols)
corrplot(cor_matrix, method = "number", type = "upper",
         tl.cex = 0.7, number.cex = 0.7,
         title = "Feature Correlation Matrix")

high_cor <- which(abs(cor_matrix) > 0.8 & cor_matrix != 1, arr.ind = TRUE)
if (nrow(high_cor) > 0) {
  cat("High correlations found:\n")
  print(high_cor)
} else {
  cat("No high correlations — safe to proceed\n")
}

# ═══════════════════════════════════════════════════════════════
# STEP 7: TRAIN / TEST SPLIT
# ═══════════════════════════════════════════════════════════════

train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "| Seasons:", min(train_df$season), "-", max(train_df$season), "\n")
cat("Test rows: ", nrow(test_df),  "| Seasons:", min(test_df$season),  "-", max(test_df$season),  "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 8: SIMPLE NB MODEL — 3 variables only
# ═══════════════════════════════════════════════════════════════

nb_simple <- glm.nb(
  goals ~
    roll_xG_openplay_adj +
    opp_roll_xG_openplay_adj +
    home_flag,
  data = train_df
)

summary(nb_simple)
cat("\nTheta:", nb_simple$theta, "\n")

test_df <- test_df %>%
  mutate(pred_nb_simple = predict(nb_simple, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 9: FULL NB MODEL — all adjusted variables
# ═══════════════════════════════════════════════════════════════

nb_full <- glm.nb(
  goals ~
    roll_xG_openplay_adj + roll_xG_corner_adj + roll_xG_freekick_adj +
    roll_xG_penalty_adj  + roll_xG_setpiece_adj +
    sd_xG +
    opp_roll_xG_openplay_adj + opp_roll_xG_corner_adj + opp_roll_xG_freekick_adj +
    opp_roll_xG_penalty_adj  + opp_roll_xG_setpiece_adj +
    opp_sd_xG +
    home_flag,
  data = train_df
)

summary(nb_full)
cat("\nTheta:", nb_full$theta, "\n")

# AIC comparison
cat("\n=== AIC Comparison ===\n")
cat("Simple NB AIC:", AIC(nb_simple), "\n")
cat("Full NB AIC:  ", AIC(nb_full),   "\n")
cat("Lower is better — difference of <2 means models are equivalent\n")

test_df <- test_df %>%
  mutate(pred_nb_full = predict(nb_full, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 10: XGBOOST — full feature set
# ═══════════════════════════════════════════════════════════════

feature_names <- c(
  "roll_xG_openplay_adj", "roll_xG_corner_adj", "roll_xG_freekick_adj",
  "roll_xG_penalty_adj",  "roll_xG_setpiece_adj",
  "sd_xG",
  "opp_roll_xG_openplay_adj", "opp_roll_xG_corner_adj", "opp_roll_xG_freekick_adj",
  "opp_roll_xG_penalty_adj",  "opp_roll_xG_setpiece_adj",
  "opp_sd_xG",
  "home_flag"
)

train_matrix <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

xgb_model <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix,
  nrounds               = 500,
  evals                 = list(train = train_matrix, test = test_matrix),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))

# ═══════════════════════════════════════════════════════════════
# STEP 11: EVALUATE ALL THREE MODELS
# ═══════════════════════════════════════════════════════════════

results <- test_df %>%
  summarise(
    rmse_nb_simple = sqrt(mean((goals - pred_nb_simple)^2)),
    rmse_nb_full   = sqrt(mean((goals - pred_nb_full)^2)),
    rmse_xgb       = sqrt(mean((goals - pred_xgb)^2)),
    mae_nb_simple  = mean(abs(goals - pred_nb_simple)),
    mae_nb_full    = mean(abs(goals - pred_nb_full)),
    mae_xgb        = mean(abs(goals - pred_xgb)),
    bias_nb_simple = mean(pred_nb_simple - goals),
    bias_nb_full   = mean(pred_nb_full   - goals),
    bias_xgb       = mean(pred_xgb       - goals)
  )

cat("\n=== Model Evaluation ===\n")
print(results)

# ═══════════════════════════════════════════════════════════════
# STEP 12: FEATURE IMPORTANCE
# ═══════════════════════════════════════════════════════════════

importance <- xgb.importance(feature_names = feature_names, model = xgb_model)
xgb.plot.importance(importance, top_n = 13)
print(importance)

# ═══════════════════════════════════════════════════════════════
# STEP 13: DISTRIBUTION CHECK — ALL THREE MODELS
# ═══════════════════════════════════════════════════════════════

test_df %>%
  count(goals) %>%
  mutate(pct_actual = round(n / sum(n), 3)) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded = round(pred_nb_simple)) %>%
      count(pred_rounded) %>%
      mutate(pct_nb_simple = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded),
    by = "goals"
  ) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded = round(pred_nb_full)) %>%
      count(pred_rounded) %>%
      mutate(pct_nb_full = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded),
    by = "goals"
  ) %>%
  left_join(
    test_df %>%
      mutate(pred_rounded = round(pred_xgb)) %>%
      count(pred_rounded) %>%
      mutate(pct_xgb = round(n / sum(n), 3)) %>%
      rename(goals = pred_rounded),
    by = "goals"
  ) %>%
  print()

# ═══════════════════════════════════════════════════════════════
# STEP 14: VISUALISE
# ═══════════════════════════════════════════════════════════════

test_df %>%
  dplyr::select(goals, pred_nb_simple, pred_nb_full, pred_xgb) %>%
  pivot_longer(-goals, names_to = "model", values_to = "predicted") %>%
  ggplot(aes(x = goals, y = predicted, color = model)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~model) +
  labs(
    title = "Predicted vs Actual Goals (Per Team Per Match)",
    x     = "Actual Goals",
    y     = "Predicted Goals"
  ) +
  theme_minimal()

# ═══════════════════════════════════════════════════════════════
# STEP 15: SAVE
# ═══════════════════════════════════════════════════════════════

saveRDS(nb_simple, "nb_simple_model.rds")
saveRDS(nb_full,   "nb_full_model.rds")
saveRDS(test_df,   "test_predictions.rds")
xgb.save(xgb_model, "xgb_model.ubj")

cat("\n=== All models saved ===\n")
```


FInal MODEL KINDA:

SO VARAIBLES LEFT ARE both Openplay, Standard deviation, Corners both, and home advantage.
They have been made to into 3 diffrent model with Negative binomial. Also the XGBOOST have same varaibles for comparison, as more data would screw the comparison




```{r}
# ═══════════════════════════════════════════════════════════════
# STEP 1: LOAD & AGGREGATE 
# ═══════════════════════════════════════════════════════════════

epl_shots <- load_understat_league_shots(league = "EPL")

match_xg <- epl_shots %>%
  filter(season >= 2016, season <= 2024) %>%
  mutate(h_a = coalesce(h_a, home_away)) %>%
  group_by(match_id, league, season, date, home_team, away_team, home_goals, away_goals) %>%
  summarise(
    home_xG_openplay = sum(xG[h_a == "h" & situation == "OpenPlay"],      na.rm = TRUE),
    away_xG_openplay = sum(xG[h_a == "a" & situation == "OpenPlay"],      na.rm = TRUE),
    home_xG_corner   = sum(xG[h_a == "h" & situation == "FromCorner"],    na.rm = TRUE),
    away_xG_corner   = sum(xG[h_a == "a" & situation == "FromCorner"],    na.rm = TRUE),
    home_xG_freekick = sum(xG[h_a == "h" & situation == "DirectFreekick"],na.rm = TRUE),
    away_xG_freekick = sum(xG[h_a == "a" & situation == "DirectFreekick"],na.rm = TRUE),
    home_xG_penalty  = sum(xG[h_a == "h" & situation == "Penalty"],       na.rm = TRUE),
    away_xG_penalty  = sum(xG[h_a == "a" & situation == "Penalty"],       na.rm = TRUE),
    home_xG_setpiece = sum(xG[h_a == "h" & situation == "SetPiece"],      na.rm = TRUE),
    away_xG_setpiece = sum(xG[h_a == "a" & situation == "SetPiece"],      na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(date)

# ═══════════════════════════════════════════════════════════════
# STEP 2: RESHAPE TO TEAM LEVEL
# ═══════════════════════════════════════════════════════════════

model_df <- bind_rows(
  match_xg %>% transmute(
    match_id, season, date,
    team        = home_team,
    opponent    = away_team,
    home_away   = "home",
    goals       = home_goals,
    xG_openplay = home_xG_openplay,
    xG_corner   = home_xG_corner,
    xG_freekick = home_xG_freekick,
    xG_penalty  = home_xG_penalty,
    xG_setpiece = home_xG_setpiece
  ),
  match_xg %>% transmute(
    match_id, season, date,
    team        = away_team,
    opponent    = home_team,
    home_away   = "away",
    goals       = away_goals,
    xG_openplay = away_xG_openplay,
    xG_corner   = away_xG_corner,
    xG_freekick = away_xG_freekick,
    xG_penalty  = away_xG_penalty,
    xG_setpiece = away_xG_setpiece
  )
) %>%
  arrange(team, date)

# ═══════════════════════════════════════════════════════════════
# STEP 3: EXPONENTIALLY WEIGHTED ROLLING MEANS + SD
# No pre-adjustment — SD enters model as its own feature
# ═══════════════════════════════════════════════════════════════

exp_weights <- exp(seq(0, 1, length.out = 5))
exp_weights <- exp_weights / sum(exp_weights)
wmean5      <- function(x) weighted.mean(x, w = exp_weights)

model_df <- model_df %>%
  group_by(team) %>%
  mutate(
    roll_xG_openplay = lag(rollapply(xG_openplay, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_corner   = lag(rollapply(xG_corner,   5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_freekick = lag(rollapply(xG_freekick, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_penalty  = lag(rollapply(xG_penalty,  5, FUN = wmean5, fill = NA, align = "right"), 1),
    roll_xG_setpiece = lag(rollapply(xG_setpiece, 5, FUN = wmean5, fill = NA, align = "right"), 1),
    # SD of total xG over last 5 games — raw, no adjustment
    sd_xG = lag(rollapply(
      xG_openplay + xG_corner + xG_freekick + xG_penalty + xG_setpiece,
      5, FUN = sd, fill = NA, align = "right"), 1)
  ) %>%
  ungroup()

# ═══════════════════════════════════════════════════════════════
# STEP 4: BUILD MODELLING DATAFRAME
# ═══════════════════════════════════════════════════════════════

team_match_df <- model_df %>%
  left_join(
    model_df %>%
      dplyr::select(
        match_id,
        opponent             = team,
        opp_roll_xG_openplay = roll_xG_openplay,
        opp_roll_xG_corner   = roll_xG_corner,
        opp_roll_xG_freekick = roll_xG_freekick,
        opp_roll_xG_penalty  = roll_xG_penalty,
        opp_roll_xG_setpiece = roll_xG_setpiece,
        opp_sd_xG            = sd_xG
      ),
    by = c("match_id", "opponent")
  ) %>%
  mutate(home_flag = if_else(home_away == "home", 1, 0)) %>%
  dplyr::select(
    match_id, season, date, team, opponent, home_away, home_flag, goals,
    roll_xG_openplay, roll_xG_corner, roll_xG_freekick,
    roll_xG_penalty,  roll_xG_setpiece, sd_xG,
    opp_roll_xG_openplay, opp_roll_xG_corner, opp_roll_xG_freekick,
    opp_roll_xG_penalty,  opp_roll_xG_setpiece, opp_sd_xG
  ) %>%
  drop_na() %>%
  arrange(date)

cat("Final modelling rows:", nrow(team_match_df), "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 5: TRAIN / TEST SPLIT
# ═══════════════════════════════════════════════════════════════

train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "| Seasons:", min(train_df$season), "-", max(train_df$season), "\n")
cat("Test rows: ", nrow(test_df),  "| Seasons:", min(test_df$season),  "-", max(test_df$season),  "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 6: MODEL 1 — Simple NB (openplay + SD + home)
# Poisson vs NB: theta will tell us if NB is needed
# If theta is very large (>20), Poisson would be nearly identical
# ═══════════════════════════════════════════════════════════════

nb_simple <- glm.nb(
  goals ~
    roll_xG_openplay +
    opp_roll_xG_openplay +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

summary(nb_simple)
cat("\nTheta:", nb_simple$theta,
    "— if >20, Poisson would give near-identical results\n")

test_df <- test_df %>%
  mutate(pred_nb_simple = predict(nb_simple, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 7: MODEL 2 — NB with freekick added
# Freekick xG is rare but high-value — worth testing separately
# ═══════════════════════════════════════════════════════════════

nb_freekick <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

summary(nb_freekick)
cat("\nTheta:", nb_freekick$theta, "\n")

cat("\n=== AIC Comparison: Simple vs Freekick ===\n")
cat("Simple NB AIC:  ", AIC(nb_simple),   "\n")
cat("Freekick NB AIC:", AIC(nb_freekick), "\n")
cat("Lower is better — difference <2 means models are equivalent\n")

test_df <- test_df %>%
  mutate(pred_nb_freekick = predict(nb_freekick, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 8: XGBOOST
# Using same features as nb_freekick model for fair comparison
# XGBoost can find non-linear interactions the NB cannot,
# so keeping feature sets aligned lets you isolate model gains
# ═══════════════════════════════════════════════════════════════

feature_names <- c(
  "roll_xG_openplay", "roll_xG_freekick",
  "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_freekick",
  "opp_sd_xG",
  "home_flag"
)

train_matrix <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

xgb_model <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix,
  nrounds               = 500,
  evals                 = list(train = train_matrix, test = test_matrix),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))

# ═══════════════════════════════════════════════════════════════
# STEP 9: EVALUATE ALL THREE MODELS
# ═══════════════════════════════════════════════════════════════

results <- test_df %>%
  summarise(
    rmse_nb_simple   = sqrt(mean((goals - pred_nb_simple)^2)),
    rmse_nb_freekick = sqrt(mean((goals - pred_nb_freekick)^2)),
    rmse_xgb         = sqrt(mean((goals - pred_xgb)^2)),
    mae_nb_simple    = mean(abs(goals - pred_nb_simple)),
    mae_nb_freekick  = mean(abs(goals - pred_nb_freekick)),
    mae_xgb          = mean(abs(goals - pred_xgb)),
    bias_nb_simple   = mean(pred_nb_simple   - goals),
    bias_nb_freekick = mean(pred_nb_freekick - goals),
    bias_xgb         = mean(pred_xgb         - goals)
  )

cat("\n=== Model Evaluation ===\n")
print(results)

# ═══════════════════════════════════════════════════════════════
# STEP 10: FEATURE IMPORTANCE
# ═══════════════════════════════════════════════════════════════

importance <- xgb.importance(feature_names = feature_names, model = xgb_model)
xgb.plot.importance(importance)
print(importance)

# ═══════════════════════════════════════════════════════════════
# STEP 11: DISTRIBUTION CHECK
# ═══════════════════════════════════════════════════════════════

test_df %>%
  count(goals) %>%
  mutate(pct_actual = round(n / sum(n), 3)) %>%
  left_join(
    test_df %>% mutate(g = round(pred_nb_simple))   %>% count(g) %>% mutate(pct_nb_simple   = round(n/sum(n),3)) %>% rename(goals=g),
    by = "goals") %>%
  left_join(
    test_df %>% mutate(g = round(pred_nb_freekick)) %>% count(g) %>% mutate(pct_nb_freekick = round(n/sum(n),3)) %>% rename(goals=g),
    by = "goals") %>%
  left_join(
    test_df %>% mutate(g = round(pred_xgb))         %>% count(g) %>% mutate(pct_xgb         = round(n/sum(n),3)) %>% rename(goals=g),
    by = "goals") %>%
  print()

# ═══════════════════════════════════════════════════════════════
# STEP 12: VISUALISE
# ═══════════════════════════════════════════════════════════════

test_df %>%
  dplyr::select(goals, pred_nb_simple, pred_nb_freekick, pred_xgb) %>%
  pivot_longer(-goals, names_to = "model", values_to = "predicted") %>%
  ggplot(aes(x = goals, y = predicted, color = model)) +
  geom_jitter(alpha = 0.3, width = 0.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~model) +
  labs(title = "Predicted vs Actual Goals (Per Team Per Match)",
       x = "Actual Goals", y = "Predicted Goals") +
  theme_minimal()

# ═══════════════════════════════════════════════════════════════
# STEP 13: SAVE
# ═══════════════════════════════════════════════════════════════

saveRDS(nb_simple,   "nb_simple_model.rds")
saveRDS(nb_freekick, "nb_freekick_model.rds")
saveRDS(test_df,     "test_predictions.rds")
xgb.save(xgb_model,  "xgb_model.ubj")

cat("\n=== All models saved ===\n")
```

NB model is kept instead of poisson, and there is still slight overdispertion.









ALT UNDER HER ER FORSÆT









#NB and XGBOOST MODELS:

```{r}
# ═══════════════════════════════════════════════════════════════
# STEP 1: PACKAGES
# ═══════════════════════════════════════════════════════════════

library(worldfootballR)
library(tidyverse)
library(zoo)
library(MASS)
library(xgboost)

select <- dplyr::select
filter <- dplyr::filter

# ═══════════════════════════════════════════════════════════════
# STEP 2: LOAD DATA
# pl        = betting market data (match level, one row per match)
# match_xg  = understat xG data
# team_match_df = your existing team-level modelling df
# ═══════════════════════════════════════════════════════════════

pl           <- readRDS("pl_clean.rds")
team_match_df <- readRDS("epl_team_match_df.rds") %>%
  rename(
    roll_xG_openplay     = roll_xG_openplay_adj,
    roll_xG_corner       = roll_xG_corner_adj,
    roll_xG_freekick     = roll_xG_freekick_adj,
    roll_xG_penalty      = roll_xG_penalty_adj,
    roll_xG_setpiece     = roll_xG_setpiece_adj,
    opp_roll_xG_openplay = opp_roll_xG_openplay_adj,
    opp_roll_xG_corner   = opp_roll_xG_corner_adj,
    opp_roll_xG_freekick = opp_roll_xG_freekick_adj,
    opp_roll_xG_penalty  = opp_roll_xG_penalty_adj,
    opp_roll_xG_setpiece = opp_roll_xG_setpiece_adj
  )
# Team name mapping to align understat vs football-data.co.uk names
team_map <- c(
  "Man City"      = "Manchester City",
  "Man United"    = "Manchester United",
  "Newcastle"     = "Newcastle United",
  "Nott'm Forest" = "Nottingham Forest",
  "West Brom"     = "West Bromwich Albion",
  "Wolves"        = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map),
    Date     = as.Date(Date)
  )

# ═══════════════════════════════════════════════════════════════
# STEP 3: TRAIN / TEST SPLIT
# ═══════════════════════════════════════════════════════════════

train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

# ═══════════════════════════════════════════════════════════════
# STEP 4: FIT NB MODEL (freekick version — best AIC from before)
# ═══════════════════════════════════════════════════════════════

nb_model <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

test_df <- test_df %>%
  mutate(pred_nb = predict(nb_model, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 5: FIT XGBOOST
# ═══════════════════════════════════════════════════════════════

feature_names <- c(
  "roll_xG_openplay", "roll_xG_freekick",
  "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_freekick",
  "opp_sd_xG",
  "home_flag"
)

train_matrix <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

xgb_model <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix,
  nrounds               = 500,
  evals                 = list(train = train_matrix, test = test_matrix),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))

# ═══════════════════════════════════════════════════════════════
# STEP 6: CONVERT TEAM-LEVEL PREDICTIONS TO MATCH-LEVEL
# Each match has two rows — join home and away predictions
# Then use Poisson scoreline matrix to get win/draw/loss probs
# ═══════════════════════════════════════════════════════════════

# Helper: given two lambdas, return home win / draw / away win probs
match_probs <- function(lambda_home, lambda_away, max_goals = 8) {
  g <- 0:max_goals
  score_matrix <- outer(dpois(g, lambda_home), dpois(g, lambda_away))
  list(
    prob_home = sum(score_matrix[lower.tri(score_matrix)]),
    prob_draw = sum(diag(score_matrix)),
    prob_away = sum(score_matrix[upper.tri(score_matrix)])
  )
}

# Split test_df into home and away rows
home_preds <- test_df %>%
  filter(home_flag == 1) %>%
  select(match_id, date, season,
         home_team  = team,
         away_team  = opponent,
         home_goals = goals,
         lambda_home_nb  = pred_nb,
         lambda_home_xgb = pred_xgb)

away_preds <- test_df %>%
  filter(home_flag == 0) %>%
  select(match_id,
         away_goals = goals,
         lambda_away_nb  = pred_nb,
         lambda_away_xgb = pred_xgb)

# Join to get one row per match with both lambdas
match_preds <- home_preds %>%
  left_join(away_preds, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals >  away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      home_goals <  away_goals ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )

# Apply scoreline matrix row by row
probs_nb <- pmap_dfr(
  list(match_preds$lambda_home_nb, match_preds$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_nb = p$prob_home, p_draw_nb = p$prob_draw, p_away_nb = p$prob_away)
  }
)

probs_xgb <- pmap_dfr(
  list(match_preds$lambda_home_xgb, match_preds$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_xgb = p$prob_home, p_draw_xgb = p$prob_draw, p_away_xgb = p$prob_away)
  }
)

match_preds <- bind_cols(match_preds, probs_nb, probs_xgb)

# ═══════════════════════════════════════════════════════════════
# STEP 7: MERGE WITH BETTING MARKET DATA
# ═══════════════════════════════════════════════════════════════

# Convert bookmaker odds to fair probabilities (remove overround)
pl_fair <- pl %>%
  filter(Date >= as.Date("2023-08-01")) %>%
  mutate(
    raw_home = 1 / B365H,
    raw_draw = 1 / B365D,
    raw_away = 1 / B365A,
    overround = raw_home + raw_draw + raw_away,
    fair_home = raw_home / overround,
    fair_draw = raw_draw / overround,
    fair_away = raw_away / overround
  ) %>%
  select(Date, HomeTeam, AwayTeam, home_win = FTR,
         fair_home, fair_draw, fair_away) %>%
  mutate(home_win = as.integer(home_win == "H"))

match_preds <- match_preds %>%
  mutate(date = as.Date(date))

# Join on date + team names
combined <- match_preds %>%
  left_join(
    pl_fair,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )

cat("Matched rows:", sum(!is.na(combined$fair_home)), "out of", nrow(combined), "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 8: EVALUATE — LOG LOSS + BRIER SCORE
# ═══════════════════════════════════════════════════════════════

log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

eval_df <- combined %>% filter(!is.na(fair_home))

results_table <- tibble(
  model = c("Bookmaker", "NB model", "XGBoost"),
  log_loss = c(
    log_loss(eval_df$home_win.y, eval_df$fair_home),
    log_loss(eval_df$home_win.y, eval_df$p_home_nb),
    log_loss(eval_df$home_win.y, eval_df$p_home_xgb)
  ),
  brier = c(
    brier_score(eval_df$home_win.y, eval_df$fair_home),
    brier_score(eval_df$home_win.y, eval_df$p_home_nb),
    brier_score(eval_df$home_win.y, eval_df$p_home_xgb)
  )
)

cat("\n=== Model Evaluation vs Bookmaker ===\n")
print(results_table)

# ═══════════════════════════════════════════════════════════════
# STEP 9: CALIBRATION CHECK
# ═══════════════════════════════════════════════════════════════

calibration_nb <- eval_df %>%
  mutate(prob_bucket = cut(p_home_nb, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  group_by(prob_bucket) %>%
  summarise(n = n(), avg_pred = mean(p_home_nb), actual_rate = mean(home_win.y))

calibration_book <- eval_df %>%
  mutate(prob_bucket = cut(fair_home, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  group_by(prob_bucket) %>%
  summarise(n = n(), avg_pred = mean(fair_home), actual_rate = mean(home_win.y))

cat("\n=== Calibration: NB Model ===\n")
print(calibration_nb)

cat("\n=== Calibration: Bookmaker ===\n")
print(calibration_book)

# ═══════════════════════════════════════════════════════════════
# STEP 10: EDGE ANALYSIS
# Where does your model disagree most with the market?
# ═══════════════════════════════════════════════════════════════

combined <- combined %>%
  mutate(
    edge_nb  = p_home_nb  - fair_home,
    edge_xgb = p_home_xgb - fair_home
  )

cat("\n=== Matches where NB disagrees most with market (top 10) ===\n")
combined %>%
  filter(!is.na(edge_nb)) %>%
  arrange(desc(abs(edge_nb))) %>%
  select(date, home_team, away_team, home_goals, away_goals,
         p_home_nb, fair_home, edge_nb) %>%
  slice(1:10) %>%
  print()

# ═══════════════════════════════════════════════════════════════
# STEP 11: SAVE
# ═══════════════════════════════════════════════════════════════

saveRDS(combined, "match_predictions_vs_market.rds")
cat("\n=== Saved ===\n")
```


See if there is any edges hidden in the models
```{r}
# ═══════════════════════════════════════════════════════════════
# EDGE ANALYSIS — WHERE DOES YOUR MODEL BEAT THE MARKET?
# ═══════════════════════════════════════════════════════════════

eval_df <- combined %>% filter(!is.na(fair_home))

# ── 1. BY EDGE SIZE ────────────────────────────────────────────
# Does betting where you disagree most with the market pay off?

edge_analysis <- eval_df %>%
  mutate(
    edge_bucket = cut(edge_nb, 
                      breaks = c(-1, -0.2, -0.1, 0, 0.1, 0.2, 1),
                      labels = c("Strong against", "Mild against", "Neutral low",
                                 "Neutral high", "Mild for", "Strong for"))
  ) %>%
  group_by(edge_bucket) %>%
  summarise(
    n             = n(),
    avg_edge      = mean(edge_nb),
    actual_rate   = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    # ROI if you bet £1 on home win when model says edge > 0
    # using bookmaker odds implied by fair prob
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("=== Performance by edge size ===\n")
print(edge_analysis)

# ── 2. BY MATCH TYPE (favourite vs underdog) ───────────────────

favourite_analysis <- eval_df %>%
  mutate(
    match_type = case_when(
      fair_home > 0.6  ~ "Strong home fav",
      fair_home > 0.45 ~ "Slight home fav",
      fair_home > 0.35 ~ "Slight away fav",
      TRUE             ~ "Strong away fav"
    )
  ) %>%
  group_by(match_type) %>%
  summarise(
    n              = n(),
    actual_rate    = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("\n=== Performance by match type ===\n")
print(favourite_analysis)

# ── 3. BY SEASON ───────────────────────────────────────────────

season_analysis <- eval_df %>%
  group_by(season) %>%
  summarise(
    n              = n(),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15))),
    model_better   = log_loss_model < log_loss_book
  )

cat("\n=== Performance by season ===\n")
print(season_analysis)

# ── 4. SIMULATED BETTING RETURNS ───────────────────────────────
# If you only bet when edge > threshold, what is your ROI?
# This is the most practically useful check

betting_returns <- function(df, edge_threshold, odds_col = "fair_home") {
  bets <- df %>%
    filter(edge_nb > edge_threshold) %>%
    mutate(
      # Decimal odds from fair prob (no margin version)
      decimal_odds = 1 / fair_home,
      # Return: win = odds - 1, lose = -1
      profit = if_else(home_win.y == 1, decimal_odds - 1, -1)
    )
  
  tibble(
    threshold  = edge_threshold,
    n_bets     = nrow(bets),
    total_profit = sum(bets$profit),
    roi        = round(sum(bets$profit) / nrow(bets) * 100, 2),
    win_rate   = round(mean(bets$home_win.y) * 100, 2),
    avg_edge   = round(mean(bets$edge_nb), 3)
  )
}

roi_table <- map_dfr(
  c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30),
  ~betting_returns(eval_df, .x)
)

cat("\n=== Simulated betting ROI by edge threshold ===\n")
print(roi_table)

# ── 5. STATISTICAL TEST — IS YOUR EDGE REAL? ───────────────────
# Diebold-Mariano style test: are log losses significantly different?

model_losses <- eval_df %>%
  mutate(
    loss_model = -(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                   (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    loss_book  = -(home_win.y * log(pmax(fair_home, 1e-15)) + 
                   (1-home_win.y) * log(pmax(1-fair_home, 1e-15))),
    diff       = loss_model - loss_book  # negative = model better
  )

t_result <- t.test(model_losses$diff, mu = 0)

cat("\n=== Statistical test: is model log loss different from bookmaker? ===\n")
cat("Mean difference (model - bookmaker):", round(mean(model_losses$diff), 4), "\n")
cat("Negative = model is better\n")
cat("p-value:", round(t_result$p.value, 4), "\n")
cat("If p < 0.05, the difference is statistically significant\n")
```


So seems to be a problem that favourites dont get acknowledge, so in order to do that, as 
we use ELO diffrences.

so cross season changes is somewhat accounted for when using elo
The rolling xg uses cross season.



# Get all unique match dates in your dataset
match_dates <- unique(as.Date(team_match_df$date))

# Fetch Elo for every match date
elo_by_date <- map_dfr(match_dates, function(d) {
  Sys.sleep(0.5)  # be polite to the API
  read_csv(paste0("http://api.clubelo.com/", d), 
           show_col_types = FALSE) %>%
    filter(Country == "ENG", Level == 1) %>%
    mutate(date = d)
}, .progress = TRUE)


```{r}
elo_by_date <- readRDS("elo_by_date.rds")
```

getting ELO rating from api

```{r}
# Save it first so you don't have to wait 24 min again
saveRDS(elo_by_date, "elo_by_date.rds")
#rename
elo_name_map <- c(
  "Man City"   = "Manchester City",
  "Man United" = "Manchester United",
  "West Brom"  = "West Bromwich Albion",
  "Newcastle"  = "Newcastle United",
  "Wolves"     = "Wolverhampton Wanderers",
  "Forest"     = "Nottingham Forest"
)

elo_by_date <- elo_by_date %>%
  mutate(Club = recode(Club, !!!elo_name_map))

# Verify no mismatches remain
unique(elo_by_date$Club)[!unique(elo_by_date$Club) %in% unique(team_match_df$team)]

```
Replacing names to match other datasets

```{r}


# Step 1: join home elo
team_match_df <- team_match_df %>%
  mutate(date = as.Date(date)) %>%
  left_join(
    elo_home %>% rename(home_elo = Elo),
    by = c("date", "team" = "Club")
  )

# Step 2: join away elo separately
team_match_df <- team_match_df %>%
  left_join(
    elo_home %>% rename(away_elo = Elo),
    by = c("date", "opponent" = "Club")
  )

# Step 3: create diff
team_match_df <- team_match_df %>%
  mutate(elo_diff = home_elo - away_elo)

cat("Missing Elo:", sum(is.na(team_match_df$elo_diff)), "\n")
```
checking for missing data and seqmenting elo
```{r}
#inclding the elo Diff in traning/testing split
# Drop the 2 NA rows
team_match_df <- team_match_df %>%
  drop_na(elo_diff)

# Re-run train/test split so elo_diff is included
train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows: ", nrow(test_df),  "\n")

# Confirm elo_diff exists in both
cat("elo_diff in train:", "elo_diff" %in% names(train_df), "\n")
cat("elo_diff in test: ", "elo_diff" %in% names(test_df),  "\n")

# ── Rerun freekick model on new split for comparison ──────────
nb_freekick <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

test_df <- test_df %>%
  mutate(pred_nb_freekick = predict(nb_freekick, 
                                    newdata = test_df, 
                                    type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 3: NB MODEL WITH ELO
# ═══════════════════════════════════════════════════════════════

nb_elo <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    elo_diff +
    home_flag,
  data = train_df
)

summary(nb_elo)
cat("\nTheta:", nb_elo$theta, "\n")

# AIC comparison vs previous best model
cat("\n=== AIC Comparison ===\n")
cat("NB freekick (no Elo) AIC:", AIC(nb_freekick), "\n")
cat("NB with Elo AIC:         ", AIC(nb_elo),       "\n")

test_df <- test_df %>%
  mutate(pred_nb_elo = predict(nb_elo, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 4: XGBOOST WITH ELO
# ═══════════════════════════════════════════════════════════════

feature_names_elo <- c(
  "roll_xG_openplay", "roll_xG_freekick",
  "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_freekick",
  "opp_sd_xG",
  "elo_diff",
  "home_flag"
)

train_matrix_elo <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names_elo]),
  label = train_df$goals
)

test_matrix_elo <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names_elo]),
  label = test_df$goals
)

xgb_elo <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix_elo,
  nrounds               = 500,
  evals                 = list(train = train_matrix_elo, test = test_matrix_elo),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb_elo = predict(xgb_elo, test_matrix_elo))

# ═══════════════════════════════════════════════════════════════
# STEP 5: EVALUATE — RMSE / MAE vs OLD MODELS
# ═══════════════════════════════════════════════════════════════

results <- test_df %>%
  summarise(
    rmse_nb_freekick = sqrt(mean((goals - pred_nb_freekick)^2)),
    rmse_nb_elo      = sqrt(mean((goals - pred_nb_elo)^2)),
    rmse_xgb_elo     = sqrt(mean((goals - pred_xgb_elo)^2)),
    mae_nb_freekick  = mean(abs(goals - pred_nb_freekick)),
    mae_nb_elo       = mean(abs(goals - pred_nb_elo)),
    mae_xgb_elo      = mean(abs(goals - pred_xgb_elo)),
    bias_nb_elo      = mean(pred_nb_elo  - goals),
    bias_xgb_elo     = mean(pred_xgb_elo - goals)
  )

cat("\n=== Model Evaluation ===\n")
print(results)

# ═══════════════════════════════════════════════════════════════
# STEP 6: FEATURE IMPORTANCE
# ═══════════════════════════════════════════════════════════════

importance_elo <- xgb.importance(feature_names = feature_names_elo, model = xgb_elo)
xgb.plot.importance(importance_elo)
print(importance_elo)

# ═══════════════════════════════════════════════════════════════
# STEP 7: CONVERT TO MATCH LEVEL + BETTING COMPARISON
# ═══════════════════════════════════════════════════════════════

home_preds <- test_df %>%
  filter(home_flag == 1) %>%
  select(match_id, date, season,
         home_team       = team,
         away_team       = opponent,
         home_goals      = goals,
         lambda_home_nb  = pred_nb_elo,
         lambda_home_xgb = pred_xgb_elo)

away_preds <- test_df %>%
  filter(home_flag == 0) %>%
  select(match_id,
         away_goals      = goals,
         lambda_away_nb  = pred_nb_elo,
         lambda_away_xgb = pred_xgb_elo)

match_preds <- home_preds %>%
  left_join(away_preds, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals >  away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      home_goals <  away_goals ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )

match_probs <- function(lambda_home, lambda_away, max_goals = 8) {
  g <- 0:max_goals
  score_matrix <- outer(dpois(g, lambda_home), dpois(g, lambda_away))
  list(
    prob_home = sum(score_matrix[lower.tri(score_matrix)]),
    prob_draw = sum(diag(score_matrix)),
    prob_away = sum(score_matrix[upper.tri(score_matrix)])
  )
}

probs_nb <- pmap_dfr(
  list(match_preds$lambda_home_nb, match_preds$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_nb = p$prob_home, p_draw_nb = p$prob_draw, p_away_nb = p$prob_away)
  }
)

probs_xgb <- pmap_dfr(
  list(match_preds$lambda_home_xgb, match_preds$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_xgb = p$prob_home, p_draw_xgb = p$prob_draw, p_away_xgb = p$prob_away)
  }
)

match_preds <- bind_cols(match_preds, probs_nb, probs_xgb)

# ═══════════════════════════════════════════════════════════════
# STEP 8: MERGE WITH BETTING MARKET
# ═══════════════════════════════════════════════════════════════

pl_fair <- pl %>%
  filter(Date >= as.Date("2023-08-01")) %>%
  mutate(
    raw_home  = 1 / B365H,
    raw_draw  = 1 / B365D,
    raw_away  = 1 / B365A,
    overround = raw_home + raw_draw + raw_away,
    fair_home = raw_home / overround,
    fair_draw = raw_draw / overround,
    fair_away = raw_away / overround
  ) %>%
  select(Date, HomeTeam, AwayTeam, home_win = FTR,
         fair_home, fair_draw, fair_away) %>%
  mutate(home_win = as.integer(home_win == "H"))

match_preds <- match_preds %>%
  mutate(date = as.Date(date))

combined <- match_preds %>%
  left_join(
    pl_fair,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )

cat("Matched rows:", sum(!is.na(combined$fair_home)), "out of", nrow(combined), "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 9: LOG LOSS + BRIER vs BOOKMAKER
# ═══════════════════════════════════════════════════════════════

log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

eval_df <- combined %>% filter(!is.na(fair_home))

results_table <- tibble(
  model = c("Bookmaker", "NB with Elo", "XGBoost with Elo"),
  log_loss = c(
    log_loss(eval_df$home_win.y, eval_df$fair_home),
    log_loss(eval_df$home_win.y, eval_df$p_home_nb),
    log_loss(eval_df$home_win.y, eval_df$p_home_xgb)
  ),
  brier = c(
    brier_score(eval_df$home_win.y, eval_df$fair_home),
    brier_score(eval_df$home_win.y, eval_df$p_home_nb),
    brier_score(eval_df$home_win.y, eval_df$p_home_xgb)
  )
)

cat("\n=== Model Evaluation vs Bookmaker ===\n")
print(results_table)

# ═══════════════════════════════════════════════════════════════
# STEP 10: EDGE ANALYSIS
# ═══════════════════════════════════════════════════════════════

combined <- combined %>%
  mutate(
    edge_nb  = p_home_nb  - fair_home,
    edge_xgb = p_home_xgb - fair_home
  )

# By edge size
edge_analysis <- eval_df %>%
  mutate(edge_nb = p_home_nb - fair_home) %>%
  mutate(
    edge_bucket = cut(edge_nb,
                      breaks = c(-1, -0.2, -0.1, 0, 0.1, 0.2, 1),
                      labels = c("Strong against", "Mild against", "Neutral low",
                                 "Neutral high", "Mild for", "Strong for"))
  ) %>%
  group_by(edge_bucket) %>%
  summarise(
    n              = n(),
    avg_edge       = mean(edge_nb),
    actual_rate    = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("\n=== Performance by edge size ===\n")
print(edge_analysis)

# By match type
favourite_analysis <- eval_df %>%
  mutate(
    match_type = case_when(
      fair_home > 0.6  ~ "Strong home fav",
      fair_home > 0.45 ~ "Slight home fav",
      fair_home > 0.35 ~ "Slight away fav",
      TRUE             ~ "Strong away fav"
    )
  ) %>%
  group_by(match_type) %>%
  summarise(
    n              = n(),
    actual_rate    = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("\n=== Performance by match type ===\n")
print(favourite_analysis)

# ROI simulation
betting_returns <- function(df, edge_threshold) {
  bets <- df %>%
    filter(edge_nb > edge_threshold) %>%
    mutate(
      decimal_odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, decimal_odds - 1, -1)
    )
  tibble(
    threshold    = edge_threshold,
    n_bets       = nrow(bets),
    total_profit = sum(bets$profit),
    roi          = round(sum(bets$profit) / nrow(bets) * 100, 2),
    win_rate     = round(mean(bets$home_win.y) * 100, 2),
    avg_edge     = round(mean(bets$edge_nb), 3)
  )
}

roi_table <- map_dfr(
  c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30),
  ~betting_returns(eval_df %>% mutate(edge_nb = p_home_nb - fair_home), .x)
)

cat("\n=== Simulated betting ROI by edge threshold ===\n")
print(roi_table)

# Statistical test
model_losses <- eval_df %>%
  mutate(
    loss_model = -(home_win.y * log(pmax(p_home_nb, 1e-15)) +
                   (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    loss_book  = -(home_win.y * log(pmax(fair_home, 1e-15)) +
                   (1-home_win.y) * log(pmax(1-fair_home, 1e-15))),
    diff       = loss_model - loss_book
  )

t_result <- t.test(model_losses$diff, mu = 0)
cat("\n=== Statistical test ===\n")
cat("Mean difference (model - bookmaker):", round(mean(model_losses$diff), 4), "\n")
cat("p-value:", round(t_result$p.value, 4), "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 11: SAVE
# ═══════════════════════════════════════════════════════════════

saveRDS(nb_elo,    "nb_elo_model.rds")
saveRDS(test_df,   "test_predictions_elo.rds")
xgb.save(xgb_elo,  "xgb_elo_model.ubj")
saveRDS(combined,  "match_predictions_elo_vs_market.rds")

cat("\n=== All models saved ===\n")
```

The addition of ELO, could seems to make multiculinarity, sow e check for it.
```{r}
cor(train_df$elo_diff, train_df$roll_xG_openplay)
cor(train_df$elo_diff, train_df$opp_roll_xG_openplay)

# Full VIF check
library(car)
vif(lm(goals ~ roll_xG_openplay + roll_xG_freekick +
        opp_roll_xG_openplay + opp_roll_xG_freekick +
        sd_xG + opp_sd_xG + elo_diff + home_flag,
        data = train_df))
```

There is correlation, but the VIF shows it is not.
Removing the opponent open play might also be relevant.
Furthermore maybe making xg rolling be residuals xg rolling instead, and have elo be main.
also the issue with ELO, VIF shows that it is not correlation with all other varaibles but maybe just pariwise, meaning it wont cause estiamation problems.
Also out of sample imporved alot,, if elo would cause noise or harm multicollinarity, it would affect out of sample performance.

Another thing, is go thorugh what varaibles should be a part of the models, the free kick models do not have elo, but that might no change anything.

There is a problem with coeffecints dropping:
The opp_roll_xG_openplay going from -0.261 to near zero is the real issue. It's not just shrinking — it flipped direction and became insignificant. That tells you Elo is absorbing the opponent quality signal almost entirely. Your model is now saying "the opponent's recent xG form doesn't matter once we know the Elo difference" which is a strong and potentially oversimplified claim.

Is this a problem in practice?
There are two ways to interpret it:
The optimistic interpretation is that Elo is simply a better measure of opponent quality than a 5-game rolling xG average, so the model is correctly downweighting the weaker signal in favour of the stronger one. This is arguably fine.
The pessimistic interpretation is that Elo and opponent xG are measuring the same thing from different angles, and by including both you're letting Elo dominate while making the xG coefficients uninterpretable. The coefficient of 0.138 on roll_xG_openplay no longer means what it used to mean — it's now only the marginal effect of recent form conditional on Elo being held constant, which is a much narrower concept.
The practical solution
Rather than worrying about interpreting individual coefficients, focus on what your model is actually for — generating lambda estimates for the scoreline matrix. For that purpose the coefficients don't need to be individually interpretable, they just need to produce good predictions. And your out-of-sample results confirm they do.
If you want interpretable coefficients alongside good predictions, the cleanest approach is to orthogonalise the features — replace roll_xG_openplay with the residual after regressing it on elo_diff:
r# Remove the part of rolling xG that is already explained by Elo
```{r}
train_df <- train_df %>%
  mutate(
    roll_xG_openplay_resid     = residuals(lm(roll_xG_openplay ~ elo_diff, 
                                               data = train_df)),
    opp_roll_xG_openplay_resid = residuals(lm(opp_roll_xG_openplay ~ elo_diff, 
                                               data = train_df))
  )

```
  
Now roll_xG_openplay_resid means "recent form above or below what Elo would predict" — a genuinely independent signal. The coefficient on it would then be interpretable as pure form effect, separate from quality. But this adds complexity and given your VIF values are fine and out-of-sample performance improved, it's more of a theoretical nicety than a practical necessity.
