---
title: "Samlet dokument"
output: html_document
date: "2026-04-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
```

Helper functions
```{r}
log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

make_calibration_table <- function(actual, pred, bucket_width = 0.1) {
  tibble(actual = actual, pred = pred) %>%
    filter(!is.na(actual), !is.na(pred)) %>%
    mutate(
      prob_bucket = cut(
        pred,
        breaks = seq(0, 1, by = bucket_width),
        include.lowest = TRUE
      )
    ) %>%
    group_by(prob_bucket) %>%
    summarise(
      n = n(),
      avg_pred = mean(pred),
      actual_rate = mean(actual),
      .groups = "drop"
    )
}

add_profit <- function(df) {
  df %>%
    mutate(
      odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, odds - 1, -1)
    )
}

select_high_lambda_bets <- function(df, prob_low = 0.4, prob_high = 0.6, q = 0.6) {
  cutoffs <- df %>%
    group_by(league) %>%
    summarise(lambda_cutoff = quantile(lambda_sum_nb, q, na.rm = TRUE), .groups = "drop")
  
  df %>%
    left_join(cutoffs, by = "league") %>%
    filter(
      fair_home > prob_low & fair_home < prob_high,
      lambda_sum_nb > lambda_cutoff
    )
}

select_low_lambda_bets <- function(df, prob_low = 0.4, prob_high = 0.6, q = 0.4) {
  cutoffs <- df %>%
    group_by(league) %>%
    summarise(lambda_cutoff = quantile(lambda_sum_nb, q, na.rm = TRUE), .groups = "drop")
  
  df %>%
    left_join(cutoffs, by = "league") %>%
    filter(
      fair_home > prob_low & fair_home < prob_high,
      lambda_sum_nb < lambda_cutoff
    )
}

select_median_lambda_bets <- function(df, prob_low = 0.4, prob_high = 0.6) {
  cutoffs <- df %>%
    group_by(league) %>%
    summarise(lambda_cutoff = median(lambda_sum_nb, na.rm = TRUE), .groups = "drop")
  
  df %>%
    left_join(cutoffs, by = "league") %>%
    filter(
      fair_home > prob_low & fair_home < prob_high,
      lambda_sum_nb > lambda_cutoff
    )
}
```

Load pooled data
```{r}
pl <- readRDS("pl_match_test_eval.rds") %>%
  mutate(league = "PL")

ll <- readRDS("ll_match_test_eval.rds") %>%
  mutate(league = "LL")

sa <- readRDS("sa_match_test_eval.rds") %>%
  mutate(league = "SA")

bl <- readRDS("bl_match_test_eval.rds") %>%
  mutate(league = "BL")

pooled <- bind_rows(pl, ll, sa, bl)
```

Check pooled sample
```{r}
nrow(pooled)

table(pooled$league)
table(pooled$season)
```
Main pooled out-of-sample comparison
```{r}
results_pooled <- tibble(
  model = c("Bookmaker", "NB model", "XGBoost model"),
  log_loss = c(
    log_loss(pooled$home_win.y, pooled$fair_home),
    log_loss(pooled$home_win.y, pooled$p_home_nb),
    log_loss(pooled$home_win.y, pooled$p_home_xgb)
  ),
  brier = c(
    brier_score(pooled$home_win.y, pooled$fair_home),
    brier_score(pooled$home_win.y, pooled$p_home_nb),
    brier_score(pooled$home_win.y, pooled$p_home_xgb)
  )
)

print(results_pooled)
```
Pooled calibration tables
```{r}
cal_book_pooled <- make_calibration_table(pooled$home_win.y, pooled$fair_home)
cal_nb_pooled   <- make_calibration_table(pooled$home_win.y, pooled$p_home_nb)
cal_xgb_pooled  <- make_calibration_table(pooled$home_win.y, pooled$p_home_xgb)

cal_book_pooled
cal_nb_pooled
cal_xgb_pooled
```

Pooled subgroup analysis: favourites vs non-favourites
```{r}
subgroup_results_pooled <- pooled %>%
  mutate(
    fav_group = case_when(
      fair_home >= 0.60 ~ "Strong home favourite",
      fair_home >= 0.45 ~ "Moderate home favourite",
      fair_home >= 0.30 ~ "Balanced / slight away favourite",
      TRUE ~ "Strong away favourite"
    )
  ) %>%
  group_by(fav_group) %>%
  summarise(
    n = n(),
    book_log_loss = log_loss(home_win.y, fair_home),
    nb_log_loss   = log_loss(home_win.y, p_home_nb),
    xgb_log_loss  = log_loss(home_win.y, p_home_xgb),
    .groups = "drop"
  )

print(subgroup_results_pooled)
```
Create pooled betting-strategy variables
```{r}
pooled <- pooled %>%
  mutate(
    lambda_diff_nb = lambda_home_nb - lambda_away_nb,
    lambda_sum_nb  = lambda_home_nb + lambda_away_nb
  )
```

Sanity check by league
```{r}
pooled %>%
  group_by(league) %>%
  summarise(
    n = n(),
    mean_lambda_sum = mean(lambda_sum_nb, na.rm = TRUE),
    q50 = quantile(lambda_sum_nb, 0.5, na.rm = TRUE),
    q60 = quantile(lambda_sum_nb, 0.6, na.rm = TRUE),
    q70 = quantile(lambda_sum_nb, 0.7, na.rm = TRUE),
    .groups = "drop"
  )
```
Main pooled betting strategy
```{r}
bets_main_pooled <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit()

mean(bets_main_pooled$profit, na.rm = TRUE)
nrow(bets_main_pooled)
table(bets_main_pooled$league)
```
Pooled betting strategy: median split
```{r}
bets_total_goals_pooled <- select_median_lambda_bets(pooled, 0.4, 0.6) %>%
  add_profit()

mean(bets_total_goals_pooled$profit, na.rm = TRUE)
nrow(bets_total_goals_pooled)
table(bets_total_goals_pooled$league)
```
Pooled betting strategy: top 40 percent
```{r}
bets_strict_pooled <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit()

mean(bets_strict_pooled$profit, na.rm = TRUE)
nrow(bets_strict_pooled)
table(bets_strict_pooled$league)
```
Pooled betting strategy: top 30 percent
```{r}
bets_stricter_pooled <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.7) %>%
  add_profit()

mean(bets_stricter_pooled$profit, na.rm = TRUE)
nrow(bets_stricter_pooled)
table(bets_stricter_pooled$league)
```
Pooled betting strategy: low expected goals
```{r}
bets_low_pooled <- select_low_lambda_bets(pooled, 0.4, 0.6, 0.4) %>%
  add_profit()

mean(bets_low_pooled$profit, na.rm = TRUE)
nrow(bets_low_pooled)
table(bets_low_pooled$league)
```
Profit by expected goal intensity quintile
```{r}
profit_by_bin_pooled <- pooled %>%
  group_by(league) %>%
  mutate(bin = ntile(lambda_sum_nb, 5)) %>%
  ungroup() %>%
  add_profit() %>%
  group_by(bin) %>%
  summarise(
    avg_profit = mean(profit, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

print(profit_by_bin_pooled)
```
Sample size checks
```{r}
sample_size_checks_pooled <- tibble(
  segment = c(
    "Mid-range odds only",
    "Mid-range + high total goals (median split)",
    "Mid-range + top 40% total goals",
    "Mid-range + top 30% total goals",
    "Mid-range + low total goals"
  ),
  n = c(
    nrow(pooled %>% filter(fair_home > 0.4 & fair_home < 0.6)),
    nrow(bets_total_goals_pooled),
    nrow(bets_strict_pooled),
    nrow(bets_stricter_pooled),
    nrow(bets_low_pooled)
  )
)

print(sample_size_checks_pooled)
```
Robustness across different expected-goal cutoffs
```{r}
robustness_cutoffs <- c(0.50, 0.60, 0.70, 0.80)

robustness_table_pooled <- purrr::map_dfr(robustness_cutoffs, function(q) {
  
  bets_q <- select_high_lambda_bets(pooled, 0.4, 0.6, q) %>%
    add_profit()
  
  tibble(
    quantile_cutoff = q,
    n_bets = nrow(bets_q),
    avg_profit = mean(bets_q$profit, na.rm = TRUE)
  )
})

print(robustness_table_pooled)
```
Robustness across alternative probability ranges
```{r}
range_results_pooled <- bind_rows(
  select_high_lambda_bets(pooled, 0.35, 0.65, 0.6) %>%
    add_profit() %>%
    summarise(range = "0.35-0.65", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  
  select_high_lambda_bets(pooled, 0.40, 0.60, 0.6) %>%
    add_profit() %>%
    summarise(range = "0.40-0.60", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE)),
  
  select_high_lambda_bets(pooled, 0.45, 0.55, 0.6) %>%
    add_profit() %>%
    summarise(range = "0.45-0.55", n_bets = n(), avg_profit = mean(profit, na.rm = TRUE))
)

print(range_results_pooled)
```
Heterogeneity across leagues
```{r}
league_results <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit() %>%
  group_by(league) %>%
  summarise(
    n_bets = n(),
    avg_profit = mean(profit, na.rm = TRUE),
    actual_home_win = mean(home_win.y),
    avg_book_prob = mean(fair_home),
    .groups = "drop"
  )

print(league_results)
```
Heterogeneity across seasons
```{r}
season_results_pooled <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit() %>%
  group_by(season) %>%
  summarise(
    n_bets = n(),
    avg_profit = mean(profit, na.rm = TRUE),
    actual_home_win = mean(home_win.y),
    avg_book_prob = mean(fair_home),
    .groups = "drop"
  )

print(season_results_pooled)
```
Heterogeneity within the mid-range segment
```{r}
mid_bin_results_pooled <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit() %>%
  mutate(
    prob_bin = cut(
      fair_home,
      breaks = c(0.4, 0.45, 0.5, 0.55, 0.6),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bin) %>%
  summarise(
    n_bets = n(),
    avg_profit = mean(profit, na.rm = TRUE),
    actual_home_win = mean(home_win.y),
    avg_book_prob = mean(fair_home),
    .groups = "drop"
  )

print(mid_bin_results_pooled)
```
Direct comparison: high vs low expected goals
```{r}
high_low_comparison_pooled <- bind_rows(
  select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
    add_profit() %>%
    summarise(
      group = "High expected goals",
      n_bets = n(),
      avg_profit = mean(profit, na.rm = TRUE),
      actual_home_win = mean(home_win.y),
      avg_book_prob = mean(fair_home)
    ),
  
  select_low_lambda_bets(pooled, 0.4, 0.6, 0.4) %>%
    add_profit() %>%
    summarise(
      group = "Low expected goals",
      n_bets = n(),
      avg_profit = mean(profit, na.rm = TRUE),
      actual_home_win = mean(home_win.y),
      avg_book_prob = mean(fair_home)
    )
)

print(high_low_comparison_pooled)
```
Check whether returns are concentrated in a few bets
```{r}
bets_concentration_pooled <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit()

summary(bets_concentration_pooled$profit)

quantile(
  bets_concentration_pooled$profit,
  probs = c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1),
  na.rm = TRUE
)
```
Bootstrap confidence interval for average profit
```{r}
set.seed(123)

bets_boot_pooled <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit()

boot_means_pooled <- replicate(2000, {
  sample_profit <- sample(
    bets_boot_pooled$profit,
    size = nrow(bets_boot_pooled),
    replace = TRUE
  )
  mean(sample_profit, na.rm = TRUE)
})

quantile(boot_means_pooled, probs = c(0.025, 0.5, 0.975))
```

Placebo test
```{r}
set.seed(123)

target_n <- nrow(
  select_high_lambda_bets(pooled, 0.4, 0.6, 0.6)
)

all_mid_bets <- pooled %>%
  filter(fair_home > 0.4 & fair_home < 0.6) %>%
  add_profit()

placebo_means <- replicate(2000, {
  placebo_sample <- all_mid_bets %>%
    slice_sample(n = target_n, replace = FALSE)
  mean(placebo_sample$profit, na.rm = TRUE)
})

quantile(placebo_means, probs = c(0.025, 0.5, 0.975))

mean(
  select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
    add_profit() %>%
    pull(profit)
)
```

Wider strategy check
```{r}
bets_wider_pooled <- pooled %>%
  filter(fair_home > 0.35 & fair_home < 0.65) %>%
  group_by(league) %>%
  filter(lambda_sum_nb > median(lambda_sum_nb, na.rm = TRUE)) %>%
  ungroup() %>%
  add_profit()

mean(bets_wider_pooled$profit, na.rm = TRUE)
nrow(bets_wider_pooled)
table(bets_wider_pooled$league)
t.test(bets_wider_pooled$profit)
```

FURTHER TESTS
```{r}
bets_main_pooled %>%
  group_by(league) %>%
  summarise(
    mean_profit = mean(profit, na.rm = TRUE),
    sd_profit = sd(profit, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

anova_profit <- aov(profit ~ league, data = bets_main_pooled)
summary(anova_profit)
TukeyHSD(anova_profit)

profit_lm <- lm(profit ~ league, data = bets_main_pooled)
summary(profit_lm)
bets_main_pooled <- bets_main_pooled %>%
  mutate(league = relevel(factor(league), ref = "PL"))

profit_lm <- lm(profit ~ league, data = bets_main_pooled)
summary(profit_lm)

high_bets <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
  add_profit() %>%
  mutate(group = "High expected goals")

low_bets <- select_low_lambda_bets(pooled, 0.4, 0.6, 0.4) %>%
  add_profit() %>%
  mutate(group = "Low expected goals")

compare_high_low <- bind_rows(high_bets, low_bets)

compare_high_low %>%
  group_by(group) %>%
  summarise(
    mean_profit = mean(profit, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

t.test(profit ~ group, data = compare_high_low)

bets_reg <- pooled %>%
  filter(fair_home > 0.4 & fair_home < 0.6) %>%
  add_profit() %>%
  mutate(
    league = factor(league),
    lambda_sum_nb = lambda_home_nb + lambda_away_nb
  )

model_interaction <- lm(profit ~ lambda_sum_nb * league, data = bets_reg)
summary(model_interaction)

bets_main_pooled %>%
  group_by(league) %>%
  summarise(
    actual_home_win = mean(home_win.y, na.rm = TRUE),
    avg_book_prob = mean(fair_home, na.rm = TRUE),
    diff = actual_home_win - avg_book_prob,
    .groups = "drop"
  )

set.seed(123)

placebo_by_league <- purrr::map_dfr(c("PL", "LL", "SA", "BL"), function(lg) {
  
  league_df <- pooled %>%
    filter(league == lg, fair_home > 0.4 & fair_home < 0.6) %>%
    add_profit()
  
  target_n <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
    filter(league == lg) %>%
    nrow()
  
  placebo_means <- replicate(2000, {
    placebo_sample <- league_df %>%
      slice_sample(n = target_n, replace = FALSE)
    mean(placebo_sample$profit, na.rm = TRUE)
  })
  
  actual_mean <- select_high_lambda_bets(pooled, 0.4, 0.6, 0.6) %>%
    filter(league == lg) %>%
    add_profit() %>%
    summarise(m = mean(profit, na.rm = TRUE)) %>%
    pull(m)
  
  tibble(
    league = lg,
    actual_mean = actual_mean,
    placebo_pvalue = mean(abs(placebo_means) >= abs(actual_mean))
  )
})

print(placebo_by_league)
```

Maybe include
```{r}
summary(aov(profit ~ league, data = bets_main_pooled))
summary(lm(profit ~ league, data = bets_main_pooled))
t.test(profit ~ group, data = compare_high_low)
```
Diebold Mariano test
```{r}
library(forecast)

dm.test(
  e1 = (pooled$home_win.y - pooled$p_home_nb)^2,
  e2 = (pooled$home_win.y - pooled$fair_home)^2,
  h = 1,
  power = 2
)

dm.test(
  e1 = (pooled$home_win.y - pooled$p_home_xgb)^2,
  e2 = (pooled$home_win.y - pooled$fair_home)^2,
  h = 1,
  power = 2
)
```


