---
title: "merged with xg"
author: "Jonathan Arve Thomsen"
date: "2026-03-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages and dataset
```{r}
library(dplyr)
library(stringr)
library(lubridate)
library(readr)

pl <- readRDS("pl_clean.rds")
match_xg <- readRDS("epl_match_xg.rds")
```

Map team names
```{r}
team_map <- c(
  "Man City" = "Manchester City",
  "Man United" = "Manchester United",
  "Newcastle" = "Newcastle United",
  "Nott'm Forest" = "Nottingham Forest",
  "West Brom" = "West Bromwich Albion",
  "Wolves" = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map)
  )

sort(unique(pl$HomeTeam))
sort(unique(match_xg$home_team))
```
Make Understat date a real date
```{r}
match_xg <- match_xg %>%
  mutate(date = as.Date(date))

summary(match_xg$date)
head(match_xg$date)
```

Include most important xg variables
```{r}
match_xg_small <- match_xg %>%
  select(
    date, season, home_team, away_team,
    home_goals, away_goals,
    home_xG, away_xG
  )
```

Merge datasets
```{r}
pl_xg <- pl %>%
  left_join(
    match_xg_small,
    by = c(
      "Date" = "date",
      "HomeTeam" = "home_team",
      "AwayTeam" = "away_team"
    )
  )
```

Check merge quality
```{r}
sum(is.na(pl_xg$home_xG))
sum(is.na(pl_xg$away_xG))

mean(is.na(pl_xg$home_xG))
mean(is.na(pl_xg$away_xG))
```
Xg feature
```{r}
pl_xg <- pl_xg %>%
  mutate(
    xG_diff = home_xG - away_xG
  )
```

Split data
```{r}
train_xg <- pl_xg %>%
  filter(Date < as.Date("2023-08-01"))

test_xg <- pl_xg %>%
  filter(Date >= as.Date("2023-08-01"))
```

Estimate model
```{r}
model_xg <- glm(
  home_win ~ p_home + xG_diff,
  data = train_xg %>% filter(!is.na(xG_diff)),
  family = binomial
)
```

Predictions
```{r}
test_xg_eval <- test_xg %>%
  filter(!is.na(xG_diff)) %>%
  mutate(
    pred_xg = predict(model_xg, newdata = ., type = "response")
  )
```

Evaluate
```{r}
log_loss(test_xg_eval$home_win, test_xg_eval$pred_xg)
brier_score(test_xg_eval$home_win, test_xg_eval$pred_xg)
```

Rolling setup
```{r}
library(zoo)

pl_xg <- pl_xg %>%
  group_by(HomeTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    home_xG_avg = lag(rollapply(home_xG, 5, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()
```

Train og test data
```{r}
train_xg <- pl_xg %>%
  filter(Date < as.Date("2023-08-01"))

test_xg <- pl_xg %>%
  filter(Date >= as.Date("2023-08-01"))
```
New model
```{r}
model_xg_real <- glm(
  home_win ~ p_home + home_xG_avg,
  data = train_xg %>% filter(!is.na(home_xG_avg)),
  family = binomial
)
```

Predictions
```{r}
test_xg_eval2 <- test_xg %>%
  filter(!is.na(home_xG_avg)) %>%
  mutate(
    pred_xg2 = predict(model_xg_real, newdata = ., type = "response")
  )
```

Evaluering
```{r}
log_loss(test_xg_eval2$home_win, test_xg_eval2$pred_xg2)
brier_score(test_xg_eval2$home_win, test_xg_eval2$pred_xg2)
```

Variance and summary statistics
```{r}
var(test$p_home)
var(test_xg_eval2$pred_xg2)

summary(test$p_home)
summary(test_xg_eval2$pred_xg2)
```

POTENTIEL UDVIDELSE
Including team strength
```{r}
pl_xg <- pl_xg %>%
  group_by(AwayTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    away_xG_avg = lag(rollapply(away_xG, 5, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()

pl_xg <- pl_xg %>%
  mutate(
    xG_diff_avg = home_xG_avg - away_xG_avg
  )
```

Train and test setup again
```{r}
train_xg <- pl_xg %>%
  filter(Date < as.Date("2023-08-01"))

test_xg <- pl_xg %>%
  filter(Date >= as.Date("2023-08-01"))
```


Model setup
```{r}
model_xg_diff <- glm(
  home_win ~ p_home + xG_diff_avg,
  data = train_xg %>% filter(!is.na(xG_diff_avg)),
  family = binomial
)
```

Predictions
```{r}
test_xg_diff_eval <- test_xg %>%
  filter(!is.na(xG_diff_avg)) %>%
  mutate(
    pred_xg_diff = predict(model_xg_diff, newdata = ., type = "response")
  )
```

Evaluate model
```{r}
log_loss(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
brier_score(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
```

Overview of all models
```{r}
results_table <- tibble(
  model = c("Bookmaker", "Logit odds", "Logit + home goals avg", "Logit + xG diff avg"),
  log_loss = c(
    log_loss(test$home_win, test$p_home),
    log_loss(test$home_win, test$pred_logit),
    log_loss(test_eval$home_win, test_eval$pred_extended),
    log_loss(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
  ),
  brier = c(
    brier_score(test$home_win, test$p_home),
    brier_score(test$home_win, test$pred_logit),
    brier_score(test_eval$home_win, test_eval$pred_extended),
    brier_score(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
  )
)

results_table
```
Calibration table for xg-model
```{r}
calibration_xg <- test_xg_diff_eval %>%
  mutate(
    prob_bucket = cut(
      pred_xg_diff,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bucket) %>%
  summarise(
    n = n(),
    avg_pred = mean(pred_xg_diff, na.rm = TRUE),
    actual_home_win = mean(home_win, na.rm = TRUE)
  )

calibration_xg
```
Calibration table for bookmaker
```{r}
calibration_book <- test %>%
  mutate(
    prob_bucket = cut(
      p_home,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bucket) %>%
  summarise(
    n = n(),
    avg_pred = mean(p_home, na.rm = TRUE),
    actual_home_win = mean(home_win, na.rm = TRUE)
  )

calibration_book
```
Compare heterogeneity
```{r}
var(test$p_home)
var(test_xg_diff_eval$pred_xg_diff)

summary(test$p_home)
summary(test_xg_diff_eval$pred_xg_diff)
```
Robustness check with other rolling windows
3 games:
```{r}
pl_xg <- pl_xg %>%
  group_by(HomeTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    home_xG_avg_3 = lag(rollapply(home_xG, 3, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()

pl_xg <- pl_xg %>%
  group_by(AwayTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    away_xG_avg_3 = lag(rollapply(away_xG, 3, mean, fill = NA, align = "right"))
  ) %>%
  ungroup() %>%
  mutate(
    xG_diff_avg_3 = home_xG_avg_3 - away_xG_avg_3
  )
```

10 games
```{r}
pl_xg <- pl_xg %>%
  group_by(HomeTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    home_xG_avg_10 = lag(rollapply(home_xG, 10, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()

pl_xg <- pl_xg %>%
  group_by(AwayTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    away_xG_avg_10 = lag(rollapply(away_xG, 10, mean, fill = NA, align = "right"))
  ) %>%
  ungroup() %>%
  mutate(
    xG_diff_avg_10 = home_xG_avg_10 - away_xG_avg_10
  )
```

Train test again
```{r}
train_xg <- pl_xg %>% filter(Date < as.Date("2023-08-01"))
test_xg  <- pl_xg %>% filter(Date >= as.Date("2023-08-01"))
```

Model for 3 games
```{r}
model_xg_diff_3 <- glm(
  home_win ~ p_home + xG_diff_avg_3,
  data = train_xg %>% filter(!is.na(xG_diff_avg_3)),
  family = binomial
)

test_xg_diff_eval_3 <- test_xg %>%
  filter(!is.na(xG_diff_avg_3)) %>%
  mutate(
    pred_xg_diff_3 = predict(model_xg_diff_3, newdata = ., type = "response")
  )

log_loss(test_xg_diff_eval_3$home_win, test_xg_diff_eval_3$pred_xg_diff_3)
brier_score(test_xg_diff_eval_3$home_win, test_xg_diff_eval_3$pred_xg_diff_3)
```
Model for 10 games
```{r}
model_xg_diff_10 <- glm(
  home_win ~ p_home + xG_diff_avg_10,
  data = train_xg %>% filter(!is.na(xG_diff_avg_10)),
  family = binomial
)

test_xg_diff_eval_10 <- test_xg %>%
  filter(!is.na(xG_diff_avg_10)) %>%
  mutate(
    pred_xg_diff_10 = predict(model_xg_diff_10, newdata = ., type = "response")
  )

log_loss(test_xg_diff_eval_10$home_win, test_xg_diff_eval_10$pred_xg_diff_10)
brier_score(test_xg_diff_eval_10$home_win, test_xg_diff_eval_10$pred_xg_diff_10)
```
# Summary:
# Adding xG-based features does not improve predictive performance
# relative to bookmaker probabilities.
# This suggests that bookmaker odds already incorporate publicly available
# performance metrics such as expected goals.





#NB and XGBOOST MODELS:

```{r}
# ═══════════════════════════════════════════════════════════════
# STEP 1: PACKAGES
# ═══════════════════════════════════════════════════════════════

library(worldfootballR)
library(tidyverse)
library(zoo)
library(MASS)
library(xgboost)

select <- dplyr::select
filter <- dplyr::filter

# ═══════════════════════════════════════════════════════════════
# STEP 2: LOAD DATA
# pl        = betting market data (match level, one row per match)
# match_xg  = understat xG data
# team_match_df = your existing team-level modelling df
# ═══════════════════════════════════════════════════════════════

pl           <- readRDS("pl_clean.rds")
team_match_df <- readRDS("epl_team_match_df.rds")

# Team name mapping to align understat vs football-data.co.uk names
team_map <- c(
  "Man City"      = "Manchester City",
  "Man United"    = "Manchester United",
  "Newcastle"     = "Newcastle United",
  "Nott'm Forest" = "Nottingham Forest",
  "West Brom"     = "West Bromwich Albion",
  "Wolves"        = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map),
    Date     = as.Date(Date)
  )

# ═══════════════════════════════════════════════════════════════
# STEP 3: TRAIN / TEST SPLIT
# ═══════════════════════════════════════════════════════════════

train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

# ═══════════════════════════════════════════════════════════════
# STEP 4: FIT NB MODEL (freekick version — best AIC from before)
# ═══════════════════════════════════════════════════════════════

nb_model <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

test_df <- test_df %>%
  mutate(pred_nb = predict(nb_model, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 5: FIT XGBOOST
# ═══════════════════════════════════════════════════════════════

feature_names <- c(
  "roll_xG_openplay", "roll_xG_freekick",
  "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_freekick",
  "opp_sd_xG",
  "home_flag"
)

train_matrix <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names]),
  label = train_df$goals
)

test_matrix <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names]),
  label = test_df$goals
)

xgb_model <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix,
  nrounds               = 500,
  evals                 = list(train = train_matrix, test = test_matrix),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb = predict(xgb_model, test_matrix))

# ═══════════════════════════════════════════════════════════════
# STEP 6: CONVERT TEAM-LEVEL PREDICTIONS TO MATCH-LEVEL
# Each match has two rows — join home and away predictions
# Then use Poisson scoreline matrix to get win/draw/loss probs
# ═══════════════════════════════════════════════════════════════

# Helper: given two lambdas, return home win / draw / away win probs
match_probs <- function(lambda_home, lambda_away, max_goals = 8) {
  g <- 0:max_goals
  score_matrix <- outer(dpois(g, lambda_home), dpois(g, lambda_away))
  list(
    prob_home = sum(score_matrix[lower.tri(score_matrix)]),
    prob_draw = sum(diag(score_matrix)),
    prob_away = sum(score_matrix[upper.tri(score_matrix)])
  )
}

# Split test_df into home and away rows
home_preds <- test_df %>%
  filter(home_flag == 1) %>%
  select(match_id, date, season,
         home_team  = team,
         away_team  = opponent,
         home_goals = goals,
         lambda_home_nb  = pred_nb,
         lambda_home_xgb = pred_xgb)

away_preds <- test_df %>%
  filter(home_flag == 0) %>%
  select(match_id,
         away_goals = goals,
         lambda_away_nb  = pred_nb,
         lambda_away_xgb = pred_xgb)

# Join to get one row per match with both lambdas
match_preds <- home_preds %>%
  left_join(away_preds, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals >  away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      home_goals <  away_goals ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )

# Apply scoreline matrix row by row
probs_nb <- pmap_dfr(
  list(match_preds$lambda_home_nb, match_preds$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_nb = p$prob_home, p_draw_nb = p$prob_draw, p_away_nb = p$prob_away)
  }
)

probs_xgb <- pmap_dfr(
  list(match_preds$lambda_home_xgb, match_preds$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_xgb = p$prob_home, p_draw_xgb = p$prob_draw, p_away_xgb = p$prob_away)
  }
)

match_preds <- bind_cols(match_preds, probs_nb, probs_xgb)

# ═══════════════════════════════════════════════════════════════
# STEP 7: MERGE WITH BETTING MARKET DATA
# ═══════════════════════════════════════════════════════════════

# Convert bookmaker odds to fair probabilities (remove overround)
pl_fair <- pl %>%
  filter(Date >= as.Date("2023-08-01")) %>%
  mutate(
    raw_home = 1 / B365H,
    raw_draw = 1 / B365D,
    raw_away = 1 / B365A,
    overround = raw_home + raw_draw + raw_away,
    fair_home = raw_home / overround,
    fair_draw = raw_draw / overround,
    fair_away = raw_away / overround
  ) %>%
  select(Date, HomeTeam, AwayTeam, home_win = FTR,
         fair_home, fair_draw, fair_away) %>%
  mutate(home_win = as.integer(home_win == "H"))

match_preds <- match_preds %>%
  mutate(date = as.Date(date))

# Join on date + team names
combined <- match_preds %>%
  left_join(
    pl_fair,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )

cat("Matched rows:", sum(!is.na(combined$fair_home)), "out of", nrow(combined), "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 8: EVALUATE — LOG LOSS + BRIER SCORE
# ═══════════════════════════════════════════════════════════════

log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

eval_df <- combined %>% filter(!is.na(fair_home))

results_table <- tibble(
  model = c("Bookmaker", "NB model", "XGBoost"),
  log_loss = c(
    log_loss(eval_df$home_win.y, eval_df$fair_home),
    log_loss(eval_df$home_win.y, eval_df$p_home_nb),
    log_loss(eval_df$home_win.y, eval_df$p_home_xgb)
  ),
  brier = c(
    brier_score(eval_df$home_win.y, eval_df$fair_home),
    brier_score(eval_df$home_win.y, eval_df$p_home_nb),
    brier_score(eval_df$home_win.y, eval_df$p_home_xgb)
  )
)

cat("\n=== Model Evaluation vs Bookmaker ===\n")
print(results_table)

# ═══════════════════════════════════════════════════════════════
# STEP 9: CALIBRATION CHECK
# ═══════════════════════════════════════════════════════════════

calibration_nb <- eval_df %>%
  mutate(prob_bucket = cut(p_home_nb, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  group_by(prob_bucket) %>%
  summarise(n = n(), avg_pred = mean(p_home_nb), actual_rate = mean(home_win.y))

calibration_book <- eval_df %>%
  mutate(prob_bucket = cut(fair_home, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  group_by(prob_bucket) %>%
  summarise(n = n(), avg_pred = mean(fair_home), actual_rate = mean(home_win.y))

cat("\n=== Calibration: NB Model ===\n")
print(calibration_nb)

cat("\n=== Calibration: Bookmaker ===\n")
print(calibration_book)

# ═══════════════════════════════════════════════════════════════
# STEP 10: EDGE ANALYSIS
# Where does your model disagree most with the market?
# ═══════════════════════════════════════════════════════════════

combined <- combined %>%
  mutate(
    edge_nb  = p_home_nb  - fair_home,
    edge_xgb = p_home_xgb - fair_home
  )

cat("\n=== Matches where NB disagrees most with market (top 10) ===\n")
combined %>%
  filter(!is.na(edge_nb)) %>%
  arrange(desc(abs(edge_nb))) %>%
  select(date, home_team, away_team, home_goals, away_goals,
         p_home_nb, fair_home, edge_nb) %>%
  slice(1:10) %>%
  print()

# ═══════════════════════════════════════════════════════════════
# STEP 11: SAVE
# ═══════════════════════════════════════════════════════════════

saveRDS(combined, "match_predictions_vs_market.rds")
cat("\n=== Saved ===\n")
```


See if there is any edges hidden in the models
```{r}
# ═══════════════════════════════════════════════════════════════
# EDGE ANALYSIS — WHERE DOES YOUR MODEL BEAT THE MARKET?
# ═══════════════════════════════════════════════════════════════

eval_df <- combined %>% filter(!is.na(fair_home))

# ── 1. BY EDGE SIZE ────────────────────────────────────────────
# Does betting where you disagree most with the market pay off?

edge_analysis <- eval_df %>%
  mutate(
    edge_bucket = cut(edge_nb, 
                      breaks = c(-1, -0.2, -0.1, 0, 0.1, 0.2, 1),
                      labels = c("Strong against", "Mild against", "Neutral low",
                                 "Neutral high", "Mild for", "Strong for"))
  ) %>%
  group_by(edge_bucket) %>%
  summarise(
    n             = n(),
    avg_edge      = mean(edge_nb),
    actual_rate   = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    # ROI if you bet £1 on home win when model says edge > 0
    # using bookmaker odds implied by fair prob
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("=== Performance by edge size ===\n")
print(edge_analysis)

# ── 2. BY MATCH TYPE (favourite vs underdog) ───────────────────

favourite_analysis <- eval_df %>%
  mutate(
    match_type = case_when(
      fair_home > 0.6  ~ "Strong home fav",
      fair_home > 0.45 ~ "Slight home fav",
      fair_home > 0.35 ~ "Slight away fav",
      TRUE             ~ "Strong away fav"
    )
  ) %>%
  group_by(match_type) %>%
  summarise(
    n              = n(),
    actual_rate    = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("\n=== Performance by match type ===\n")
print(favourite_analysis)

# ── 3. BY SEASON ───────────────────────────────────────────────

season_analysis <- eval_df %>%
  group_by(season) %>%
  summarise(
    n              = n(),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) + 
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15))),
    model_better   = log_loss_model < log_loss_book
  )

cat("\n=== Performance by season ===\n")
print(season_analysis)

# ── 4. SIMULATED BETTING RETURNS ───────────────────────────────
# If you only bet when edge > threshold, what is your ROI?
# This is the most practically useful check

betting_returns <- function(df, edge_threshold, odds_col = "fair_home") {
  bets <- df %>%
    filter(edge_nb > edge_threshold) %>%
    mutate(
      # Decimal odds from fair prob (no margin version)
      decimal_odds = 1 / fair_home,
      # Return: win = odds - 1, lose = -1
      profit = if_else(home_win.y == 1, decimal_odds - 1, -1)
    )
  
  tibble(
    threshold  = edge_threshold,
    n_bets     = nrow(bets),
    total_profit = sum(bets$profit),
    roi        = round(sum(bets$profit) / nrow(bets) * 100, 2),
    win_rate   = round(mean(bets$home_win.y) * 100, 2),
    avg_edge   = round(mean(bets$edge_nb), 3)
  )
}

roi_table <- map_dfr(
  c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30),
  ~betting_returns(eval_df, .x)
)

cat("\n=== Simulated betting ROI by edge threshold ===\n")
print(roi_table)

# ── 5. STATISTICAL TEST — IS YOUR EDGE REAL? ───────────────────
# Diebold-Mariano style test: are log losses significantly different?

model_losses <- eval_df %>%
  mutate(
    loss_model = -(home_win.y * log(pmax(p_home_nb, 1e-15)) + 
                   (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    loss_book  = -(home_win.y * log(pmax(fair_home, 1e-15)) + 
                   (1-home_win.y) * log(pmax(1-fair_home, 1e-15))),
    diff       = loss_model - loss_book  # negative = model better
  )

t_result <- t.test(model_losses$diff, mu = 0)

cat("\n=== Statistical test: is model log loss different from bookmaker? ===\n")
cat("Mean difference (model - bookmaker):", round(mean(model_losses$diff), 4), "\n")
cat("Negative = model is better\n")
cat("p-value:", round(t_result$p.value, 4), "\n")
cat("If p < 0.05, the difference is statistically significant\n")
```


So seems to be a problem that favourites dont get acknowledge, so in order to do that, as 
we use ELO diffrences.

so cross season changes is somewhat accounted for when using elo
The rolling xg uses cross season.


```{r}
# Get all unique match dates in your dataset
match_dates <- unique(as.Date(team_match_df$date))

# Fetch Elo for every match date
elo_by_date <- map_dfr(match_dates, function(d) {
  Sys.sleep(0.5)  # be polite to the API
  read_csv(paste0("http://api.clubelo.com/", d), 
           show_col_types = FALSE) %>%
    filter(Country == "ENG", Level == 1) %>%
    mutate(date = d)
}, .progress = TRUE)

```
getting ELO rating from api

```{r}
# Save it first so you don't have to wait 24 min again
saveRDS(elo_by_date, "elo_by_date.rds")
#rename
elo_name_map <- c(
  "Man City"   = "Manchester City",
  "Man United" = "Manchester United",
  "West Brom"  = "West Bromwich Albion",
  "Newcastle"  = "Newcastle United",
  "Wolves"     = "Wolverhampton Wanderers",
  "Forest"     = "Nottingham Forest"
)

elo_by_date <- elo_by_date %>%
  mutate(Club = recode(Club, !!!elo_name_map))

# Verify no mismatches remain
unique(elo_by_date$Club)[!unique(elo_by_date$Club) %in% unique(team_match_df$team)]

```
Replacing names to match other datasets

```{r}


# Step 1: join home elo
team_match_df <- team_match_df %>%
  mutate(date = as.Date(date)) %>%
  left_join(
    elo_home %>% rename(home_elo = Elo),
    by = c("date", "team" = "Club")
  )

# Step 2: join away elo separately
team_match_df <- team_match_df %>%
  left_join(
    elo_home %>% rename(away_elo = Elo),
    by = c("date", "opponent" = "Club")
  )

# Step 3: create diff
team_match_df <- team_match_df %>%
  mutate(elo_diff = home_elo - away_elo)

cat("Missing Elo:", sum(is.na(team_match_df$elo_diff)), "\n")
```
checking for missing data and seqmenting elo
```{r}
#inclding the elo Diff in traning/testing split
# Drop the 2 NA rows
team_match_df <- team_match_df %>%
  drop_na(elo_diff)

# Re-run train/test split so elo_diff is included
train_df <- team_match_df %>% filter(season <= 2022)
test_df  <- team_match_df %>% filter(season >= 2023)

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows: ", nrow(test_df),  "\n")

# Confirm elo_diff exists in both
cat("elo_diff in train:", "elo_diff" %in% names(train_df), "\n")
cat("elo_diff in test: ", "elo_diff" %in% names(test_df),  "\n")

# ── Rerun freekick model on new split for comparison ──────────
nb_freekick <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    home_flag,
  data = train_df
)

test_df <- test_df %>%
  mutate(pred_nb_freekick = predict(nb_freekick, 
                                    newdata = test_df, 
                                    type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 3: NB MODEL WITH ELO
# ═══════════════════════════════════════════════════════════════

nb_elo <- glm.nb(
  goals ~
    roll_xG_openplay +
    roll_xG_freekick +
    opp_roll_xG_openplay +
    opp_roll_xG_freekick +
    sd_xG +
    opp_sd_xG +
    elo_diff +
    home_flag,
  data = train_df
)

summary(nb_elo)
cat("\nTheta:", nb_elo$theta, "\n")

# AIC comparison vs previous best model
cat("\n=== AIC Comparison ===\n")
cat("NB freekick (no Elo) AIC:", AIC(nb_freekick), "\n")
cat("NB with Elo AIC:         ", AIC(nb_elo),       "\n")

test_df <- test_df %>%
  mutate(pred_nb_elo = predict(nb_elo, newdata = test_df, type = "response"))

# ═══════════════════════════════════════════════════════════════
# STEP 4: XGBOOST WITH ELO
# ═══════════════════════════════════════════════════════════════

feature_names_elo <- c(
  "roll_xG_openplay", "roll_xG_freekick",
  "sd_xG",
  "opp_roll_xG_openplay", "opp_roll_xG_freekick",
  "opp_sd_xG",
  "elo_diff",
  "home_flag"
)

train_matrix_elo <- xgb.DMatrix(
  data  = as.matrix(train_df[, feature_names_elo]),
  label = train_df$goals
)

test_matrix_elo <- xgb.DMatrix(
  data  = as.matrix(test_df[, feature_names_elo]),
  label = test_df$goals
)

xgb_elo <- xgb.train(
  params = list(
    objective        = "count:poisson",
    eta              = 0.05,
    max_depth        = 4,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = train_matrix_elo,
  nrounds               = 500,
  evals                 = list(train = train_matrix_elo, test = test_matrix_elo),
  early_stopping_rounds = 30,
  verbose               = 1,
  print_every_n         = 50
)

test_df <- test_df %>%
  mutate(pred_xgb_elo = predict(xgb_elo, test_matrix_elo))

# ═══════════════════════════════════════════════════════════════
# STEP 5: EVALUATE — RMSE / MAE vs OLD MODELS
# ═══════════════════════════════════════════════════════════════

results <- test_df %>%
  summarise(
    rmse_nb_freekick = sqrt(mean((goals - pred_nb_freekick)^2)),
    rmse_nb_elo      = sqrt(mean((goals - pred_nb_elo)^2)),
    rmse_xgb_elo     = sqrt(mean((goals - pred_xgb_elo)^2)),
    mae_nb_freekick  = mean(abs(goals - pred_nb_freekick)),
    mae_nb_elo       = mean(abs(goals - pred_nb_elo)),
    mae_xgb_elo      = mean(abs(goals - pred_xgb_elo)),
    bias_nb_elo      = mean(pred_nb_elo  - goals),
    bias_xgb_elo     = mean(pred_xgb_elo - goals)
  )

cat("\n=== Model Evaluation ===\n")
print(results)

# ═══════════════════════════════════════════════════════════════
# STEP 6: FEATURE IMPORTANCE
# ═══════════════════════════════════════════════════════════════

importance_elo <- xgb.importance(feature_names = feature_names_elo, model = xgb_elo)
xgb.plot.importance(importance_elo)
print(importance_elo)

# ═══════════════════════════════════════════════════════════════
# STEP 7: CONVERT TO MATCH LEVEL + BETTING COMPARISON
# ═══════════════════════════════════════════════════════════════

home_preds <- test_df %>%
  filter(home_flag == 1) %>%
  select(match_id, date, season,
         home_team       = team,
         away_team       = opponent,
         home_goals      = goals,
         lambda_home_nb  = pred_nb_elo,
         lambda_home_xgb = pred_xgb_elo)

away_preds <- test_df %>%
  filter(home_flag == 0) %>%
  select(match_id,
         away_goals      = goals,
         lambda_away_nb  = pred_nb_elo,
         lambda_away_xgb = pred_xgb_elo)

match_preds <- home_preds %>%
  left_join(away_preds, by = "match_id") %>%
  mutate(
    actual_result = case_when(
      home_goals >  away_goals ~ "home",
      home_goals == away_goals ~ "draw",
      home_goals <  away_goals ~ "away"
    ),
    home_win = as.integer(actual_result == "home")
  )

match_probs <- function(lambda_home, lambda_away, max_goals = 8) {
  g <- 0:max_goals
  score_matrix <- outer(dpois(g, lambda_home), dpois(g, lambda_away))
  list(
    prob_home = sum(score_matrix[lower.tri(score_matrix)]),
    prob_draw = sum(diag(score_matrix)),
    prob_away = sum(score_matrix[upper.tri(score_matrix)])
  )
}

probs_nb <- pmap_dfr(
  list(match_preds$lambda_home_nb, match_preds$lambda_away_nb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_nb = p$prob_home, p_draw_nb = p$prob_draw, p_away_nb = p$prob_away)
  }
)

probs_xgb <- pmap_dfr(
  list(match_preds$lambda_home_xgb, match_preds$lambda_away_xgb),
  function(lh, la) {
    p <- match_probs(lh, la)
    tibble(p_home_xgb = p$prob_home, p_draw_xgb = p$prob_draw, p_away_xgb = p$prob_away)
  }
)

match_preds <- bind_cols(match_preds, probs_nb, probs_xgb)

# ═══════════════════════════════════════════════════════════════
# STEP 8: MERGE WITH BETTING MARKET
# ═══════════════════════════════════════════════════════════════

pl_fair <- pl %>%
  filter(Date >= as.Date("2023-08-01")) %>%
  mutate(
    raw_home  = 1 / B365H,
    raw_draw  = 1 / B365D,
    raw_away  = 1 / B365A,
    overround = raw_home + raw_draw + raw_away,
    fair_home = raw_home / overround,
    fair_draw = raw_draw / overround,
    fair_away = raw_away / overround
  ) %>%
  select(Date, HomeTeam, AwayTeam, home_win = FTR,
         fair_home, fair_draw, fair_away) %>%
  mutate(home_win = as.integer(home_win == "H"))

match_preds <- match_preds %>%
  mutate(date = as.Date(date))

combined <- match_preds %>%
  left_join(
    pl_fair,
    by = c("date" = "Date", "home_team" = "HomeTeam", "away_team" = "AwayTeam")
  )

cat("Matched rows:", sum(!is.na(combined$fair_home)), "out of", nrow(combined), "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 9: LOG LOSS + BRIER vs BOOKMAKER
# ═══════════════════════════════════════════════════════════════

log_loss <- function(actual, predicted) {
  predicted <- pmax(pmin(predicted, 1 - 1e-15), 1e-15)
  -mean(actual * log(predicted) + (1 - actual) * log(1 - predicted))
}

brier_score <- function(actual, predicted) {
  mean((actual - predicted)^2)
}

eval_df <- combined %>% filter(!is.na(fair_home))

results_table <- tibble(
  model = c("Bookmaker", "NB with Elo", "XGBoost with Elo"),
  log_loss = c(
    log_loss(eval_df$home_win.y, eval_df$fair_home),
    log_loss(eval_df$home_win.y, eval_df$p_home_nb),
    log_loss(eval_df$home_win.y, eval_df$p_home_xgb)
  ),
  brier = c(
    brier_score(eval_df$home_win.y, eval_df$fair_home),
    brier_score(eval_df$home_win.y, eval_df$p_home_nb),
    brier_score(eval_df$home_win.y, eval_df$p_home_xgb)
  )
)

cat("\n=== Model Evaluation vs Bookmaker ===\n")
print(results_table)

# ═══════════════════════════════════════════════════════════════
# STEP 10: EDGE ANALYSIS
# ═══════════════════════════════════════════════════════════════

combined <- combined %>%
  mutate(
    edge_nb  = p_home_nb  - fair_home,
    edge_xgb = p_home_xgb - fair_home
  )

# By edge size
edge_analysis <- eval_df %>%
  mutate(edge_nb = p_home_nb - fair_home) %>%
  mutate(
    edge_bucket = cut(edge_nb,
                      breaks = c(-1, -0.2, -0.1, 0, 0.1, 0.2, 1),
                      labels = c("Strong against", "Mild against", "Neutral low",
                                 "Neutral high", "Mild for", "Strong for"))
  ) %>%
  group_by(edge_bucket) %>%
  summarise(
    n              = n(),
    avg_edge       = mean(edge_nb),
    actual_rate    = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("\n=== Performance by edge size ===\n")
print(edge_analysis)

# By match type
favourite_analysis <- eval_df %>%
  mutate(
    match_type = case_when(
      fair_home > 0.6  ~ "Strong home fav",
      fair_home > 0.45 ~ "Slight home fav",
      fair_home > 0.35 ~ "Slight away fav",
      TRUE             ~ "Strong away fav"
    )
  ) %>%
  group_by(match_type) %>%
  summarise(
    n              = n(),
    actual_rate    = mean(home_win.y),
    avg_model_prob = mean(p_home_nb),
    avg_fair_prob  = mean(fair_home),
    log_loss_model = -mean(home_win.y * log(pmax(p_home_nb, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    log_loss_book  = -mean(home_win.y * log(pmax(fair_home, 1e-15)) +
                           (1-home_win.y) * log(pmax(1-fair_home, 1e-15)))
  )

cat("\n=== Performance by match type ===\n")
print(favourite_analysis)

# ROI simulation
betting_returns <- function(df, edge_threshold) {
  bets <- df %>%
    filter(edge_nb > edge_threshold) %>%
    mutate(
      decimal_odds = 1 / fair_home,
      profit = if_else(home_win.y == 1, decimal_odds - 1, -1)
    )
  tibble(
    threshold    = edge_threshold,
    n_bets       = nrow(bets),
    total_profit = sum(bets$profit),
    roi          = round(sum(bets$profit) / nrow(bets) * 100, 2),
    win_rate     = round(mean(bets$home_win.y) * 100, 2),
    avg_edge     = round(mean(bets$edge_nb), 3)
  )
}

roi_table <- map_dfr(
  c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30),
  ~betting_returns(eval_df %>% mutate(edge_nb = p_home_nb - fair_home), .x)
)

cat("\n=== Simulated betting ROI by edge threshold ===\n")
print(roi_table)

# Statistical test
model_losses <- eval_df %>%
  mutate(
    loss_model = -(home_win.y * log(pmax(p_home_nb, 1e-15)) +
                   (1-home_win.y) * log(pmax(1-p_home_nb, 1e-15))),
    loss_book  = -(home_win.y * log(pmax(fair_home, 1e-15)) +
                   (1-home_win.y) * log(pmax(1-fair_home, 1e-15))),
    diff       = loss_model - loss_book
  )

t_result <- t.test(model_losses$diff, mu = 0)
cat("\n=== Statistical test ===\n")
cat("Mean difference (model - bookmaker):", round(mean(model_losses$diff), 4), "\n")
cat("p-value:", round(t_result$p.value, 4), "\n")

# ═══════════════════════════════════════════════════════════════
# STEP 11: SAVE
# ═══════════════════════════════════════════════════════════════

saveRDS(nb_elo,    "nb_elo_model.rds")
saveRDS(test_df,   "test_predictions_elo.rds")
xgb.save(xgb_elo,  "xgb_elo_model.ubj")
saveRDS(combined,  "match_predictions_elo_vs_market.rds")

cat("\n=== All models saved ===\n")
```

The addition of ELO, could seems to make multiculinarity, sow e check for it.
```{r}
cor(train_df$elo_diff, train_df$roll_xG_openplay)
cor(train_df$elo_diff, train_df$opp_roll_xG_openplay)

# Full VIF check
library(car)
vif(lm(goals ~ roll_xG_openplay + roll_xG_freekick +
        opp_roll_xG_openplay + opp_roll_xG_freekick +
        sd_xG + opp_sd_xG + elo_diff + home_flag,
        data = train_df))
```

There is correlation, but the VIF shows it is not.
Removing the opponent open play might also be relevant.
Furthermore maybe making xg rolling be residuals xg rolling instead, and have elo be main.
also the issue with ELO, VIF shows that it is not correlation with all other varaibles but maybe just pariwise, meaning it wont cause estiamation problems.
Also out of sample imporved alot,, if elo would cause noise or harm multicollinarity, it would affect out of sample performance.

Another thing, is go thorugh what varaibles should be a part of the models, the free kick models do not have elo, but that might no change anything.

There is a problem with coeffecints dropping:
The opp_roll_xG_openplay going from -0.261 to near zero is the real issue. It's not just shrinking — it flipped direction and became insignificant. That tells you Elo is absorbing the opponent quality signal almost entirely. Your model is now saying "the opponent's recent xG form doesn't matter once we know the Elo difference" which is a strong and potentially oversimplified claim.

Is this a problem in practice?
There are two ways to interpret it:
The optimistic interpretation is that Elo is simply a better measure of opponent quality than a 5-game rolling xG average, so the model is correctly downweighting the weaker signal in favour of the stronger one. This is arguably fine.
The pessimistic interpretation is that Elo and opponent xG are measuring the same thing from different angles, and by including both you're letting Elo dominate while making the xG coefficients uninterpretable. The coefficient of 0.138 on roll_xG_openplay no longer means what it used to mean — it's now only the marginal effect of recent form conditional on Elo being held constant, which is a much narrower concept.
The practical solution
Rather than worrying about interpreting individual coefficients, focus on what your model is actually for — generating lambda estimates for the scoreline matrix. For that purpose the coefficients don't need to be individually interpretable, they just need to produce good predictions. And your out-of-sample results confirm they do.
If you want interpretable coefficients alongside good predictions, the cleanest approach is to orthogonalise the features — replace roll_xG_openplay with the residual after regressing it on elo_diff:
r# Remove the part of rolling xG that is already explained by Elo
```{r}
train_df <- train_df %>%
  mutate(
    roll_xG_openplay_resid     = residuals(lm(roll_xG_openplay ~ elo_diff, 
                                               data = train_df)),
    opp_roll_xG_openplay_resid = residuals(lm(opp_roll_xG_openplay ~ elo_diff, 
                                               data = train_df))
  )

```
  
Now roll_xG_openplay_resid means "recent form above or below what Elo would predict" — a genuinely independent signal. The coefficient on it would then be interpretable as pure form effect, separate from quality. But this adds complexity and given your VIF values are fine and out-of-sample performance improved, it's more of a theoretical nicety than a practical necessity.