---
title: "merged with xg"
author: "Jonathan Arve Thomsen"
date: "2026-03-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages and dataset
```{r}
library(dplyr)
library(stringr)
library(lubridate)
library(readr)

pl <- readRDS("pl_clean.rds")
match_xg <- readRDS("epl_match_xg.rds")
```

Map team names
```{r}
team_map <- c(
  "Man City" = "Manchester City",
  "Man United" = "Manchester United",
  "Newcastle" = "Newcastle United",
  "Nott'm Forest" = "Nottingham Forest",
  "West Brom" = "West Bromwich Albion",
  "Wolves" = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map)
  )

sort(unique(pl$HomeTeam))
sort(unique(match_xg$home_team))
```
Make Understat date a real date
```{r}
match_xg <- match_xg %>%
  mutate(date = as.Date(date))

summary(match_xg$date)
head(match_xg$date)
```

Include most important xg variables
```{r}
match_xg_small <- match_xg %>%
  select(
    date, season, home_team, away_team,
    home_goals, away_goals,
    home_xG, away_xG
  )
```

Merge datasets
```{r}
pl_xg <- pl %>%
  left_join(
    match_xg_small,
    by = c(
      "Date" = "date",
      "HomeTeam" = "home_team",
      "AwayTeam" = "away_team"
    )
  )
```

Check merge quality
```{r}
sum(is.na(pl_xg$home_xG))
sum(is.na(pl_xg$away_xG))

mean(is.na(pl_xg$home_xG))
mean(is.na(pl_xg$away_xG))
```
Xg feature
```{r}
pl_xg <- pl_xg %>%
  mutate(
    xG_diff = home_xG - away_xG
  )
```

Split data
```{r}
train_xg <- pl_xg %>%
  filter(Date < as.Date("2023-08-01"))

test_xg <- pl_xg %>%
  filter(Date >= as.Date("2023-08-01"))
```

Estimate model
```{r}
model_xg <- glm(
  home_win ~ p_home + xG_diff,
  data = train_xg %>% filter(!is.na(xG_diff)),
  family = binomial
)
```

Predictions
```{r}
test_xg_eval <- test_xg %>%
  filter(!is.na(xG_diff)) %>%
  mutate(
    pred_xg = predict(model_xg, newdata = ., type = "response")
  )
```

Evaluate
```{r}
log_loss(test_xg_eval$home_win, test_xg_eval$pred_xg)
brier_score(test_xg_eval$home_win, test_xg_eval$pred_xg)
```

Rolling setup
```{r}
library(zoo)

pl_xg <- pl_xg %>%
  group_by(HomeTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    home_xG_avg = lag(rollapply(home_xG, 5, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()
```

Train og test data
```{r}
train_xg <- pl_xg %>%
  filter(Date < as.Date("2023-08-01"))

test_xg <- pl_xg %>%
  filter(Date >= as.Date("2023-08-01"))
```
New model
```{r}
model_xg_real <- glm(
  home_win ~ p_home + home_xG_avg,
  data = train_xg %>% filter(!is.na(home_xG_avg)),
  family = binomial
)
```

Predictions
```{r}
test_xg_eval2 <- test_xg %>%
  filter(!is.na(home_xG_avg)) %>%
  mutate(
    pred_xg2 = predict(model_xg_real, newdata = ., type = "response")
  )
```

Evaluering
```{r}
log_loss(test_xg_eval2$home_win, test_xg_eval2$pred_xg2)
brier_score(test_xg_eval2$home_win, test_xg_eval2$pred_xg2)
```

Variance and summary statistics
```{r}
var(test$p_home)
var(test_xg_eval2$pred_xg2)

summary(test$p_home)
summary(test_xg_eval2$pred_xg2)
```

POTENTIEL UDVIDELSE
Including team strength
```{r}
pl_xg <- pl_xg %>%
  group_by(AwayTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    away_xG_avg = lag(rollapply(away_xG, 5, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()

pl_xg <- pl_xg %>%
  mutate(
    xG_diff_avg = home_xG_avg - away_xG_avg
  )
```

Train and test setup again
```{r}
train_xg <- pl_xg %>%
  filter(Date < as.Date("2023-08-01"))

test_xg <- pl_xg %>%
  filter(Date >= as.Date("2023-08-01"))
```


Model setup
```{r}
model_xg_diff <- glm(
  home_win ~ p_home + xG_diff_avg,
  data = train_xg %>% filter(!is.na(xG_diff_avg)),
  family = binomial
)
```

Predictions
```{r}
test_xg_diff_eval <- test_xg %>%
  filter(!is.na(xG_diff_avg)) %>%
  mutate(
    pred_xg_diff = predict(model_xg_diff, newdata = ., type = "response")
  )
```

Evaluate model
```{r}
log_loss(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
brier_score(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
```

Overview of all models
```{r}
results_table <- tibble(
  model = c("Bookmaker", "Logit odds", "Logit + home goals avg", "Logit + xG diff avg"),
  log_loss = c(
    log_loss(test$home_win, test$p_home),
    log_loss(test$home_win, test$pred_logit),
    log_loss(test_eval$home_win, test_eval$pred_extended),
    log_loss(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
  ),
  brier = c(
    brier_score(test$home_win, test$p_home),
    brier_score(test$home_win, test$pred_logit),
    brier_score(test_eval$home_win, test_eval$pred_extended),
    brier_score(test_xg_diff_eval$home_win, test_xg_diff_eval$pred_xg_diff)
  )
)

results_table
```
Calibration table for xg-model
```{r}
calibration_xg <- test_xg_diff_eval %>%
  mutate(
    prob_bucket = cut(
      pred_xg_diff,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bucket) %>%
  summarise(
    n = n(),
    avg_pred = mean(pred_xg_diff, na.rm = TRUE),
    actual_home_win = mean(home_win, na.rm = TRUE)
  )

calibration_xg
```
Calibration table for bookmaker
```{r}
calibration_book <- test %>%
  mutate(
    prob_bucket = cut(
      p_home,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bucket) %>%
  summarise(
    n = n(),
    avg_pred = mean(p_home, na.rm = TRUE),
    actual_home_win = mean(home_win, na.rm = TRUE)
  )

calibration_book
```
Compare heterogeneity
```{r}
var(test$p_home)
var(test_xg_diff_eval$pred_xg_diff)

summary(test$p_home)
summary(test_xg_diff_eval$pred_xg_diff)
```
Robustness check with other rolling windows
3 games:
```{r}
pl_xg <- pl_xg %>%
  group_by(HomeTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    home_xG_avg_3 = lag(rollapply(home_xG, 3, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()

pl_xg <- pl_xg %>%
  group_by(AwayTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    away_xG_avg_3 = lag(rollapply(away_xG, 3, mean, fill = NA, align = "right"))
  ) %>%
  ungroup() %>%
  mutate(
    xG_diff_avg_3 = home_xG_avg_3 - away_xG_avg_3
  )
```

10 games
```{r}
pl_xg <- pl_xg %>%
  group_by(HomeTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    home_xG_avg_10 = lag(rollapply(home_xG, 10, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()

pl_xg <- pl_xg %>%
  group_by(AwayTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    away_xG_avg_10 = lag(rollapply(away_xG, 10, mean, fill = NA, align = "right"))
  ) %>%
  ungroup() %>%
  mutate(
    xG_diff_avg_10 = home_xG_avg_10 - away_xG_avg_10
  )
```

Train test again
```{r}
train_xg <- pl_xg %>% filter(Date < as.Date("2023-08-01"))
test_xg  <- pl_xg %>% filter(Date >= as.Date("2023-08-01"))
```

Model for 3 games
```{r}
model_xg_diff_3 <- glm(
  home_win ~ p_home + xG_diff_avg_3,
  data = train_xg %>% filter(!is.na(xG_diff_avg_3)),
  family = binomial
)

test_xg_diff_eval_3 <- test_xg %>%
  filter(!is.na(xG_diff_avg_3)) %>%
  mutate(
    pred_xg_diff_3 = predict(model_xg_diff_3, newdata = ., type = "response")
  )

log_loss(test_xg_diff_eval_3$home_win, test_xg_diff_eval_3$pred_xg_diff_3)
brier_score(test_xg_diff_eval_3$home_win, test_xg_diff_eval_3$pred_xg_diff_3)
```
Model for 10 games
```{r}
model_xg_diff_10 <- glm(
  home_win ~ p_home + xG_diff_avg_10,
  data = train_xg %>% filter(!is.na(xG_diff_avg_10)),
  family = binomial
)

test_xg_diff_eval_10 <- test_xg %>%
  filter(!is.na(xG_diff_avg_10)) %>%
  mutate(
    pred_xg_diff_10 = predict(model_xg_diff_10, newdata = ., type = "response")
  )

log_loss(test_xg_diff_eval_10$home_win, test_xg_diff_eval_10$pred_xg_diff_10)
brier_score(test_xg_diff_eval_10$home_win, test_xg_diff_eval_10$pred_xg_diff_10)
```
# Summary:
# Adding xG-based features does not improve predictive performance
# relative to bookmaker probabilities.
# This suggests that bookmaker odds already incorporate publicly available
# performance metrics such as expected goals.


