---
title: "residual chat"
output: html_document
date: "2026-04-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(readr)
library(dplyr)
library(purrr)
library(stringr)
library(zoo)
library(tidyr)
library(lubridate)
```

Helper functions
```{r}
log_loss <- function(actual, p) {
  keep <- !is.na(actual) & !is.na(p)
  actual <- actual[keep]
  p <- p[keep]
  p <- pmin(pmax(p, 1e-15), 1 - 1e-15)
  -mean(actual * log(p) + (1 - actual) * log(1 - p))
}

brier_score <- function(actual, p) {
  keep <- !is.na(actual) & !is.na(p)
  actual <- actual[keep]
  p <- p[keep]
  mean((actual - p)^2)
}

clip_prob <- function(p) {
  pmin(pmax(p, 1e-6), 1 - 1e-6)
}
```

Load Football-Data files
```{r}
files <- c(
  "PL 16-17.csv",
  "PL 17-18.csv",
  "PL 18-19.csv",
  "PL 19-20.csv",
  "PL 20-21.csv",
  "PL 21-22.csv",
  "PL 22-23.csv",
  "PL 23-24.csv",
  "PL 24-25.csv"
)

pl_raw <- map_dfr(files, function(file) {
  read_csv(
    file,
    col_types = cols(Date = col_character()),
    show_col_types = FALSE
  ) %>%
    mutate(season = str_remove(file, "\\.csv$"))
})
```

Select variables and clean data
```{r}
bookmakers <- c("B365", "BW", "IW", "PS", "WH", "VC")

book_cols <- c(
  unlist(lapply(bookmakers, function(b) paste0(b, c("H", "D", "A"))))
)

base_cols <- c("Date", "HomeTeam", "AwayTeam", "FTHG", "FTAG", "FTR", "season")

pl <- pl_raw %>%
  select(any_of(c(base_cols, book_cols))) %>%
  mutate(Date = trimws(Date))

pl$Date <- ifelse(
  nchar(pl$Date) == 8,
  as.character(as.Date(pl$Date, format = "%d/%m/%y")),
  as.character(as.Date(pl$Date, format = "%d/%m/%Y"))
)

pl$Date <- as.Date(pl$Date)

pl <- pl %>%
  arrange(Date) %>%
  mutate(
    match_result = case_when(
      FTR == "H" ~ "HomeWin",
      FTR == "D" ~ "Draw",
      FTR == "A" ~ "AwayWin",
      TRUE ~ NA_character_
    ),
    total_goals = FTHG + FTAG,
    over_2_5 = ifelse(total_goals > 2.5, 1, 0),
    home_win = ifelse(FTR == "H", 1, 0)
  ) %>%
  filter(
    !is.na(Date),
    !is.na(HomeTeam),
    !is.na(AwayTeam),
    !is.na(FTHG),
    !is.na(FTAG),
    !is.na(FTR)
  )
```

Convert odds to implied probabilities
```{r}
make_probs <- function(df, prefix) {
  h_col <- paste0(prefix, "H")
  d_col <- paste0(prefix, "D")
  a_col <- paste0(prefix, "A")

  if (!all(c(h_col, d_col, a_col) %in% names(df))) {
    return(df)
  }

  p_home_raw <- 1 / df[[h_col]]
  p_draw_raw <- 1 / df[[d_col]]
  p_away_raw <- 1 / df[[a_col]]

  prob_sum <- p_home_raw + p_draw_raw + p_away_raw

  df[[paste0("p_home_", prefix)]] <- p_home_raw / prob_sum
  df[[paste0("p_draw_", prefix)]] <- p_draw_raw / prob_sum
  df[[paste0("p_away_", prefix)]] <- p_away_raw / prob_sum

  df
}

for (b in bookmakers) {
  pl <- make_probs(pl, b)
}

available_books <- bookmakers[
  sapply(bookmakers, function(b) paste0("p_home_", b) %in% names(pl))
]

available_books
```

Build market forecast variables
```{r}
home_prob_cols <- paste0("p_home_", available_books)
draw_prob_cols <- paste0("p_draw_", available_books)
away_prob_cols <- paste0("p_away_", available_books)

pl <- pl %>%
  mutate(
    p_home_market = rowMeans(across(all_of(home_prob_cols)), na.rm = TRUE),
    p_draw_market = rowMeans(across(all_of(draw_prob_cols)), na.rm = TRUE),
    p_away_market = rowMeans(across(all_of(away_prob_cols)), na.rm = TRUE),

    p_home_sd = apply(select(., all_of(home_prob_cols)), 1, sd, na.rm = TRUE),
    p_home_min = apply(select(., all_of(home_prob_cols)), 1, min, na.rm = TRUE),
    p_home_max = apply(select(., all_of(home_prob_cols)), 1, max, na.rm = TRUE)
  )
```

Keeping Bet365 as benchmark
```{r}
pl <- pl %>%
  mutate(
    p_home_b365 = if ("p_home_B365" %in% names(.)) p_home_B365 else NA_real_
  )
```

Merge xg data
```{r}
match_xg <- readRDS("epl_match_xg.rds")

team_map <- c(
  "Man City" = "Manchester City",
  "Man United" = "Manchester United",
  "Newcastle" = "Newcastle United",
  "Nott'm Forest" = "Nottingham Forest",
  "West Brom" = "West Bromwich Albion",
  "Wolves" = "Wolverhampton Wanderers"
)

pl <- pl %>%
  mutate(
    HomeTeam = recode(HomeTeam, !!!team_map),
    AwayTeam = recode(AwayTeam, !!!team_map)
  )

match_xg <- match_xg %>%
  mutate(date = as.Date(date))

match_xg_small <- match_xg %>%
  select(
    date, season, home_team, away_team,
    home_xG, away_xG
  )

pl_xg <- pl %>%
  left_join(
    match_xg_small,
    by = c(
      "Date" = "date",
      "HomeTeam" = "home_team",
      "AwayTeam" = "away_team"
    )
  ) %>%
  mutate(
    xG_diff = home_xG - away_xG
  )
```

Rolling features
```{r}
pl_xg <- pl_xg %>%
  group_by(HomeTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    home_xG_avg_5 = lag(rollapply(home_xG, 5, mean, fill = NA, align = "right"))
  ) %>%
  ungroup()

pl_xg <- pl_xg %>%
  group_by(AwayTeam) %>%
  arrange(Date, .by_group = TRUE) %>%
  mutate(
    away_xG_avg_5 = lag(rollapply(away_xG, 5, mean, fill = NA, align = "right"))
  ) %>%
  ungroup() %>%
  mutate(
    xG_diff_avg_5 = home_xG_avg_5 - away_xG_avg_5,
    xG_diff_avg_5_sq = xG_diff_avg_5^2
  )
```

Train/test split
```{r}
train <- pl_xg %>%
  filter(Date < as.Date("2023-08-01"))

test <- pl_xg %>%
  filter(Date >= as.Date("2023-08-01"))
```

MODEL 1: Bet365 raw
```{r}
test <- test %>%
  mutate(pred_b365 = p_home_b365)
```

MODEL 2: Market average raw
```{r}
test <- test %>%
  mutate(pred_market_raw = p_home_market)
```

MODEL 3: Market recalibration
```{r}
model_recal <- glm(
  home_win ~ p_home_market,
  data = train,
  family = binomial
)

test <- test %>%
  mutate(
    pred_market_recal = predict(model_recal, newdata = ., type = "response")
  )
```

MODEL 4: Residual correction with xg
```{r}
train <- train %>%
  mutate(error_market = home_win - p_home_market)

model_error_xg <- lm(
  error_market ~ xG_diff_avg_5,
  data = train %>% filter(!is.na(xG_diff_avg_5), !is.na(p_home_market))
)

test <- test %>%
  mutate(
    error_hat_xg = predict(model_error_xg, newdata = .),
    pred_corrected_xg = clip_prob(p_home_market + error_hat_xg)
  )
```

MODEL 5: Residual correction with xg+nonlinearity
```{r}
model_error_xg_nl <- lm(
  error_market ~ xG_diff_avg_5 + I(xG_diff_avg_5^2),
  data = train %>% filter(!is.na(xG_diff_avg_5), !is.na(p_home_market))
)

test <- test %>%
  mutate(
    error_hat_xg_nl = predict(model_error_xg_nl, newdata = .),
    pred_corrected_xg_nl = clip_prob(p_home_market + error_hat_xg_nl)
  )
```

MODEL 6: Residual correction with xg+nonlinearity+dispersion
```{r}
model_error_full <- lm(
  error_market ~ xG_diff_avg_5 + I(xG_diff_avg_5^2) + p_home_sd,
  data = train %>%
    filter(!is.na(xG_diff_avg_5), !is.na(p_home_market), !is.na(p_home_sd))
)

test <- test %>%
  mutate(
    error_hat_full = predict(model_error_full, newdata = .),
    pred_corrected_full = clip_prob(p_home_market + error_hat_full)
  )
```

MODEL 7: Direct logit robustness model
```{r}
model_direct <- glm(
  home_win ~ p_home_market + xG_diff_avg_5 + I(xG_diff_avg_5^2) + p_home_sd,
  data = train %>%
    filter(!is.na(xG_diff_avg_5), !is.na(p_home_market), !is.na(p_home_sd)),
  family = binomial
)

test <- test %>%
  mutate(
    pred_direct = predict(model_direct, newdata = ., type = "response")
  )
```

Evaluating models
```{r}
results_table <- tibble(
  model = c(
    "Bet365 raw",
    "Market average raw",
    "Market recalibration",
    "Residual correction: xG",
    "Residual correction: xG + nonlinearity",
    "Residual correction: xG + nonlinearity + dispersion",
    "Direct logit robustness"
  ),
  log_loss = c(
    log_loss(test$home_win, test$pred_b365),
    log_loss(test$home_win, test$pred_market_raw),
    log_loss(test$home_win, test$pred_market_recal),
    log_loss(test$home_win, test$pred_corrected_xg),
    log_loss(test$home_win, test$pred_corrected_xg_nl),
    log_loss(test$home_win, test$pred_corrected_full),
    log_loss(test$home_win, test$pred_direct)
  ),
  brier = c(
    brier_score(test$home_win, test$pred_b365),
    brier_score(test$home_win, test$pred_market_raw),
    brier_score(test$home_win, test$pred_market_recal),
    brier_score(test$home_win, test$pred_corrected_xg),
    brier_score(test$home_win, test$pred_corrected_xg_nl),
    brier_score(test$home_win, test$pred_corrected_full),
    brier_score(test$home_win, test$pred_direct)
  )
)

results_table
```

EXTRA: Testing if market is biased

Calibration
```{r}
calibration_market <- test %>%
  mutate(
    prob_bucket = cut(
      pred_market_raw,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(prob_bucket) %>%
  summarise(
    n = n(),
    avg_pred = mean(pred_market_raw, na.rm = TRUE),
    actual_home_win = mean(home_win, na.rm = TRUE),
    .groups = "drop"
  )

calibration_market
```
Residual diagnostics
```{r}
summary(model_error_xg)
summary(model_error_xg_nl)
summary(model_error_full)
```

ADVANCED MODELS

Interaction model
```{r}
model_error_interact <- lm(
  error_market ~ xG_diff_avg_5 * p_home_market + p_home_sd,
  data = train %>%
    filter(!is.na(xG_diff_avg_5), !is.na(p_home_market), !is.na(p_home_sd))
)

test <- test %>%
  mutate(
    error_hat_interact = predict(model_error_interact, newdata = .),
    pred_corrected_interact = clip_prob(p_home_market + error_hat_interact)
  )
```

Interaction+nonlinearity
```{r}
model_error_interact_nl <- lm(
  error_market ~ xG_diff_avg_5 * p_home_market +
                 I(xG_diff_avg_5^2) * p_home_market +
                 p_home_sd,
  data = train %>%
    filter(!is.na(xG_diff_avg_5), !is.na(p_home_market), !is.na(p_home_sd))
)

test <- test %>%
  mutate(
    error_hat_interact_nl = predict(model_error_interact_nl, newdata = .),
    pred_corrected_interact_nl = clip_prob(p_home_market + error_hat_interact_nl)
  )
```

"Disagreement" instead of sd
```{r}
train <- train %>%
  mutate(
    disagreement = p_home_max - p_home_min
  )

test <- test %>%
  mutate(
    disagreement = p_home_max - p_home_min
  )

model_error_disagree <- lm(
  error_market ~ xG_diff_avg_5 * p_home_market +
                 I(xG_diff_avg_5^2) +
                 disagreement,
  data = train %>%
    filter(!is.na(xG_diff_avg_5), !is.na(disagreement))
)

test <- test %>%
  mutate(
    error_hat_disagree = predict(model_error_disagree, newdata = .),
    pred_corrected_disagree = clip_prob(p_home_market + error_hat_disagree)
  )
```

Segmented model
```{r}
train_mid <- train %>%
  filter(p_home_market > 0.3 & p_home_market < 0.7)

train_extreme <- train %>%
  filter(p_home_market <= 0.3 | p_home_market >= 0.7)

model_mid <- lm(
  error_market ~ xG_diff_avg_5 + I(xG_diff_avg_5^2),
  data = train_mid
)

model_extreme <- lm(
  error_market ~ xG_diff_avg_5 + I(xG_diff_avg_5^2),
  data = train_extreme
)

test <- test %>%
  mutate(
    error_hat_segment = ifelse(
      p_home_market > 0.3 & p_home_market < 0.7,
      predict(model_mid, newdata = .),
      predict(model_extreme, newdata = .)
    ),
    pred_corrected_segment = clip_prob(p_home_market + error_hat_segment)
  )
```

Logit transformation
```{r}
logit <- function(p) log(p / (1 - p))
inv_logit <- function(x) exp(x) / (1 + exp(x))

train <- train %>%
  mutate(
    logit_p = logit(p_home_market)
  )

test <- test %>%
  mutate(
    logit_p = logit(p_home_market)
  )

model_logit_adj <- lm(
  home_win ~ logit_p + xG_diff_avg_5 + I(xG_diff_avg_5^2) + p_home_sd,
  data = train %>%
    filter(!is.na(xG_diff_avg_5), !is.na(p_home_sd))
)

test <- test %>%
  mutate(
    pred_logit_adj = clip_prob(predict(model_logit_adj, newdata = .))
  )
```

Updated evaluation table
```{r}
results_table_extended <- tibble(
  model = c(
    "Bet365 raw",
    "Market average raw",
    "Market recalibration",
    "Residual correction full",
    "Interaction model",
    "Interaction + nonlinear",
    "Disagreement model",
    "Segmented model",
    "Logit adjusted model"
  ),
  log_loss = c(
    log_loss(test$home_win, test$pred_b365),
    log_loss(test$home_win, test$pred_market_raw),
    log_loss(test$home_win, test$pred_market_recal),
    log_loss(test$home_win, test$pred_corrected_full),
    log_loss(test$home_win, test$pred_corrected_interact),
    log_loss(test$home_win, test$pred_corrected_interact_nl),
    log_loss(test$home_win, test$pred_corrected_disagree),
    log_loss(test$home_win, test$pred_corrected_segment),
    log_loss(test$home_win, test$pred_logit_adj)
  ),
  brier = c(
    brier_score(test$home_win, test$pred_b365),
    brier_score(test$home_win, test$pred_market_raw),
    brier_score(test$home_win, test$pred_market_recal),
    brier_score(test$home_win, test$pred_corrected_full),
    brier_score(test$home_win, test$pred_corrected_interact),
    brier_score(test$home_win, test$pred_corrected_interact_nl),
    brier_score(test$home_win, test$pred_corrected_disagree),
    brier_score(test$home_win, test$pred_corrected_segment),
    brier_score(test$home_win, test$pred_logit_adj)
  )
)

results_table_extended
```

Test variance og mean i market error
```{r}
test <- test %>%
  mutate(
    error_market = home_win - p_home_market
  )
var(test$error_market)
mean(abs(test$error_market))
```


